### 2025-11-13 draft_group-6.Rmd
draft outline

---
title: "draft_group-6"
output: pdf_document
---

*Paper (40%)* A paper detailing the motivation, methodology, and results of your project, complete with references. (https://dsci445-csu.github.io/project/)

# Outline

## motivation
(10% of the content)\
The Bank Marketing dataset contains information collected from a Portuguese banking institution during multiple direct telemarketing campaigns (source1). *Why??* (We want to justify the efforts of modeling this data)

**Description of dataset** \
Each row represents an individual client that was contacted in an attempt to sell a long-term deposit product. The dataset includes 16 predictor variables describing client demographics, financial attributes, and details about the marketing campaign itself (such as: contact method, call duration, number of contacts, outcome of previous campaigns).
These variables are a mix of categorical and quantitative data, which makes this dataset well-suited for comparing multiple classification models. The target variable indicates whether the client subscribed to a term deposit (yes/no). As usual, the majority of clients do not subscribe (source 2). This causes an imbalance in our data by increasing modeling bias.

**Problem statement/Assumptions** \
We can go over what we are focusing on and what factors we may ignore.

We would like to maintain interpretability in our model. Emphasis is predicated for the model to provide actionable insght to the manager. After all humans love stories.

## methodology 
(60%)

- Data wrangling/cleaning
- For each model (Logistic, LDA, QDA, KNN, Boosting )
  - Model Formulation (training data, with k-fold CV)
  - Feature Selection and model tuning (subset, regularization, dimension reduction)

## results
(25%)

- Model comparison and assessment (Test error, confusion matrix, AUC, BIC, etc.)
- Discussion (make inferrences, determine the favored model)

## references
(5%)

### 2025-11-18 Intro.Rmd
update

---
title: "Introduction"
output: html_document
---

Load data

```{r}
library(readr)
library(ggplot2)

bank_full <- read_delim("bank-full.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
head(bank_full)
```

The data set consists of 16 variables describing the customer and marketing campaign along with temporal aspects:
* age: Age in years 
* job: Occupation (qualitative variable)
* marital: Marital status ( qualitative variable)
* education: Highest level of education (qualitative variable)
* default: has credit in default? (binary)
* balance: average yearly balance
* housing: has housing loan? (binary)
* loan: has personal loan? (binary)
* Contract: contact communication type (qualitative)
* day: last contact day of the week (month)
* month: last contact month of year
* duration: last contact duration, in seconds
* campaign: number of contacts performed during this campaign and for this client
* pdays: number of days that passed by after the client was last contacted from a previous campaign (-1 means client was not previously contacted)
* previous: number of contacts performed before this campaign and for this client
* poutcome: outcome of the previous marketing campaign (qualitative)
* y: Indicator of whether the client subscribed a term deposit (Binary target)

## Distribution

```{r}
library(tidyverse)
library(ggplot2)

bank_long <- bank_full %>%
  select(where(is.numeric)) %>% # numeric columns
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(bank_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "darkblue", color = "white", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol = 4) + 
  labs(
    title = "Distribution of Numeric Variables in bank_full Dataset",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()

```

From the plots one can observe that the numeric variables are heavily skewed (balance, campaign, duration, pdays, previous). The dissemblance of the observations should take into account when considering the bias in our model fit.

```{r}
summary(bank_full)
```

The categorical variables need to be set to categorical (or binary) variables, since R is reading the data as mere strings.

```{r}
bank_full$job <- as.factor(bank_full$job)
bank_full$marital <- as.factor(bank_full$marital)
bank_full$education <- as.factor(bank_full$education)
bank_full$contact <- as.factor(bank_full$contact)
bank_full$poutcome <- as.factor(bank_full$poutcome)
bank_full$day <- as.factor(bank_full$day)
bank_full$month <- as.factor(bank_full$month)

bank_full$default <- ifelse(bank_full$default == "yes", 1, 0)
bank_full$housing <- ifelse(bank_full$housing == "yes", 1, 0)
bank_full$loan <- ifelse(bank_full$loan == "yes", 1, 0)

```

Plots for categorical variables

```{r}

```

## We plan to use the k-fold cross validation approach. 
For consideration: Are we able to make our training sets a sort of proper stratification of our population dataset (Imbalance in amounts of customers subscribed).

```{r}
bank_train <- 
bank_test <- 
```

Model set up...

## Lgoistic Regression

```{r}
library(glmnet)
library(ggplot2)

bank_glm <- glm(y ~., data = bank_train, family = binomial)
summary(bank_glm)
```

estimate test error

## LDA and QDA

```{r}
library(MASS)

bank_lda <- lda(y ~., data = bank_full, subset = train)
bank_lda

bank_qda <- qda(y ~., data = bank_full, subset = train)
```

estimate test error

## KNN

```{r}
library(class)

bank_knn <- knn()
```

estimate test error

Optimize K value (w.r.t cross validation tesst error)

## GAM

```{r}
library(gam)
bank_gam <- gam(y ~ , data = bank_train) # basis functions

```

## Boosting Decision Trees
require post-hoc explanation

```{r}

```

Potential model Mixed effects model (attribute set variables to have a random effect)

### 2025-11-18 draft_group-6.Rmd
update

@@
 title: "draft_group-6"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
 ---
@@
 ## motivation
-(10% of the content)\
+(13% of the content)\
 The Bank Marketing dataset contains information collected from a Portuguese banking institution during multiple direct telemarketing campaigns (source1).
@@
 - Data wrangling/cleaning
-- For each model (Logistic, LDA, QDA, KNN, Boosting )
+- For each model (Logistic, LDA, QDA, KNN, GAM, Boosting)
   - Model Formulation (training data, with k-fold CV)
@@
 ## references
-(5%)
+(2%)

### 2025-11-18 Intro.Rmd
update1
model_fit file

@@
 
-bank_full <- read_delim("bank-full.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
+bank_full <- read_delim("data/bank-full_1.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
 head(bank_full)
@@
 
-Plots for categorical variables
+### Plots for categorical variables
 
@@
 
-## We plan to use the k-fold cross validation approach. 
-For consideration: Are we able to make our training sets a sort of proper stratification of our population dataset (Imbalance in amounts of customers subscribed).
-
-```{r}
-bank_train <- 
-bank_test <- 
-```
-
-Model set up...
-
-## Lgoistic Regression
-
-```{r}
-library(glmnet)
-library(ggplot2)
-
-bank_glm <- glm(y ~., data = bank_train, family = binomial)
-summary(bank_glm)
-```
-
-estimate test error
-
-## LDA and QDA
-
-```{r}
-library(MASS)
-
-bank_lda <- lda(y ~., data = bank_full, subset = train)
-bank_lda
-
-bank_qda <- qda(y ~., data = bank_full, subset = train)
-```
-
-estimate test error
-
-## KNN
-
-```{r}
-library(class)
-
-bank_knn <- knn()
-```
-
-estimate test error
-
-Optimize K value (w.r.t cross validation tesst error)
-
-## GAM
-
-```{r}
-library(gam)
-bank_gam <- gam(y ~ , data = bank_train) # basis functions
-
-```
-
-## Boosting Decision Trees
-require post-hoc explanation
-
-```{r}
-
-```
-
-Potential model Mixed effects model (attribute set variables to have a random effect)
+*We don't need to do much on exploratory data analysis as this course is on machine learning*

### 2025-11-18 model_fit.Rmd
model_fit file

---
title: "model_fit"
output: html_document
---

load clean data

```{r}

```

## We plan to use the k-fold cross validation approach. 
For consideration: Are we able to make our training sets a sort of proper stratification of our population dataset (Imbalance in amounts of customers subscribed).

```{r}
bank_train <- 
bank_test <- 
```

## Lgoistic Regression

```{r}
library(glmnet)
library(ggplot2)

bank_glm <- glm(y ~., data = bank_train, family = binomial)
summary(bank_glm)
```

estimate test error

## LDA and QDA

```{r}
library(MASS)

bank_lda <- lda(y ~., data = bank_full, subset = train)
bank_lda

bank_qda <- qda(y ~., data = bank_full, subset = train)
```

estimate test error

## KNN
Interpretable in a different way... Can be used to identify customers similar to those that are subscirbed.

```{r}
library(class)

bank_knn <- knn()
```

estimate test error

Optimize K value (w.r.t cross validation test error)

## GAM

```{r}
library(gam)
bank_gam <- gam(y ~ , data = bank_train) # basis functions

```

## Boosting Decision Trees
require post-hoc explanation

```{r}

```

Potential model Mixed effects model (attribute set variables to have a random effect)

### 2025-11-19 Intro.Rmd
Added EDA visualizations and fixed knitting issues

--- 2025-11-18 Intro.Rmd
@@
 library(ggplot2)
+library(dplyr)
+library(tidyr)
 
@@
 ```{r}
+
+# Calculate and print the success rate
+success_rate <- mean(bank_full$y == "yes") * 100
+cat("Overall Subscription Success Rate:", round(success_rate, 2), "%\n")
+
+# bar plot for the target variable 'y'
+ggplot(bank_full, aes(x = y, fill = y)) +
+  geom_bar() +
+  labs(title = "Distribution of Term Deposit Subscription (y)",
+       x = "Subscribed to Term Deposit?",
+       y = "Count") +
+  theme_minimal()
+
+```
+
+# Convert variables to factors for plotting
+
+```{r}
+
+bank_full$y <- as.factor(bank_full$y)
+bank_full$job <- as.factor(bank_full$job)
+bank_full$marital <- as.factor(bank_full$marital)
+bank_full$education <- as.factor(bank_full$education)
+bank_full$contact <- as.factor(bank_full$contact)
+bank_full$month <- as.factor(bank_full$month)
+bank_full$poutcome <- as.factor(bank_full$poutcome)
+bank_full$housing <- as.factor(bank_full$housing)
+bank_full$loan <- as.factor(bank_full$loan)
+
+```
+
+```{r}
+
+ggplot(bank_full, aes(x = job, fill = y)) +
+  geom_bar(position = "fill") +
+  coord_flip() +
+  labs(title = "Subscription Rate by Job",
+       y = "Proportion") +
+  theme_minimal()
+
+```
+
+```{r}
+ggplot(bank_full, aes(x = education, fill = y)) +
+  geom_bar(position = "fill") +
+  coord_flip() +
+  labs(title = "Subscription Rate by Education Level",
+       y = "Proportion") +
+  theme_minimal()
+```
+
+```{r}
+
+ggplot(bank_full, aes(x = marital, fill = y)) +
+  geom_bar(position = "fill") +
+  coord_flip() +
+  labs(title = "Subscription Rate by Marital Status",
+       y = "Proportion") +
+  theme_minimal()
+
+```
+
+```{r}
+ggplot(bank_full, aes(x = contact, fill = y)) +
+  geom_bar(position = "fill") +
+  labs(title = "Subscription Rate by Contact Type",
+       y = "Proportion") +
+  theme_minimal()
+```
+
+```{r}
+ggplot(bank_full, aes(x = poutcome, fill = y)) +
+  geom_bar(position = "fill") +
+  labs(title = "Subscription Rate by Previous Marketing Campaign Outcome",
+       y = "Proportion") +
+  theme_minimal()
+```
+
+```{r}
+bank_full %>%
+  group_by(campaign) %>%
+  summarise(success_rate = mean(y == "yes")) %>%
+  ggplot(aes(x = campaign, y = success_rate)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Subscription Success Rate vs. Number of Campaign Contacts",
+       x = "Number of Contacts During This Campaign",
+       y = "Success Rate") +
+  theme_minimal()
+```
+
+```{r}
 summary(bank_full)
@@
 ```{r}
-bank_train <- 
-bank_test <- 
+#bank_train <- 
+#bank_test <- 
 ```
@@
 
-bank_glm <- glm(y ~., data = bank_train, family = binomial)
-summary(bank_glm)
+#bank_glm <- glm(y ~., data = bank_train, family = binomial)
+#summary(bank_glm)
 ```
@@
 
-bank_lda <- lda(y ~., data = bank_full, subset = train)
-bank_lda
+#bank_lda <- lda(y ~., data = bank_full, subset = train)
+#bank_lda
 
-bank_qda <- qda(y ~., data = bank_full, subset = train)
+#bank_qda <- qda(y ~., data = bank_full, subset = train)
 ```
@@
 ## KNN
+Interpretable in a different way... Can be used to identify customers similar to those that are subscirbed.
 
@@
 
-bank_knn <- knn()
+#bank_knn <- knn()
 ```
@@
 
-Optimize K value (w.r.t cross validation tesst error)
+Optimize K value (w.r.t cross validation test error)
 
@@
 library(gam)
-bank_gam <- gam(y ~ , data = bank_train) # basis functions
+#bank_gam <- gam(y ~ , data = bank_train) # basis functions
 

### 2025-11-20 Intro.Rmd
Updated plots, k-fold cross-validation, and some of the model setup

@@
 
+cat_vars <- c("job", "marital", "education", "contact", "month", "poutcome")
+
+for (var in cat_vars) {
+p <- ggplot(bank_full, aes_string(x = var, fill = "y")) +
+geom_bar(position = "fill") +
+coord_flip() +
+labs(
+title = paste("Proportion of Subscription by", var),
+x = var,
+y = "Proportion"
+) +
+theme_minimal()
+
+print(p) 
+}
+
 ```
@@
 #bank_test <- 
+library(caret)
+
+kfold_cv_model <- function(data, target_col, model_func, k = 5, seed = 123, ...) {
+set.seed(seed)
+
+folds <- createFolds(data[[target_col]], k = k, list = TRUE, returnTrain = TRUE)
+
+accuracy_list <- numeric(k)
+
+for (i in 1:k) {
+train_idx <- folds[[i]]
+train_data <- data[train_idx, ]
+test_data  <- data[-train_idx, ]
+model <- model_func(train_data, ...)
+
+pred_probs <- predict(model, newdata = test_data, type = "response")
+
+if (is.numeric(pred_probs)) {
+  pred <- ifelse(pred_probs > 0.5, "yes", "no")
+} else {
+  pred <- pred_probs
+}
+actual <- test_data[[target_col]]
+accuracy_list[i] <- mean(pred == actual)
+}
+mean_accuracy <- mean(accuracy_list)
+return(mean_accuracy)
+}
 ```
@@
 #summary(bank_glm)
+
+logistic_model <- function(train_data, ...) {
+glm(y ~ ., data = train_data, family = binomial)
+}
+
+accuracy_logit <- kfold_cv_model(bank_full, "y", logistic_model, k = 5)
+accuracy_logit
+
+cat("Logistic Regression 5-Fold CV Accuracy:", round(accuracy_logit * 100, 2), "%\n")
+
 ```

### 2025-12-02 Intro.Rmd
Add files via upload

@@
 #bank_qda <- qda(y ~., data = bank_full, subset = train)
+
+set.seed(123)
+train_index <- sample(seq_len(nrow(bank_full)), size = 0.8 * nrow(bank_full))
+bank_train <- bank_full[train_index, ]
+bank_test <- bank_full[-train_index, ]
+```
+
+```{r}
+lda_model <- lda(y ~ ., data = bank_train)
+lda_model
+```
+
+```{r}
+lda_pred <- predict(lda_model, newdata = bank_test)$class
+lda_actual <- bank_test$y
+```
+
+```{r}
+lda_accuracy <- mean(lda_pred == lda_actual)
+lda_error <- 1 - lda_accuracy
+cat("LDA Test Accuracy:", round(lda_accuracy * 100, 2), "%\n")
+cat("LDA Test Error:", round(lda_error * 100, 2), "%\n")
+```
+
+```{r}
+qda_model <- qda(y ~ ., data = bank_train)
+qda_model
+```
+
+```{r}
+qda_pred <- predict(qda_model, newdata = bank_test)$class
+qda_actual <- bank_test$y
+```
+
+```{r}
+qda_accuracy <- mean(qda_pred == qda_actual)
+qda_error <- 1 - qda_accuracy
+cat("QDA Test Accuracy:", round(qda_accuracy * 100, 2), "%\n")
+cat("QDA Test Error:", round(qda_error * 100, 2), "%\n")
 ```
@@
 
+```{r}
+numeric_vars <- bank_full %>%
+select(-y) %>%
+select(where(is.numeric)) %>%
+names()
+train_x <- bank_train[, numeric_vars]
+test_x <- bank_test[, numeric_vars]
+```
+
+```{r}
+train_x_scaled <- scale(train_x)
+test_x_scaled <- scale(
+  test_x,
+  center = attr(train_x_scaled, "scaled:center"),
+  scale  = attr(train_x_scaled, "scaled:scale")
+)
+train_y <- bank_train$y
+test_y  <- bank_test$y
+```
+
+```{r}
+knn_pred <- knn(train = train_x_scaled,
+test = test_x_scaled,
+cl = train_y,
+k = 5)
+```
+
+```{r}
+knn_accuracy <- mean(knn_pred == test_y)
+knn_error <- 1 - knn_accuracy
+cat("KNN Test Accuracy:", round(knn_accuracy * 100, 2), "%\n")
+cat("KNN Test Error:", round(knn_error * 100, 2), "%\n")
+```
+
+```{r}
+k_values <- 1:25
+cv_accuracy <- numeric(length(k_values))
+folds <- createFolds(train_y, k = 5)
+for (i in seq_along(k_values)) {
+k <- k_values[i]
+fold_acc <- c()
+for (f in folds) {
+fold_train_x <- train_x_scaled[-f, ]
+fold_test_x <- train_x_scaled[f, ]
+fold_train_y <- train_y[-f]
+fold_test_y <- train_y[f]
+pred <- knn(
+  train = fold_train_x,
+  test  = fold_test_x,
+  cl    = fold_train_y,
+  k     = k
+)
+
+fold_acc <- c(fold_acc, mean(pred == fold_test_y))
+}
+cv_accuracy[i] <- mean(fold_acc)
+}
+best_k <- k_values[which.max(cv_accuracy)]
+best_k
+```
+
+```{r}
+plot(k_values, cv_accuracy, type = "b",
+xlab = "K Value", ylab = "Cross-Validated Accuracy",
+main = "K Optimization for KNN")
+abline(v = best_k, col = "red", lty = 2)
+```
+
+```{r}
+knn_best_pred <- knn(
+train = train_x_scaled,
+test = test_x_scaled,
+cl = train_y,
+k = best_k
+)
+final_accuracy <- mean(knn_best_pred == test_y)
+final_error <- 1 - final_accuracy
+cat("Best K:", best_k, "\n")
+cat("Test Accuracy with Best K:", round(final_accuracy * 100, 2), "%\n")
+cat("Test Error with Best K:", round(final_error * 100, 2), "%\n")
+```
+
 estimate test error
@@
 #bank_gam <- gam(y ~ , data = bank_train) # basis functions
+```
 
+```{r}
+gam_model <- gam(
+y ~ s(age) + s(balance) + s(duration) + s(campaign) + s(pdays),
+family = binomial,
+data = bank_train
+)
+summary(gam_model)
+```
+
+```{r}
+gam_prob <- predict(gam_model, newdata = bank_test, type = "response")
+gam_pred <- ifelse(gam_prob > 0.5, "yes", "no")
+gam_pred <- as.factor(gam_pred)
+```
+
+```{r}
+actual <- bank_test$y
+gam_accuracy <- mean(gam_pred == actual)
+gam_error <- 1 - gam_accuracy
+cat("GAM Test Accuracy:", round(gam_accuracy * 100, 2), "%\n")
+cat("GAM Test Error:", round(gam_error * 100, 2), "%\n")
 ```
@@
 require post-hoc explanation
+```{r}
+library(gbm)
+```
+```{r}
+bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
+bank_test$y_num <- ifelse(bank_test$y == "yes", 1, 0)
+set.seed(123)
+boost_model <- gbm(
+formula = y_num ~ . - y,
+data = bank_train,
+distribution = "bernoulli",
+n.trees = 1000,
+interaction.depth = 3,
+shrinkage = 0.01,
+n.minobsinnode = 10,
+verbose = FALSE
+)
+```
 
 ```{r}
-
+boost_prob <- predict(boost_model, newdata = bank_test, n.trees = 1000, type = "response")
+boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")
 ```
 
+```{r}
+boost_accuracy <- mean(boost_pred == bank_test$y)
+boost_error <- 1 - boost_accuracy
+cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
+cat("Boosting Test Error:", round(boost_error * 100, 2), "%\n")
+summary(boost_model)
+```
+Post-hoc explanation: Boosting combines the small decision trees to not make the same mistake the previous one has made. Which creates a very good and accurate model.
+
+Ω
 Potential model Mixed effects model (attribute set variables to have a random effect)

### 2025-12-03 Intro.Rmd
split intro file

--- 2025-11-19 Intro.Rmd
@@
 
-bank_full <- read_delim("bank-full.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
+bank_full <- read_delim("data/bank-full_1.csv", delim = ';', escape_double = FALSE, trim_ws = TRUE)
 head(bank_full)
@@
 
+What does this mean? ^
+
 ```{r}
@@
 
-```
+cat_vars <- c("job", "marital", "education", "contact", "month", "poutcome")
 
-## We plan to use the k-fold cross validation approach. 
-For consideration: Are we able to make our training sets a sort of proper stratification of our population dataset (Imbalance in amounts of customers subscribed).
+for (var in cat_vars) {
+p <- ggplot(bank_full, aes_string(x = var, fill = "y")) +
+geom_bar(position = "fill") +
+coord_flip() +
+labs(
+title = paste("Proportion of Subscription by", var),
+x = var,
+y = "Proportion"
+) +
+theme_minimal()
 
-```{r}
-#bank_train <- 
-#bank_test <- 
-```
-
-Model set up...
-
-## Lgoistic Regression
-
-```{r}
-library(glmnet)
-library(ggplot2)
-
-#bank_glm <- glm(y ~., data = bank_train, family = binomial)
-#summary(bank_glm)
-```
-
-estimate test error
-
-## LDA and QDA
-
-```{r}
-library(MASS)
-
-#bank_lda <- lda(y ~., data = bank_full, subset = train)
-#bank_lda
-
-#bank_qda <- qda(y ~., data = bank_full, subset = train)
-```
-
-estimate test error
-
-## KNN
-Interpretable in a different way... Can be used to identify customers similar to those that are subscirbed.
-
-```{r}
-library(class)
-
-#bank_knn <- knn()
-```
-
-estimate test error
-
-Optimize K value (w.r.t cross validation test error)
-
-## GAM
-
-```{r}
-library(gam)
-#bank_gam <- gam(y ~ , data = bank_train) # basis functions
+print(p) 
+}
 
 ```
-
-## Boosting Decision Trees
-require post-hoc explanation
-
-```{r}
-
-```
-
-Potential model Mixed effects model (attribute set variables to have a random effect)

### 2025-12-03 draft_group-6.Rmd
test
split intro file

@@
 
-## results
-(25%)
+## 3. Data and Preprocessing
+3.1 Dataset Description
 
-- Model comparison and assessment (Test error, confusion matrix, AUC, BIC, etc.)
-- Discussion (make inferrences, determine the favored model)
+Source: UCI Machine Learning Repository.
 
-## references
-(2%)
+Number of observations and predictors.
+
+Summary of variable types (categorical, numeric, campaign history).
+
+3.3 Exploratory Data Analysis
+
+Distribution of the response variable (class imbalance).
+
+Univariate summaries: histograms, bar plots.
+
+Key relationships between predictors and response.
+
+## 4. Methodology
+4.1 Overview of Statistical Learning Framework
+
+Supervised binary classification.
+
+Tradeoff between interpretability and predictive power.
+
+4.2 Resampling Methods (ISLR Ch. 5)
+4.2.1 Train/Test Split
+
+Rationale for honest evaluation.
+
+4.2.2 k-Fold Cross-Validation
+
+Applied to Logistic Regression, KNN, and GAM (if included).
+
+Used for model tuning and error estimation.
+
+4.2.3 Bootstrap (Optional)
+
+Estimating coefficient variability in Logistic Regression.
+
+4.3 Candidate Models (ISLR Ch. 4 & 6)
+
+Logistic Regression
+
+Linear Discriminant Analysis (LDA)
+
+Quadratic Discriminant Analysis (QDA)
+
+K-Nearest Neighbors (KNN) with CV-selected k
+
+Generalized Additive Models (GAM) (optional but interpretable)
+
+4.4 Assessment Metrics (ISLR Ch. 6)
+
+Test misclassification rate.
+
+Sensitivity & specificity.
+
+Confusion matrix.
+
+ROC curve and AUC.
+
+BIC (when applicable).
+
+## 5. Results
+5.1 Cross-Validation Findings
+
+Optimal tuning parameters (e.g., best k for KNN).
+
+CV-estimated test errors for each model.
+
+5.2 Test Set Performance
+
+Summary table: accuracy, sensitivity, specificity, AUC.
+
+ROC curves for model comparison.
+
+5.3 Interpretation of Key Predictors
+
+Coefficient signs and magnitudes (Logistic).
+
+Group separation interpretation (LDA/QDA).
+
+Partial dependence/smooth terms (GAM).
+
+## 6. Discussion
+- Which model offers the best predictive accuracy?
+- Which model is most interpretable for a marketing manager?
+- How did resampling methods influence the final choice?
+- Effects of class imbalance.
+- Limitations and suggested future modeling directions.
+
+Future directions: oversampling methods, modern ML models.

### 2025-12-03 model_fit.Rmd
split intro file

@@
 
-## We plan to use the k-fold cross validation approach. 
-For consideration: Are we able to make our training sets a sort of proper stratification of our population dataset (Imbalance in amounts of customers subscribed).
+## 2.3 Train/Test split
+- Explain purpose of an honest out-of-sample evaluation.
+- k-fold CV, k=5
 
-```{r}
-bank_train <- 
-bank_test <- 
-```
+---
 
-## Lgoistic Regression
+# 3. Modeling
 
-```{r}
-library(glmnet)
-library(ggplot2)
+## 3.1 Logistic Regression (Baseline Model)
+- Fit logistic regression.
+- k-fold CV for estimated test error.
+- Assess linearity and consider interactions if appropriate.
+- Optional: **regularized logistic regression** (Ridge/Lasso) using CV-selected $\lambda$.
 
-bank_glm <- glm(y ~., data = bank_train, family = binomial)
-summary(bank_glm)
-```
+## 3.2 Linear Discriminant Analysis (LDA)
+- Fit LDA model.
+- k-fold CV for estimated test error.
+- Check assumptions (normality, equal covariance).
+- Discuss bias-variance
 
-estimate test error
+## 3.3 Quadratic Discriminant Analysis (QDA)
+- Fit QDA model.
+- k-fold CV for estimated test error.
+- Discuss overfitting risk with higher model flexibility.
+- Compare performance with LDA.
 
-## LDA and QDA
+## 3.4 K-Nearest Neighbors (KNN)
+- Standardize numeric predictors.
+- Perform CV to tune **k** .
+- Report best k and corresponding CV error.
+- Discuss bias-variance behavior.
 
-```{r}
-library(MASS)
+## 3.5 Generalized Additive Models (GAM)
+- Fit GAM with smoothing splines for continuous variables.
+- Use CV to choose smoothing parameters.
+- Examine partial effect plots for interpretability.
 
-bank_lda <- lda(y ~., data = bank_full, subset = train)
-bank_lda
+## 3.6 Boosting (Most Flexible Model)
+- Fit gradient boosting classifier (e.g., `gbm` or `xgboost`).
+- Use cross-validation for:
+  - number of trees,
+  - interaction depth,
+  - learning rate.
+- Discuss complexity vs interpretability tradeoff. This model will be our benchmark non-interpretable model to compare interpretable models fitted above.
 
-bank_qda <- qda(y ~., data = bank_full, subset = train)
-```
+---
 
-estimate test error
+# 4. Results — Model Assessment & Comparison
 
-## KNN
-Interpretable in a different way... Can be used to identify customers similar to those that are subscirbed.
+## 4.1 Test Set Performance
+For each model:
+- Confusion matrix  
+  - Accuracy and misclassification rate  
+  - Sensitivity / specificity (important for minority class)  
+- ROC curves and AUC
 
-```{r}
-library(class)
-
-bank_knn <- knn()
-```
-
-estimate test error
-
-Optimize K value (w.r.t cross validation test error)
-
-## GAM
-
-```{r}
-library(gam)
-bank_gam <- gam(y ~ , data = bank_train) # basis functions
-
-```
-
-## Boosting Decision Trees
-require post-hoc explanation
-
-```{r}
-
-```
-
-Potential model Mixed effects model (attribute set variables to have a random effect)
+## 4.2 Comparative Evaluation
+- Table summarizing all test-set performance metrics.
+- Visual comparison of ROC curves.
+- Discuss interpretability vs predictive quality.

### 2025-12-03 Intro.Rmd
Add descriptions for visualization figures

@@
 
+**Description:** A simple bar chart showing the imbalance between clients who subscribed to a term deposit (“yes”) and those who did not (“no”), with the majority belonging to the “no” category.
+
 # Convert variables to factors for plotting
@@
 
+**Description:** A stacked bar chart illustrating the proportion of subscription outcomes across different job types, with retired and student occupations showing relatively higher subscription rates.
+
 ```{r}
@@
 ```
+
+**Description:** A proportional bar chart comparing subscription rates across education levels, indicating that clients with tertiary education have a higher tendency to subscribe.
 
@@
 
+**Description:** A stacked bar chart showing subscription proportions by marital status, revealing that single clients have a slightly higher subscription rate compared to married or divorced clients.
+
 ```{r}
@@
 
+**Description:** A proportional bar chart comparing subscription outcomes by contact method, with cellular contacts showing a notably higher subscription rate than telephone or unknown contact types.
+
 ```{r}
@@
 ```
+
+**Description:** A stacked bar chart depicting how the outcome of previous marketing campaigns influences current subscription rates, with previous “success” outcomes leading to higher subscription proportions.
 
@@
 
-What does this mean? ^
+**Description:** A line plot with points showing how the subscription success rate declines as the number of contacts during the campaign increases, suggesting diminishing returns with repeated contact attempts.
 

### 2025-12-05 model_fit.Rmd
update2

@@
 
-load clean data
+load libraries and clean data
+
+```{r}
+library(readr)
+library(tidyverse)
+library(boot)
+library(glmnet)
+library(caret)
+library(pROC)
+library(ggplot2)
+
+set.seed(445)
+
+bank_full <- read_delim("data/bank-full_1.csv", 
+    delim = ";", escape_double = FALSE, trim_ws = TRUE)
+```
+
+```{r}
+bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))
+
+bank_full <- bank_full %>%
+  mutate(
+    job = as.factor(job),
+    marital = as.factor(marital),
+    education = as.factor(education),
+    default = as.factor(default),
+    housing = as.factor(housing),
+    loan = as.factor(loan),
+    contact = as.factor(contact),
+    month = as.factor(month),
+    poutcome = as.factor(poutcome))
+
+```
+
+## Train/Test split
+- Explain purpose of an honest out-of-sample evaluation.
+- k-fold CV, k=5
+
+```{r}
+bank_full$previously_contacted <- factor(ifelse(bank_full$pdays == -1, "No", "Yes"), levels = c("No", "Yes"))
+bank_full$pdays_non_miss <- ifelse(bank_full$pdays == -1, NA, bank_full$pdays)
+
+train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
+train_data <- bank_full[train_index, ]
+test_data <- bank_full[-train_index, ]
+```
+
+## subset selection
+- Backward step-wise elimination (ecalutes fit using AIC)
+
+```{r}
+initial_formula <- y ~ age + job + marital + education + default + balance + housing + loan + 
+  contact + day + month + campaign + pdays + previous + poutcome + previously_contacted
+
+# Fit the model with maximum covaraints on the training set
+glm.full <- glm(initial_formula, data = train_data, family = binomial)
+
+# Use the step function to perform backward elimination based on AIC
+glm.selected <- step(
+  glm.full, 
+  direction = "backward", 
+  trace = 0
+)
+
+selected_formula <- formula(glm.selected)
+selected_formula
+
+```
 
@@
 
+interact_formula <- update(selected_formula, . ~ .^2)
+
+# Fit the maximal interaction model (CAUTION: This can be slow and unstable with many factors)
+glm.full.interact <- glm(interact_formula, data = train_data, family = binomial)
+glm.full.interact.summary <- summary(glm.full.interact)
 ```
 
-## 2.3 Train/Test split
-- Explain purpose of an honest out-of-sample evaluation.
-- k-fold CV, k=5
+```{r}
+coef_table <- glm.full.interact.summary$coefficients
+coef_df <- as.data.frame(coef_table)
+
+view(coef_df %>%  filter(`Pr(>|z|)`< 0.01))
+```
+
+job:contact
+job:month
+education:housing
+housing:month
+contact:month
+day:month
+month:poutcome
+
+```{r}
+interact_formula_significant <- y ~ job + marital + education + balance + housing + loan + contact + 
+  day + month + campaign + previous + poutcome + job:contact + job:month + education:housing + housing:month +
+  contact:month + day:month + month:poutcome
+
+glm.full.interact_significant <- glm(interact_formula_significant, data = train_data, family = binomial)
+
+glm.interact.selected <- step(
+  glm.full.interact_significant, 
+  direction = "backward", 
+  trace = 1 
+)
+
+selected_interact_formula <- formula(glm.interact.selected)
+selected_interact_formula
+```
 
@@
 
-## 3.2 Linear Discriminant Analysis (LDA)
-- Fit LDA model.
-- k-fold CV for estimated test error.
-- Check assumptions (normality, equal covariance).
-- Discuss bias-variance
+*Standard raw logistic reg (without tuning)*
 
-## 3.3 Quadratic Discriminant Analysis (QDA)
-- Fit QDA model.
-- k-fold CV for estimated test error.
-- Discuss overfitting risk with higher model flexibility.
-- Compare performance with LDA.
+```{r}
+glm_out <- glm(y ~ ., data = bank_full, family = binomial)
+summary(glm_out)
+```
+
+Discuss accuracy (cv test error) and amount of parameters
+
+```{r}
+glm.full.cv <- glm(initial_formula, data = data, family = binomial, x = TRUE)
+
+cv.error.full <- cv.glm(data, glm.full.cv, K = 5)$delta[2]
+length(coef(glm.full.cv))
+```
+
+This model is complex due to the large number of dummy variables.
+The CV test error provides an estimate of how well this complexity generalizes to new, out of sample data.
+If the CV error is significantly higher than the training error, the model is likely overfitting
+
+*With step-wise feature selection* (our base line)
+
+```{r}
+# Re-fit the final selected model
+glm.baseline <- glm(selected_formula, data = train_data, family = binomial)
+summary(glm.baseline)
+
+```
+
+```{r}
+glm_out <- glm(y ~ job + marital + education + balance + housing + loan + contact + 
+    day + month + campaign + previous + poutcome,
+    data = bank_full,
+    family = binomial, # Or whatever family you used
+    x = TRUE,
+    model = TRUE
+)
+
+cv.error.baseline <- cv.glm(
+    bank_full, #only include features that are fit
+    glm_out, K = 5)
+
+cv.error.baseline$delta[2]
+```
+
+```{r}
+glm.probs.baseline <- predict(glm.baseline, newdata = test_data, type = "response")
+
+glm.pred.baseline <- factor(ifelse(glm.probs.baseline > 0.5, "yes", "no"), levels = c("no", "yes"))
+confusionMatrix(glm.pred.baseline, test_data$y, positive = "yes")
+
+```
+
+- analysis of parameters and example predictions with interpretation
+
+residual plots (numeric features)... this will lead to inclusion of non-linear features
+
+```{r}
+glm.resids <- residuals(glm.full, type = "deviance")
+
+# Function to plot residuals vs a numeric predictor
+plot_residuals_vs_predictor <- function(predictor_name, data, residuals) {
+  df <- data.frame(
+    Residuals = residuals,
+    Predictor = data[[predictor_name]]
+  ) %>% filter(!is.na(Predictor))
+
+  p <- ggplot(df, aes(x = Predictor, y = Residuals)) +
+    geom_point(alpha = 0.3) +
+    geom_smooth(method = "loess", color = "red", se = FALSE) + # Add a smoothing line
+    geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+    labs(
+      title = paste("Deviance Residuals vs.", predictor_name),
+      y = "Deviance Residuals",
+      x = predictor_name
+    ) +
+    theme_minimal()
+  print(p)
+}
+
+# Plot residuals for key numeric features
+plot_residuals_vs_predictor("age", train_data, residuals(glm.full, type = "deviance"))
+plot_residuals_vs_predictor("balance", train_data, residuals(glm.full, type = "deviance"))
+```
+
+*polynomials logit reg (on Balance or other numeric variables)*
+- test models with transformations (polynomial or Log)
+- test models with interaction terms
+
+```{r}
+
+```
+
+```{r}
+glm.test.linearity <- glm(y ~ job + marital + education + balance + housing + loan + contact + 
+    day + month + campaign + previous + poutcome + I(balance^2), data = train_data, family = binomial)
+
+summary(glm.test.linearity)$coefficients["I(balance^2)", "Pr(>|z|)"]
+
+# Interaction: Test interaction between 'age' and 'campaign' (common drivers).
+glm.test.interaction <- glm(selected_formula, data = train_data, family = binomial)
+glm.test.interaction <- update(glm.test.interaction, . ~ . + age:campaign)
+
+print(anova(glm.full, glm.test.interaction, test = "Chisq"))
+```
+
+*Regularized logit reg*
+
+Ridge
+- test across lambda values (and plot) with mean CV test error
+- confussion matrix and roc curve
+
+```{r}
+full_predictors_formula <- y ~ age + job + marital + education + default + balance + housing + loan + 
+  contact + day + month + campaign + pdays + previous + poutcome + previously_contacted
+
+X_train_ridge <- model.matrix(full_predictors_formula, data = train_data)[, -1] 
+y_train_ridge <- train_data$y
+
+```
+
+k-fold CV for optimal lambda
+
+```{r}
+cv.ridge.fit <- cv.glmnet(
+  X_train_ridge, 
+  y_train_ridge, 
+  alpha = 0, 
+  family = "binomial", 
+  nfolds = 5
+)
+
+lambda_min <- cv.ridge.fit$lambda.min
+lambda_min
+```
+
+fit using optimal lambda
+
+```{r}
+X_test_ridge <- model.matrix(full_predictors_formula, data = test_data)[, -1]
+
+ridge.probs <- predict(cv.ridge.fit, newx = X_test_ridge, s = lambda_min, type = "response") %>%
+  as.vector()
+
+plot(cv.ridge.fit)
+
+ridge.pred <- factor(ifelse(ridge.probs > 0.5, "yes", "no"), levels = c("no", "yes"))
+confusionMatrix(ridge.pred, test_data$y, positive = "yes")
+```
+
+LASSO
+- test across lambda values (and plot) with mean CV test error
+- confussion matrix and roc curve
+
+```{r}
+X_train_lasso <- X_train_ridge
+y_train_lasso <- y_train_ridge
+```
+
+k-fold CV for optimal lambda
+
+```{r}
+cv.lasso.fit <- cv.glmnet(
+  X_train_lasso, 
+  y_train_lasso, 
+  alpha = 1, 
+  family = "binomial", 
+  nfolds = 5
+)
+
+lambda_min_lasso <- cv.lasso.fit$lambda.min
+lambda_min_lasso
+
+plot(cv.lasso.fit)
+title("LASSO CV Error vs. Log(Lambda)", line = 2.5)
+abline(v = log(lambda_min_lasso), col = "red", lty = 2)
+
+```
+
+fit using optimal lambda
+
+```{r}
+X_test_lasso <- X_test_ridge # Use the same test design matrix
+
+lasso.probs <- predict(cv.lasso.fit, newx = X_test_lasso, s = lambda_min_lasso, type = "response") %>% 
+  as.vector()
+
+lasso.pred <- factor(ifelse(lasso.probs > 0.5, "yes", "no"), levels = c("no", "yes"))
+confusionMatrix(lasso.pred, test_data$y, positive = "yes")
+
+```
+
+*plot ROC curves and table confusion matrix for the above models*
+
+```{r}
+roc_baseline <- roc(test_data$y, glm.probs.baseline, direction = "<")
+roc_ridge <- roc(test_data$y, ridge.probs, direction = "<")
+
+roc_data <- tibble(
+  Specificity = c(roc_baseline$specificities, roc_ridge$specificities), # include the raw logit model
+  Sensitivity = c(roc_baseline$sensitivities, roc_ridge$sensitivities),
+  Model = c(rep(paste0("Baseline (AUC=", round(auc(roc_baseline), 3), ")"), length(roc_baseline$specificities)), 
+            rep(paste0("Ridge (AUC=", round(auc(roc_ridge), 3), ")"), length(roc_ridge$specificities)))
+)
+
+# Plot the ROC curves 
+ggplot(roc_data, aes(x = 1 - Specificity, y = Sensitivity, color = Model)) +
+  geom_line(linewidth = 1) +
+  geom_abline(lty = 2, color = "gray50") + 
+  labs( title = "ROC Curve Comparison: Baseline vs. Ridge Logit",
+    x = "False Positive Rate (1 - Specificity)",
+    y = "True Positive Rate (Sensitivity)",
+    color = "Model") +
+  theme_minimal(base_size = 14)
+
+```
+
+---
 
@@
 
----
-
 # 4. Results — Model Assessment & Comparison

### 2025-12-06 draft_group-6.Rmd
update2
finished draft_group-6 Rmd file

--- 2025-11-13 draft_group-6.Rmd
@@
 title: "draft_group-6"
-output: pdf_document
+author: "Sandra, Tigist, Mussa, Waddah"
+output:
+  pdf_document: default
+  html_document:
+    df_print: paged
 ---
@@
 
-## motivation
-(10% of the content)\
+## Motivation
+(13% of the content)\
 The Bank Marketing dataset contains information collected from a Portuguese banking institution during multiple direct telemarketing campaigns (source1).
-*Why??* (We want to justify the efforts of modeling this data)
+*Why??* (We want to justify the efforts of modeling this data, then whwy it needs to be accurate)
 
 **Description of dataset** \
-Each row represents an individual client that was contacted in an attempt to sell a long-term deposit product.
-The dataset includes 16 predictor variables describing client demographics, financial attributes, and details about the marketing campaign itself (such as: contact method, call duration, number of contacts, outcome of previous campaigns).
-These variables are a mix of categorical and quantitative data, which makes this dataset well-suited for comparing multiple classification models.
-The target variable indicates whether the client subscribed to a term deposit (yes/no).
-As usual, the majority of clients do not subscribe (source 2).
-This causes an imbalance in our data by increasing modeling bias.
+Each observation in the dataset corresponds to a single client contacted during the bank’s telemarketing campaigns. The dataset includes 16 predictor variables, which cover:
 
-**Problem statement/Assumptions** \
-We can go over what we are focusing on and what factors we may ignore.
+- Demographics: age, occupation, marital status, education
+- Financial indicators: loan status, credit history, average account balance
+- Marketing campaign features: type of contact, month/day of contact, call duration
+- Past campaign outcomes: number of previous contacts and whether they were successful
 
-We would like to maintain interpretability in our model. Emphasis is predicated for the model to provide actionable insght to the manager. After all humans love stories.
+These variables include both categorical and quantitative features, making the dataset suitable for comparing a variety of classification methods.
 
-## methodology 
-(60%)
+The target variable, y, is a binary indicator specifying whether the client subscribed to a term deposit (yes/no).
+As is typical in marketing data, the positive class (“yes”) is rare, meaning most clients do not subscribe (source 2).
+This creates a class imbalance, which can bias models toward predicting the majority class unless addressed appropriately.
 
-- Data wrangling/cleaning
-- For each model (Logistic, LDA, QDA, KNN, Boosting )
-  - Model Formulation (training data, with k-fold CV)
-  - Feature Selection and model tuning (subset, regularization, dimension reduction)
+**Problem Statement/Assumptions** \
+The goal of this project is to build statistical models that can predict whether a client will subscribe to a term deposit based on their demographic, financial, and campaign-related characteristics.
+This is a binary classification problem.
 
-## results
-(25%)
+Our focus is on models that provide both good predictive performance and interpretability, since managers need results they can explain and translate into marketing strategies.
+Interpretability is especially valuable in domains such as banking, where decisions must be justified for ethical, customer-centric, and regulatory reasons.
 
-- Model comparison and assessment (Test error, confusion matrix, AUC, BIC, etc.)
-- Discussion (make inferrences, determine the favored model)
+To keep the scope focused, we make several simplifying assumptions:
 
-## references
-(5%)
+- We treat all observations as independent, even though clients may appear in multiple campaigns.
+- We do not use time-series methods, even though campaign month/day could be part of a temporal structure.
+- We prioritize models covered in class, especially those that illustrate differences in linearity, flexibility, and underlying assumptions (e.g., covariance structure for LDA and QDA).
+- We assume the data used for modeling reflects the population targeted by the campaign, meaning the model can be applied to future clients with similar characteristics.
+
+These choices allow us to compare several classification approaches under a consistent framework.
+
+## Methodology
+
+#### Data Wrangling and Cleaning:
+
+We began by loading the dataset, converting categorical variables into factors, and identifying “unknown” levels used in place of missing data.
+We evaluated several approaches to handling these values—including keeping “unknown” as its own category and treating them as missing, but ultimately selected the method that preserved the most information without distorting class balance.
+
+A train/test split was created, and all models were trained using the same set of cleaned predictors for fair comparison.
+
+### Models Considered:
+
+We evaluated five main models from the course:
+
+- Logistic Regression — A linear classifier that estimates the log-odds of subscription. It is highly interpretable and provides coefficients that can be directly translated into marketing insights.
+
+- Linear Discriminant Analysis (LDA) — A generative model that assumes equal covariance matrices for each class and normally distributed predictors. LDA often performs well in structured tabular data.
+
+- Quadratic Discriminant Analysis (QDA) — Similar to LDA, but allows each class to have its own covariance matrix, giving the model more flexibility.
+- K-Nearest Neighbors (KNN) — A nonparametric method that makes predictions by finding the k most similar clients.
+It captures nonlinear relationships but can be sensitive to scaling and high-dimensional data.
+
+- Boosting (e.g., AdaBoost or Gradient Boosting) — An ensemble method that sequentially builds weak learners to minimize classification error.
+Boosting is often superior in predictive accuracy but is less interpretable.
+
+### Model Training, Formulation, and Validation
+
+All models were trained using k-fold cross-validation to estimate out-of-sample performance. This approach reduces variance and ensures that results do not depend on a single train/test split.
+
+We also carried out:
+- Feature selection using stepwise regression and subset selection for logistic regression
+- Regularization techniques (ridge/lasso) when appropriate
+- Dimensionality reduction (e.g., PCA) if needed to improve KNN performance
+- Hyperparameter tuning for KNN (choice of k), QDA (regularization), and boosting (learning rate, number of trees)
+
+This section of the paper will include the formal model equations and the details of the validation procedure.
+
+## Results
+
+Once our models are trained, we will compare their performance using test error rate, confusion matrices, and metrics such as accuracy, recall, and AUC.
+Because the dataset is imbalanced, we expect recall and AUC to be especially important for evaluating how well each model identifies clients who subscribe.
+
+We will also create ROC curves to compare models visually and use information criteria (AIC/BIC) for parametric models where appropriate.
+After summarizing these results, we will determine which model performs best overall and which provides the most useful insights for guiding future marketing decisions
+
+## Discussion
+
+Since we are still exploring the dataset and fitting our models, we have not yet determined which classification method will perform best.
+However, based on the structure of the Bank Marketing data and findings from similar studies, we expect that boosting is likely to give the best predictive accuracy, while logistic regression will likely be the most interpretable and actionable for managers.
+
+In the final version of this section, we will compare the models based on their test performance, interpretability, and robustness to class imbalance.
+We will also discuss any limitations of our approach and suggest future improvements, such as trying resampling methods or evaluating additional machine learning models.
+
+## References
+
+Dua, D., & Graff, C. (2019). UCI Machine Learning Repository: Bank Marketing Dataset. 
+https://archive.ics.uci.edu/dataset/222/bank+marketing

### 2025-12-06 draft_group-6.Rmd
updated final draft

@@
 
-- Linear Discriminant Analysis (LDA) — A generative model that assumes equal covariance matrices for each class and normally distributed predictors. LDA often performs well in structured tabular data.
-
-- Quadratic Discriminant Analysis (QDA) — Similar to LDA, but allows each class to have its own covariance matrix, giving the model more flexibility.
 - K-Nearest Neighbors (KNN) — A nonparametric method that makes predictions by finding the k most similar clients.

### 2025-12-06 draft_group-6.Rmd
final finished updated draft

@@
 - Dimensionality reduction (e.g., PCA) if needed to improve KNN performance
-- Hyperparameter tuning for KNN (choice of k), QDA (regularization), and boosting (learning rate, number of trees)
+- Hyperparameter tuning for KNN (choice of k)
 

### 2025-12-07 Slides.Rmd
slides update

---
title: "Statistical Machine Learning Models on A Bank Marketing dataset"
subtitle: "Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee Kaplan"
author: "Sandra, Tigist, Mussa Hassen, Waddah"
output: ioslides_presentation
---

## Project Motivation & Business Problem
* **Discuss:** The bank's need for efficient marketing and the high cost of telemarketing campaigns. The goal is to maximize the success rate by predicting customer subscription.
* **Discuss:** The challenge of less positive class (imbalanced data) and why prediction accuracy is crucial.
* **Image/Plot Idea:** A simple, high-level info graphic of a marketing campaign funnel, highlighting the bottleneck/drop-off point.

[Image of a marketing funnel illustrating customer conversion]

## The Bank Marketing Dataset (Slide 2)
* **Discuss:** The dataset structure: $\approx 45,000$ client observations from a Portuguese banking institution's telemarketing campaigns.
* **Discuss:** Brief overview of the four main predictor categories: Client Demographics, Financial Indicators, Campaign Features, and Past Outcomes. Mention the paper had more features, fro privacy reasons the public data set has less
* **Discuss:** The Target Variable ($y$) is binary: Did the client subscribe to a term deposit (Yes/No)?
* **Discuss:** Emphasize the severity of the class imbalance (e.g., typically $\approx 11\%$ 'Yes' subscriptions). This may yield higher false negatives.
* **Image/Plot Idea:** A small table summarizing the list of the 4 key variable groups (e.g., Demographics, Financial, Campaign, Past). 
* **Image/Plot Idea:** A **Bar Plot** showing the distribution of the target variable ($y$), visually highlighting the imbalance.

## Problem Statement & Methodology Overview (Slide 3)
* **Discuss:** The project aims for a **Comparative Study**—comparing models based on two criteria: **Predictive Power** and **Interpretability**.
* **Discuss:** The project goal is **Binary Classification**: Building statistical models that are both **predictive** and **interpretable** for management.
* **Discuss:** Established an **80/20 Train/Test Split** and used **5-Fold Cross-Validation (CV)** on the training set for tuning and estimating out-of-sample error.
* **Image/Plot Idea:** An icon representing **Predictive Power** next to an icon representing **Interpretability** (The two core goals).

## Methodology: Data Preparation (Slide 5)
* **Discuss:** **Data Wrangling** involved converting categorical variables to factors and strategies for handling 'unknown' levels (treating them as a separate category vs. missing).
* **Image/Plot Idea:** A simple flow chart showing: Raw Data $\rightarrow$ Wrangling/Cleaning $\rightarrow$ Train/Test Split. 
* **Image/Plot Idea:** Summary stats or informational plots

[Image of a data science workflow diagram]

## LogReg 1/3: Feature Selection & Linearity  #MUSSA
* **Discuss:** Initial feature selection was performed using **Backward Stepwise Elimination** (based on **AIC**) on the full model to find the most parsimonious set of main effects.
* **Discuss:** **Deviance Residual Plots** were used to assess the linearity assumption for numeric predictors (e.g., `age`, `balance`). This led to testing non-linear terms.
* **Image/Plot Idea:** The **Residual Plot** for a key numeric feature (e.g., `balance`), showing a curve suggesting non-linearity.

## LogReg 2/3: Regularization & Tuning   #MUSSA
* **Discuss:** We explored **Ridge ($\alpha=0$)** and **LASSO ($\alpha=1$) Regularization** to manage model complexity and feature selection, particularly with the large number of dummy variables.
* **Discuss:** **5-Fold Cross-Validation** was used to select the optimal penalty parameter, $\lambda$, which minimized the mean CV error for each regularized model.
* **Image/Plot Idea:** A plot showing the **CV Error vs. $\text{Log}(\lambda)$** for the LASSO model, highlighting the optimal $\lambda_{min}$.

## LogReg 3/3: Evaluation & Coefficients  #MUSSA
* **Discuss:** The final selected LogReg model (likely the best regularized version) was evaluated on the held-out **Test Set**. Metrics focused on **AUC** and **Recall**.
* **Discuss:** Analyze the **Coefficients** of the best interpretable model, highlighting variables with large magnitudes (e.g., `duration`, `poutcome`) for actionable insight.
* **Image/Plot Idea:** The **Confusion Matrix** for the best LogReg model, showing the trade-off between True Positives and False Negatives.

---
# 2. K-Nearest Neighbors (KNN) Model
---

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*.

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.

## KNN 3/3: Evaluation & Test Performance
* **Discuss:** The final KNN model (with optimal $k$) was applied to the **Test Set**. Evaluate performance using **AUC** and compare it to the LogReg models.
* **Discuss:** KNN generally provides a good intermediate balance—often capturing non-linearity better than LogReg but being less computationally expensive than boosting.
* **Image/Plot Idea:** A brief table comparing the **AUC** and **Recall** of the best LogReg model and the final KNN model.

---
# 3. Boosting Model (High Predictive Benchmark)
---

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner).

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.

---

## Results: Comparative Evaluation (Slide 13)
* **Discuss:** Present the **Final Summary Table** comparing the Test Set performance (AUC, Recall, Accuracy) of all models (**Best LogReg**, **KNN**, **Boosting**).
* **Discuss:** Visually compare the models using the **Combined ROC Curve Plot** and discuss which model dominates the space.
* **Image/Plot Idea:** The **Combined ROC Curve Plot** comparing all three final models (LogReg, KNN, and Boosting).

## Discussion: The Trade-Off (Slide 14)
* **Discuss:** Reiterate the central trade-off: **Boosting** is the best *predictor*, but the **Regularized Logistic Regression** offers the most *interpretable* and *actionable* insights for the bank.
* **Discuss:** Final actionable advice for the bank: use **LogReg insights** to design campaigns and **Boosting** for a final scoring model if maximum predictive accuracy is needed.
* **Image/Plot Idea:** A visual metaphor (e.g., a scale or seesaw) contrasting "High Performance" vs. "High Interpretability."

---

## Discussion: Business Impact (Slide 13)
* **Discuss:** How the **chosen model** (Balancing performance and interpretability) can lead to **cost savings** and **higher ROI** for future campaigns.
* **Discuss:** Briefly touch upon **regulatory/ethical implications** in banking and the need to justify predictions (further supporting the value of interpretability).
* **Image/Plot Idea:** A currency symbol or a bar chart showing the conceptual increase in ROI due to better targeting.

## Discussion: Limitations & Future Work (Slide 14)
* **Discuss:** Acknowledge limitations, such as not using **time-series analysis** or fully accounting for the dependency of the 'duration' feature.
* **Discuss:** Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying **Deep Learning** models.
* **Image/Plot Idea:** A bulleted list of 2-3 **future research questions** (e.g., Time-Series, SMOTE, Deep Learning).

## References (Slide 15)
* **Discuss:** Simply list the required references for the data source and any major methodology papers.
* **Discuss:** Provide contact information for questions.
* Thank you
* **Image/Plot Idea:** The official UCI ML Repository logo and your contact email.

### 2025-12-07 Slides.tex
slides update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  ignorenonframetext,
]{beamer}
\newif\ifbibliography
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% remove section numbering
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{section title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{subsection title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Statistical Machine Learning Models on A Bank Marketing dataset},
  pdfauthor={Sandra, Tigist, Mussa, Waddah},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Statistical Machine Learning Models on A Bank Marketing dataset}
\subtitle{Colorado State University, DSCI 445: Statistical Machine
Learning, Dr.Andee Kaplan}
\author{Sandra, Tigist, Mussa, Waddah}

\begin{document}
\frame{\titlepage}

\begin{frame}{Project Motivation \& Business Problem}
\protect\phantomsection\label{project-motivation-business-problem}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} The bank's need for \textbf{efficient marketing} and
  the high cost of telemarketing campaigns. The goal is to maximize the
  success rate by predicting customer subscription.
\item
  \textbf{Discuss:} The challenge of the \textbf{rare positive class}
  (imbalanced data) and why prediction accuracy is crucial for
  \textbf{resource allocation}.
\item
  \textbf{Image/Plot Idea:} A simple, high-level infographic of a
  marketing campaign funnel, highlighting the bottleneck/drop-off point.
\end{itemize}

{[}Image of a marketing funnel illustrating customer conversion{]}
\end{frame}

\begin{frame}{The Bank Marketing Dataset (Slide 2)}
\protect\phantomsection\label{the-bank-marketing-dataset-slide-2}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} The dataset structure: \(\approx 45,000\) client
  observations from a Portuguese banking institution's telemarketing
  campaigns.
\item
  \textbf{Discuss:} Brief overview of the \textbf{four main predictor
  categories}: Demographics, Financial Indicators, Campaign Features,
  and Past Outcomes.
\item
  \textbf{Image/Plot Idea:} A small table summarizing the number of
  observations and the list of the 4 key variable groups (e.g.,
  Demographics, Financial, Campaign, Past).
\end{itemize}
\end{frame}

\begin{frame}{Target Variable \& Class Imbalance (Slide 3)}
\protect\phantomsection\label{target-variable-class-imbalance-slide-3}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} The \textbf{Target Variable (\(y\))} is binary: Did
  the client subscribe to a term deposit (Yes/No)?
\item
  \textbf{Discuss:} Emphasize the severity of the \textbf{class
  imbalance} (e.g., typically \(\approx 11\%\) `Yes' subscriptions).
  This means models must be evaluated beyond simple accuracy.
\item
  \textbf{Image/Plot Idea:} A \textbf{Bar Plot} showing the distribution
  of the target variable (\(y\)), visually highlighting the imbalance.
\end{itemize}
\end{frame}

\begin{frame}{Problem Statement \& Key Assumptions (Slide 4)}
\protect\phantomsection\label{problem-statement-key-assumptions-slide-4}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} The project goal is \textbf{Binary Classification}:
  Building statistical models that are both \textbf{predictive} and
  \textbf{interpretable} for management.
\item
  \textbf{Discuss:} Two key assumptions: Data observations are
  \textbf{independent} (no time-series methods), and we focus on
  \textbf{course-covered models} for direct comparison.
\item
  \textbf{Image/Plot Idea:} An icon representing \textbf{Predictive
  Power} next to an icon representing \textbf{Interpretability} (The two
  core goals).
\end{itemize}
\end{frame}

\begin{frame}{Methodology: Data Preparation (Slide 5)}
\protect\phantomsection\label{methodology-data-preparation-slide-5}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} \textbf{Data Wrangling} involved converting
  categorical variables to factors and strategies for handling `unknown'
  levels (treating them as a separate category vs.~missing).
\item
  \textbf{Discuss:} Established a \textbf{train/test split} (e.g.,
  70/30) to ensure models are evaluated on unseen data for a fair
  comparison of out-of-sample error.
\item
  \textbf{Image/Plot Idea:} A simple flow chart showing: Raw Data
  \(\rightarrow\) Wrangling/Cleaning \(\rightarrow\) Train/Test Split.
\end{itemize}

{[}Image of a data science workflow diagram{]}
\end{frame}

\begin{frame}{Methodology: Models Selected (Slide 6)}
\protect\phantomsection\label{methodology-models-selected-slide-6}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} \textbf{Logistic Regression:} Chosen for its
  \textbf{linearity} and \textbf{high interpretability} (provides clear
  marketing insights via coefficients).
\item
  \textbf{Discuss:} \textbf{K-Nearest Neighbors (KNN):} A
  \textbf{non-parametric} and \textbf{flexible} model to capture
  non-linearities, though sensitive to scale.
\item
  \textbf{Image/Plot Idea:} A side-by-side visual comparing the
  conceptual difference between a linear boundary (LogReg) and a local,
  non-linear one (KNN).
\end{itemize}
\end{frame}

\begin{frame}{Methodology: Ensemble/Advanced Models (Slide 7)}
\protect\phantomsection\label{methodology-ensembleadvanced-models-slide-7}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} \textbf{Boosting (e.g., AdaBoost/Gradient
  Boosting):} An ensemble method expected to provide the \textbf{highest
  predictive accuracy} by sequentially correcting errors of weak
  learners.
\item
  \textbf{Discuss:} \textbf{Model Training Strategy:} All models used
  \textbf{k-fold Cross-Validation} (CV) on the training set to prevent
  overfitting and select optimal hyperparameters.
\item
  \textbf{Image/Plot Idea:} A conceptual diagram illustrating the
  boosting process (e.g., weak learners combining to form a strong
  learner).
\end{itemize}
\end{frame}

\begin{frame}{Model Tuning \& Validation Details (Slide 8)}
\protect\phantomsection\label{model-tuning-validation-details-slide-8}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Mentioned specific tuning steps:
  \textbf{Hyperparameter tuning} (e.g., finding optimal \(k\) for KNN)
  and \textbf{Feature Selection} (e.g., stepwise for LogReg) to simplify
  models.
\item
  \textbf{Discuss:} For models like LogReg, we also explored
  \textbf{Regularization} (Ridge/Lasso) to prevent overfitting and
  improve generalization.
\item
  \textbf{Image/Plot Idea:} A plot showing CV error vs.~\(k\) (for KNN)
  or \(\lambda\) (for Lasso/Ridge) to demonstrate hyperparameter
  selection.
\end{itemize}
\end{frame}

\begin{frame}{Results: Model Evaluation Metrics (Slide 9)}
\protect\phantomsection\label{results-model-evaluation-metrics-slide-9}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Primary metrics are \textbf{Test Error Rate} and
  \textbf{AUC (Area Under the ROC Curve)}, which are robust to class
  imbalance.
\item
  \textbf{Discuss:} Also emphasize \textbf{Recall} (sensitivity) because
  identifying the few `Yes' subscribers is more critical than high
  accuracy on the majority `No' class.
\item
  \textbf{Image/Plot Idea:} A visual diagram of a \textbf{Confusion
  Matrix}, highlighting the meaning of True Positives (TP) and False
  Negatives (FN).
\end{itemize}
\end{frame}

\begin{frame}{Results: Performance Comparison (Slide 10)}
\protect\phantomsection\label{results-performance-comparison-slide-10}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Present the \textbf{final performance comparison},
  identifying the model with the lowest Test Error/Highest AUC (e.g.,
  \textbf{Boosting}).
\item
  \textbf{Discuss:} Acknowledge the trade-off: The \textbf{best
  predictive model} may not be the \textbf{most interpretable}.
\item
  \textbf{Image/Plot Idea:} A \textbf{Table} or \textbf{Bar Chart}
  summarizing AUC, Accuracy, and Recall for \textbf{Logistic
  Regression}, \textbf{KNN}, and \textbf{Boosting}.
\end{itemize}
\end{frame}

\begin{frame}{Results: ROC Curve Analysis (Slide 11)}
\protect\phantomsection\label{results-roc-curve-analysis-slide-11}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} A visual comparison of models using the
  \textbf{Receiver Operating Characteristic (ROC) Curve}. A curve closer
  to the top-left is better.
\item
  \textbf{Discuss:} This curve provides insight into how each model
  balances the \textbf{True Positive Rate} (Recall) vs.~\textbf{False
  Positive Rate}.
\item
  \textbf{Image/Plot Idea:} The final \textbf{ROC Curve plot} comparing
  the performance of all tested models.
\end{itemize}
\end{frame}

\begin{frame}{Discussion: Interpreting the Best Model (Slide 12)}
\protect\phantomsection\label{discussion-interpreting-the-best-model-slide-12}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Focus on the \textbf{most interpretable model
  (Logistic Regression)}---what were the key coefficients/insights?
  (e.g., `Duration of last contact' is a strong predictor).
\item
  \textbf{Discuss:} Discuss how these feature importances can be
  directly translated into \textbf{actionable marketing strategies} for
  the bank.
\item
  \textbf{Image/Plot Idea:} A \textbf{Feature Importance Plot} (for
  Boosting) or a \textbf{Coefficient Plot} (for LogReg), highlighting
  the top 3-5 most influential variables.
\end{itemize}
\end{frame}

\begin{frame}{Discussion: Business Impact (Slide 13)}
\protect\phantomsection\label{discussion-business-impact-slide-13}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} How the \textbf{chosen model} (Balancing performance
  and interpretability) can lead to \textbf{cost savings} and
  \textbf{higher ROI} for future campaigns.
\item
  \textbf{Discuss:} Briefly touch upon \textbf{regulatory/ethical
  implications} in banking and the need to justify predictions (further
  supporting the value of interpretability).
\item
  \textbf{Image/Plot Idea:} A currency symbol or a bar chart showing the
  conceptual increase in ROI due to better targeting.
\end{itemize}
\end{frame}

\begin{frame}{Discussion: Limitations \& Future Work (Slide 14)}
\protect\phantomsection\label{discussion-limitations-future-work-slide-14}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Acknowledge limitations, such as not using
  \textbf{time-series analysis} or fully accounting for the dependency
  of the `duration' feature.
\item
  \textbf{Discuss:} Suggest future improvements: Exploring advanced
  techniques like \textbf{resampling methods} (SMOTE) to handle
  imbalance better or trying \textbf{Deep Learning} models.
\item
  \textbf{Image/Plot Idea:} A bulleted list of 2-3 \textbf{future
  research questions} (e.g., Time-Series, SMOTE, Deep Learning).
\end{itemize}
\end{frame}

\begin{frame}{References (Slide 15)}
\protect\phantomsection\label{references-slide-15}
\begin{itemize}
\tightlist
\item
  \textbf{Discuss:} Simply list the required references for the data
  source and any major methodology papers.
\item
  \textbf{Discuss:} Provide contact information for questions.
\item
  \textbf{Image/Plot Idea:} The official UCI ML Repository logo and your
  contact email.
\end{itemize}
\end{frame}

\end{document}

### 2025-12-07 Slides.Rmd
slides update2

@@
 
-## Discussion: Business Impact (Slide 13)
+## Discussion: Business Impact (Slide 13)  #MUSSA
 * **Discuss:** How the **chosen model** (Balancing performance and interpretability) can lead to **cost savings** and **higher ROI** for future campaigns.
@@
 
-## Discussion: Limitations & Future Work (Slide 14)
+## Discussion: Limitations & Future Work (Slide 14)  #MUSSA
 * **Discuss:** Acknowledge limitations, such as not using **time-series analysis** or fully accounting for the dependency of the 'duration' feature.

### 2025-12-07 model_fit.Rmd
slides update2

@@
 
-view(coef_df %>%  filter(`Pr(>|z|)`< 0.01))
+sample(coef_df %>%  filter(`Pr(>|z|)`< 0.01))
 ```
@@
 - Optional: **regularized logistic regression** (Ridge/Lasso) using CV-selected $\lambda$.
-
-*Standard raw logistic reg (without tuning)*
-
-```{r}
-glm_out <- glm(y ~ ., data = bank_full, family = binomial)
-summary(glm_out)
-```
-
-Discuss accuracy (cv test error) and amount of parameters
-
-```{r}
-glm.full.cv <- glm(initial_formula, data = data, family = binomial, x = TRUE)
-
-cv.error.full <- cv.glm(data, glm.full.cv, K = 5)$delta[2]
-length(coef(glm.full.cv))
-```
-
-This model is complex due to the large number of dummy variables.
-The CV test error provides an estimate of how well this complexity generalizes to new, out of sample data.
-If the CV error is significantly higher than the training error, the model is likely overfitting
 

### 2025-12-07 Slides.Rmd
slides update3

@@
 title: "Statistical Machine Learning Models on A Bank Marketing dataset"
-subtitle: "Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee Kaplan"
 author: "Sandra, Tigist, Mussa Hassen, Waddah"
-output: ioslides_presentation
+subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
+  Kaplan'
+output:
+  ioslides_presentation: 
+    css: styles.css
+
 ---
+
+```{r appendix, include=FALSE}
+library(readr)
+library(knitr)
+library(tidyverse)
+library(boot)
+library(glmnet)
+library(caret)
+library(pROC)
+library(MASS)
+library(ggplot2)
+
+set.seed(445)
+
+bank_full <- read_delim("data/bank-full_1.csv", 
+    delim = ";", escape_double = FALSE, trim_ws = TRUE)
+
+bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))
+
+bank_full <- bank_full %>%
+  mutate(
+    job = as.factor(job),
+    marital = as.factor(marital),
+    education = as.factor(education),
+    default = as.factor(default),
+    housing = as.factor(housing),
+    loan = as.factor(loan),
+    contact = as.factor(contact),
+    month = as.factor(month),
+    poutcome = as.factor(poutcome)) %>% 
+  mutate(previously_contacted = as.factor(ifelse(pdays == -1, "no", "yes")))
+
+train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
+train_data <- bank_full[train_index, ]
+test_data  <- bank_full[-train_index, ]
+
+cat(paste("Training set size:", nrow(train_data), "\n"))
+cat(paste("Test set size:", nrow(test_data), "\n"))
+
+full_formula <- y ~ age + job + marital + education + default + balance + 
+  housing + loan + contact + day + month + duration + campaign + 
+  pdays + previous + poutcome + previously_contacted
+
+glm.full <- glm(full_formula, data = train_data, family = "binomial")
+
+# Perform backward selection
+glm.step <- step(glm.full, direction = "backward", trace = 0)
+
+selected_formula <- formula(glm.step)
+selected_formula
+
+glm.baseline <- glm(selected_formula, data = train_data, family = "binomial")
+summary(glm.baseline)
+
+glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
+    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
+summary(glm.tran)
+
+residuals <- residuals(glm.tran, type = "deviance")
+train_data$residuals <- residuals
+
+cost <- function(r, pi) {
+  mean(r != (pi > 0.5))
+}
+
+cv.error.baseline <- cv.glm(
+  data = train_data, 
+  glmfit = glm.tran, 
+  cost = cost, 
+  K = 5
+)
+
+cv_error_rate <- cv.error.baseline$delta[2]
+round(cv_error_rate, 5)
+
+X_full <- model.matrix(y ~ . - 1, data = bank_full)
+Y_full <- as.numeric(bank_full$y) - 1
+
+X_train <- X_full[train_index, ]
+Y_train <- Y_full[train_index]
+
+X_test <- X_full[-train_index, ]
+Y_test <- Y_full[-train_index]
+
+cv.ridge <- cv.glmnet(X_train, Y_train, 
+                      alpha = 0, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.ridge <- cv.ridge$lambda.min
+
+cv.lasso <- cv.glmnet(X_train, Y_train, 
+                      alpha = 1, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.lasso <- cv.lasso$lambda.min
+
+baseline_prob <- predict(glm.tran, newdata = test_data, type = "response")
+baseline_pred <- factor(ifelse(baseline_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+ridge_prob <- predict(cv.ridge, s = lambda.min.ridge, newx = X_test, type = "response")
+ridge_pred <- factor(ifelse(ridge_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+lasso_prob <- predict(cv.lasso, s = lambda.min.lasso, newx = X_test, type = "response")
+lasso_pred <- factor(ifelse(lasso_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+cm_baseline <- confusionMatrix(baseline_pred, test_data$y, positive = "yes")
+cm_ridge <- confusionMatrix(ridge_pred, test_data$y, positive = "yes")
+cm_lasso <- confusionMatrix(lasso_pred, test_data$y, positive = "yes")
+
+metrics <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  Accuracy = c(cm_baseline$overall['Accuracy'], cm_ridge$overall['Accuracy'], cm_lasso$overall['Accuracy']),
+  Sensitivity = c(cm_baseline$byClass['Sensitivity'], cm_ridge$byClass['Sensitivity'], cm_lasso$byClass['Sensitivity']),
+  Specificity = c(cm_baseline$byClass['Specificity'], cm_ridge$byClass['Specificity'], cm_lasso$byClass['Specificity'])
+)
+
+roc_baseline <- roc(test_data$y, baseline_prob)
+roc_ridge <- roc(test_data$y, as.numeric(ridge_prob))
+roc_lasso <- roc(test_data$y, as.numeric(lasso_prob))
+
+roc_data <- data.frame(
+  fpr = c(1 - roc_baseline$specificities, 1 - roc_ridge$specificities, 1 - roc_lasso$specificities),
+  tpr = c(roc_baseline$sensitivities, roc_ridge$sensitivities, roc_lasso$sensitivities),
+  Model = factor(c(rep("Baseline Logit", length(roc_baseline$specificities)),
+                   rep("Ridge Logit", length(roc_ridge$specificities)),
+                   rep("LASSO Logit", length(roc_lasso$specificities))),
+                 levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit"))
+)
+
+auc_values <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  AUC = c(auc(roc_baseline), auc(roc_ridge), auc(roc_lasso))
+)
+
+```
 
@@
 
-## LogReg 1/3: Feature Selection & Linearity  #MUSSA
-* **Discuss:** Initial feature selection was performed using **Backward Stepwise Elimination** (based on **AIC**) on the full model to find the most parsimonious set of main effects.
-* **Discuss:** **Deviance Residual Plots** were used to assess the linearity assumption for numeric predictors (e.g., `age`, `balance`). This led to testing non-linear terms.
-* **Image/Plot Idea:** The **Residual Plot** for a key numeric feature (e.g., `balance`), showing a curve suggesting non-linearity.
+## LogReg: Feature Selection & Linearity  #MUSSA
+* Initial feature selection was performed using **Backward Stepwise Elimination** on the full model to find the most useful set of main effects.
+* Our selected formula resutled:
 
-## LogReg 2/3: Regularization & Tuning   #MUSSA
-* **Discuss:** We explored **Ridge ($\alpha=0$)** and **LASSO ($\alpha=1$) Regularization** to manage model complexity and feature selection, particularly with the large number of dummy variables.
-* **Discuss:** **5-Fold Cross-Validation** was used to select the optimal penalty parameter, $\lambda$, which minimized the mean CV error for each regularized model.
-* **Image/Plot Idea:** A plot showing the **CV Error vs. $\text{Log}(\lambda)$** for the LASSO model, highlighting the optimal $\lambda_{min}$.
+$y \sim job + marital + education + balance + housing + $
 
-## LogReg 3/3: Evaluation & Coefficients  #MUSSA
-* **Discuss:** The final selected LogReg model (likely the best regularized version) was evaluated on the held-out **Test Set**. Metrics focused on **AUC** and **Recall**.
-* **Discuss:** Analyze the **Coefficients** of the best interpretable model, highlighting variables with large magnitudes (e.g., `duration`, `poutcome`) for actionable insight.
-* **Image/Plot Idea:** The **Confusion Matrix** for the best LogReg model, showing the trade-off between True Positives and False Negatives.
+$loan + contact + day + month + duration + $
+
+$campaign + previous+ poutcome$
+* **Residual Plots** were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.
+
+---
+
+```{r residuals_slide, echo=FALSE, results='asis'}
+residuals <- residuals(glm.baseline, type = "deviance")
+train_data$residuals <- residuals
+
+ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+```
+
+## LogReg: Models with non-linear components and interactions
+* A **quadratic term** $I(\text{balance}^4)$—was added to the formula to reduce the model's bias.
+* We then tested for crucial **interaction terms** (between campaign features and demographics). 
+* Our model had a cross-validated test error rate 0.09846
+
+## LogReg: Regularization & Tuning   #MUSSA
+* We explored Ridge and LASSO regularization as an alternative approach for feature selection.
+* **5-Fold Cross-Validation** was used to select the optimal penalty parameter, $\lambda$, which minimized the mean CV error for each regularized model.
+
+```{r, echo=FALSE, results='asis', fig.width=4, fig.height=3}
+
+plot(cv.ridge) 
+plot(cv.lasso)
+```
+
+## LogReg: Evaluation & Coefficients  #MUSSA
+* The final selected LogReg model was evaluated on the 5-fold CV error rate, along with metrics focused on the confusion matrix.
+* Models performed relatively similar. We would prefer a model with higher **sensitivity** (few false negatives)
+
+```{r, echo=FALSE, results='asis'}
+metrics$AUC <- auc_values[, ncol(auc_values)]
+metrics %>% kable()
+```
 
@@
 
----
-
-## Results: Comparative Evaluation (Slide 13)
+## Results: Comparative Evaluation (Slide 13). #MUSSA
 * **Discuss:** Present the **Final Summary Table** comparing the Test Set performance (AUC, Recall, Accuracy) of all models (**Best LogReg**, **KNN**, **Boosting**).
@@
 
-## Discussion: The Trade-Off (Slide 14)  
+## The Trade-Off / Business Impact  #MUSSA
 * **Discuss:** Reiterate the central trade-off: **Boosting** is the best *predictor*, but the **Regularized Logistic Regression** offers the most *interpretable* and *actionable* insights for the bank.
@@
 * **Image/Plot Idea:** A visual metaphor (e.g., a scale or seesaw) contrasting "High Performance" vs. "High Interpretability."
-
----
-
-## Discussion: Business Impact (Slide 13)  #MUSSA
 * **Discuss:** How the **chosen model** (Balancing performance and interpretability) can lead to **cost savings** and **higher ROI** for future campaigns.
@@
 * **Discuss:** Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying **Deep Learning** models.
-* **Image/Plot Idea:** A bulleted list of 2-3 **future research questions** (e.g., Time-Series, SMOTE, Deep Learning).
 

### 2025-12-07 Slides.tex
slides update3

@@
 \makeatother
+\usepackage{graphicx}
+\makeatletter
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
+% Set default figure placement to htbp
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   pdftitle={Statistical Machine Learning Models on A Bank Marketing dataset},
-  pdfauthor={Sandra, Tigist, Mussa, Waddah},
+  pdfauthor={Sandra, Tigist, Mussa Hassen, Waddah},
   hidelinks,
@@
 Learning, Dr.Andee Kaplan}
-\author{Sandra, Tigist, Mussa, Waddah}
+\author{Sandra, Tigist, Mussa Hassen, Waddah}
 
@@
 \item
-  \textbf{Discuss:} The bank's need for \textbf{efficient marketing} and
-  the high cost of telemarketing campaigns. The goal is to maximize the
-  success rate by predicting customer subscription.
+  \textbf{Discuss:} The bank's need for efficient marketing and the high
+  cost of telemarketing campaigns. The goal is to maximize the success
+  rate by predicting customer subscription.
 \item
-  \textbf{Discuss:} The challenge of the \textbf{rare positive class}
-  (imbalanced data) and why prediction accuracy is crucial for
-  \textbf{resource allocation}.
+  \textbf{Discuss:} The challenge of less positive class (imbalanced
+  data) and why prediction accuracy is crucial.
 \item
-  \textbf{Image/Plot Idea:} A simple, high-level infographic of a
+  \textbf{Image/Plot Idea:} A simple, high-level info graphic of a
   marketing campaign funnel, highlighting the bottleneck/drop-off point.
@@
 \item
-  \textbf{Discuss:} Brief overview of the \textbf{four main predictor
-  categories}: Demographics, Financial Indicators, Campaign Features,
-  and Past Outcomes.
+  \textbf{Discuss:} Brief overview of the four main predictor
+  categories: Client Demographics, Financial Indicators, Campaign
+  Features, and Past Outcomes. Mention the paper had more features, fro
+  privacy reasons the public data set has less
 \item
-  \textbf{Image/Plot Idea:} A small table summarizing the number of
-  observations and the list of the 4 key variable groups (e.g.,
-  Demographics, Financial, Campaign, Past).
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Target Variable \& Class Imbalance (Slide 3)}
-\protect\phantomsection\label{target-variable-class-imbalance-slide-3}
-\begin{itemize}
-\tightlist
+  \textbf{Discuss:} The Target Variable (\(y\)) is binary: Did the
+  client subscribe to a term deposit (Yes/No)?
 \item
-  \textbf{Discuss:} The \textbf{Target Variable (\(y\))} is binary: Did
-  the client subscribe to a term deposit (Yes/No)?
+  \textbf{Discuss:} Emphasize the severity of the class imbalance (e.g.,
+  typically \(\approx 11\%\) `Yes' subscriptions). This may yield higher
+  false negatives.
 \item
-  \textbf{Discuss:} Emphasize the severity of the \textbf{class
-  imbalance} (e.g., typically \(\approx 11\%\) `Yes' subscriptions).
-  This means models must be evaluated beyond simple accuracy.
+  \textbf{Image/Plot Idea:} A small table summarizing the list of the 4
+  key variable groups (e.g., Demographics, Financial, Campaign, Past).
 \item
@@
 
-\begin{frame}{Problem Statement \& Key Assumptions (Slide 4)}
-\protect\phantomsection\label{problem-statement-key-assumptions-slide-4}
+\begin{frame}{Problem Statement \& Methodology Overview (Slide 3)}
+\protect\phantomsection\label{problem-statement-methodology-overview-slide-3}
 \begin{itemize}
 \tightlist
+\item
+  \textbf{Discuss:} The project aims for a \textbf{Comparative
+  Study}---comparing models based on two criteria: \textbf{Predictive
+  Power} and \textbf{Interpretability}.
 \item
@@
 \item
-  \textbf{Discuss:} Two key assumptions: Data observations are
-  \textbf{independent} (no time-series methods), and we focus on
-  \textbf{course-covered models} for direct comparison.
+  \textbf{Discuss:} Established an \textbf{80/20 Train/Test Split} and
+  used \textbf{5-Fold Cross-Validation (CV)} on the training set for
+  tuning and estimating out-of-sample error.
 \item
@@
 \item
-  \textbf{Discuss:} Established a \textbf{train/test split} (e.g.,
-  70/30) to ensure models are evaluated on unseen data for a fair
-  comparison of out-of-sample error.
-\item
   \textbf{Image/Plot Idea:} A simple flow chart showing: Raw Data
   \(\rightarrow\) Wrangling/Cleaning \(\rightarrow\) Train/Test Split.
+\item
+  \textbf{Image/Plot Idea:} Summary stats or informational plots
 \end{itemize}
@@
 
-\begin{frame}{Methodology: Models Selected (Slide 6)}
-\protect\phantomsection\label{methodology-models-selected-slide-6}
+\begin{frame}[fragile]{LogReg: Feature Selection \& Linearity \#MUSSA}
+\protect\phantomsection\label{logreg-feature-selection-linearity-mussa}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} \textbf{Logistic Regression:} Chosen for its
-  \textbf{linearity} and \textbf{high interpretability} (provides clear
-  marketing insights via coefficients).
+  \textbf{Discuss:} Initial feature selection was performed using
+  \textbf{Backward Stepwise Elimination} (based on \textbf{AIC}) on the
+  full model to find the most parsimonious set of main effects.
 \item
-  \textbf{Discuss:} \textbf{K-Nearest Neighbors (KNN):} A
-  \textbf{non-parametric} and \textbf{flexible} model to capture
-  non-linearities, though sensitive to scale.
+  \textbf{Discuss:} \textbf{Deviance Residual Plots} were used to assess
+  the linearity assumption for numeric predictors (e.g., \texttt{age},
+  \texttt{balance}). This led to testing non-linear terms.
 \item
-  \textbf{Image/Plot Idea:} A side-by-side visual comparing the
-  conceptual difference between a linear boundary (LogReg) and a local,
-  non-linear one (KNN).
+  \textbf{Image/Plot Idea:} The \textbf{Residual Plot} for a key numeric
+  feature (e.g., \texttt{balance}), showing a curve suggesting
+  non-linearity.
+\end{itemize}
+
+\pandocbounded{\includegraphics[keepaspectratio]{Slides_files/figure-beamer/residuals_slide-1.pdf}}
+\end{frame}
+
+\begin{frame}{LogReg: Models with nonlinear components and interactions}
+\protect\phantomsection\label{logreg-models-with-nonlinear-components-and-interactions}
+\begin{itemize}
+\tightlist
+\item
 \end{itemize}
@@
 
-\begin{frame}{Methodology: Ensemble/Advanced Models (Slide 7)}
-\protect\phantomsection\label{methodology-ensembleadvanced-models-slide-7}
+\begin{frame}{LogReg: Regularization \& Tuning \#MUSSA}
+\protect\phantomsection\label{logreg-regularization-tuning-mussa}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} \textbf{Boosting (e.g., AdaBoost/Gradient
-  Boosting):} An ensemble method expected to provide the \textbf{highest
-  predictive accuracy} by sequentially correcting errors of weak
-  learners.
+  \textbf{Discuss:} We explored \textbf{Ridge (\(\alpha=0\))} and
+  \textbf{LASSO (\(\alpha=1\)) Regularization} to manage model
+  complexity and feature selection, particularly with the large number
+  of dummy variables.
 \item
-  \textbf{Discuss:} \textbf{Model Training Strategy:} All models used
-  \textbf{k-fold Cross-Validation} (CV) on the training set to prevent
-  overfitting and select optimal hyperparameters.
+  \textbf{Discuss:} \textbf{5-Fold Cross-Validation} was used to select
+  the optimal penalty parameter, \(\lambda\), which minimized the mean
+  CV error for each regularized model.
 \item
-  \textbf{Image/Plot Idea:} A conceptual diagram illustrating the
-  boosting process (e.g., weak learners combining to form a strong
-  learner).
+  \textbf{Image/Plot Idea:} A plot showing the \textbf{CV Error
+  vs.~\(\text{Log}(\lambda)\)} for the LASSO model, highlighting the
+  optimal \(\lambda_{min}\).
 \end{itemize}
@@
 
-\begin{frame}{Model Tuning \& Validation Details (Slide 8)}
-\protect\phantomsection\label{model-tuning-validation-details-slide-8}
+\begin{frame}[fragile]{LogReg: Evaluation \& Coefficients \#MUSSA}
+\protect\phantomsection\label{logreg-evaluation-coefficients-mussa}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Mentioned specific tuning steps:
-  \textbf{Hyperparameter tuning} (e.g., finding optimal \(k\) for KNN)
-  and \textbf{Feature Selection} (e.g., stepwise for LogReg) to simplify
-  models.
+  \textbf{Discuss:} The final selected LogReg model (likely the best
+  regularized version) was evaluated on the held-out \textbf{Test Set}.
+  Metrics focused on \textbf{AUC} and \textbf{Recall}.
 \item
-  \textbf{Discuss:} For models like LogReg, we also explored
-  \textbf{Regularization} (Ridge/Lasso) to prevent overfitting and
-  improve generalization.
+  \textbf{Discuss:} Analyze the \textbf{Coefficients} of the best
+  interpretable model, highlighting variables with large magnitudes
+  (e.g., \texttt{duration}, \texttt{poutcome}) for actionable insight.
 \item
-  \textbf{Image/Plot Idea:} A plot showing CV error vs.~\(k\) (for KNN)
-  or \(\lambda\) (for Lasso/Ridge) to demonstrate hyperparameter
-  selection.
+  \textbf{Image/Plot Idea:} The \textbf{Confusion Matrix} for the best
+  LogReg model, showing the trade-off between True Positives and False
+  Negatives.
 \end{itemize}
@@
 
-\begin{frame}{Results: Model Evaluation Metrics (Slide 9)}
-\protect\phantomsection\label{results-model-evaluation-metrics-slide-9}
+\begin{frame}{KNN 1/3: Feature Preparation}
+\protect\phantomsection\label{knn-13-feature-preparation}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Primary metrics are \textbf{Test Error Rate} and
-  \textbf{AUC (Area Under the ROC Curve)}, which are robust to class
-  imbalance.
+  \textbf{Discuss:} \textbf{KNN} requires scaling: all numeric
+  predictors were \textbf{standardized} (mean=0, SD=1) to prevent
+  features with larger scales from dominating the distance calculation.
 \item
-  \textbf{Discuss:} Also emphasize \textbf{Recall} (sensitivity) because
-  identifying the few `Yes' subscribers is more critical than high
-  accuracy on the majority `No' class.
+  \textbf{Discuss:} The non-parametric nature means \textbf{no explicit
+  feature selection} via coefficients, but the scaling step is vital for
+  ensuring equal contribution from all dimensions.
 \item
-  \textbf{Image/Plot Idea:} A visual diagram of a \textbf{Confusion
-  Matrix}, highlighting the meaning of True Positives (TP) and False
-  Negatives (FN).
+  \textbf{Image/Plot Idea:} A visual comparison: a scatter plot of two
+  features \emph{unscaled} vs.~the same features
+  \emph{scaled/standardized}.
 \end{itemize}
@@
 
-\begin{frame}{Results: Performance Comparison (Slide 10)}
-\protect\phantomsection\label{results-performance-comparison-slide-10}
+\begin{frame}{KNN 2/3: Hyperparameter Tuning}
+\protect\phantomsection\label{knn-23-hyperparameter-tuning}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Present the \textbf{final performance comparison},
-  identifying the model with the lowest Test Error/Highest AUC (e.g.,
-  \textbf{Boosting}).
+  \textbf{Discuss:} The sole hyperparameter, \textbf{\(k\)} (the number
+  of neighbors), determines the model's complexity (i.e., the
+  bias-variance trade-off).
 \item
-  \textbf{Discuss:} Acknowledge the trade-off: The \textbf{best
-  predictive model} may not be the \textbf{most interpretable}.
+  \textbf{Discuss:} \textbf{5-Fold Cross-Validation} was performed
+  across a range of \(k\) values to find the optimal \(k\) that
+  minimized the mean CV error.
 \item
-  \textbf{Image/Plot Idea:} A \textbf{Table} or \textbf{Bar Chart}
-  summarizing AUC, Accuracy, and Recall for \textbf{Logistic
-  Regression}, \textbf{KNN}, and \textbf{Boosting}.
+  \textbf{Image/Plot Idea:} A plot showing \textbf{CV Test Error
+  vs.~\(k\)}, illustrating the U-shape curve where the optimal \(k\) is
+  selected.
 \end{itemize}
@@
 
-\begin{frame}{Results: ROC Curve Analysis (Slide 11)}
-\protect\phantomsection\label{results-roc-curve-analysis-slide-11}
+\begin{frame}{KNN 3/3: Evaluation \& Test Performance}
+\protect\phantomsection\label{knn-33-evaluation-test-performance}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} A visual comparison of models using the
-  \textbf{Receiver Operating Characteristic (ROC) Curve}. A curve closer
-  to the top-left is better.
+  \textbf{Discuss:} The final KNN model (with optimal \(k\)) was applied
+  to the \textbf{Test Set}. Evaluate performance using \textbf{AUC} and
+  compare it to the LogReg models.
 \item
-  \textbf{Discuss:} This curve provides insight into how each model
-  balances the \textbf{True Positive Rate} (Recall) vs.~\textbf{False
-  Positive Rate}.
+  \textbf{Discuss:} KNN generally provides a good intermediate
+  balance---often capturing non-linearity better than LogReg but being
+  less computationally expensive than boosting.
 \item
-  \textbf{Image/Plot Idea:} The final \textbf{ROC Curve plot} comparing
-  the performance of all tested models.
+  \textbf{Image/Plot Idea:} A brief table comparing the \textbf{AUC} and
+  \textbf{Recall} of the best LogReg model and the final KNN model.
 \end{itemize}
@@
 
-\begin{frame}{Discussion: Interpreting the Best Model (Slide 12)}
-\protect\phantomsection\label{discussion-interpreting-the-best-model-slide-12}
+\begin{frame}{Boosting 1/3: Model Formulation}
+\protect\phantomsection\label{boosting-13-model-formulation}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Focus on the \textbf{most interpretable model
-  (Logistic Regression)}---what were the key coefficients/insights?
-  (e.g., `Duration of last contact' is a strong predictor).
+  \textbf{Discuss:} \textbf{Boosting} (e.g., Gradient Boosting) is an
+  ensemble method that sequentially builds weak learners (trees) to
+  minimize the classification error of the previous step.
 \item
-  \textbf{Discuss:} Discuss how these feature importances can be
-  directly translated into \textbf{actionable marketing strategies} for
-  the bank.
+  \textbf{Discuss:} This model is expected to be our
+  \textbf{highest-performing benchmark} but sacrifices interpretability
+  due to its complex, additive structure.
 \item
-  \textbf{Image/Plot Idea:} A \textbf{Feature Importance Plot} (for
-  Boosting) or a \textbf{Coefficient Plot} (for LogReg), highlighting
-  the top 3-5 most influential variables.
+  \textbf{Image/Plot Idea:} A conceptual diagram illustrating the
+  boosting process (weak learners combining sequentially to form a
+  strong learner).
 \end{itemize}
@@
 
-\begin{frame}{Discussion: Business Impact (Slide 13)}
-\protect\phantomsection\label{discussion-business-impact-slide-13}
+\begin{frame}{Boosting 2/3: Hyperparameter Tuning}
+\protect\phantomsection\label{boosting-23-hyperparameter-tuning}
 \begin{itemize}
 \tightlist
+\item
+  \textbf{Discuss:} Multiple hyperparameters were tuned using
+  \textbf{Cross-Validation}, including the \textbf{number of trees} (M),
+  \textbf{interaction depth} (complexity of weak learners), and
+  \textbf{learning rate} (\(\eta\)).
+\item
+  \textbf{Discuss:} The tuning process involves finding the right
+  balance between these parameters to ensure the model generalizes well
+  without overfitting the training data.
+\item
+  \textbf{Image/Plot Idea:} A plot showing the \textbf{Error Rate
+  vs.~Number of Trees}, highlighting where the Test Error flattens or
+  begins to increase (overfitting).
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Boosting 3/3: Evaluation \& Importance}
+\protect\phantomsection\label{boosting-33-evaluation-importance}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Discuss:} The final Boosting model was evaluated on the
+  \textbf{Test Set}. Confirm that it achieved the \textbf{highest AUC},
+  setting the benchmark for predictive accuracy.
+\item
+  \textbf{Discuss:} While coefficients are not available, discuss the
+  model's \textbf{Variable Importance Plot} as the only glimpse into its
+  internal workings.
+\item
+  \textbf{Image/Plot Idea:} The \textbf{Variable Importance Plot} from
+  the Boosting model, which should confirm that the most important
+  features align reasonably well with LogReg's significant coefficients.
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Results: Comparative Evaluation (Slide 13). \#MUSSA}
+\protect\phantomsection\label{results-comparative-evaluation-slide-13.-mussa}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Discuss:} Present the \textbf{Final Summary Table} comparing
+  the Test Set performance (AUC, Recall, Accuracy) of all models
+  (\textbf{Best LogReg}, \textbf{KNN}, \textbf{Boosting}).
+\item
+  \textbf{Discuss:} Visually compare the models using the
+  \textbf{Combined ROC Curve Plot} and discuss which model dominates the
+  space.
+\item
+  \textbf{Image/Plot Idea:} The \textbf{Combined ROC Curve Plot}
+  comparing all three final models (LogReg, KNN, and Boosting).
+\end{itemize}
+\end{frame}
+
+\begin{frame}{The Trade-Off / Business Impact \#MUSSA}
+\protect\phantomsection\label{the-trade-off-business-impact-mussa}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Discuss:} Reiterate the central trade-off: \textbf{Boosting}
+  is the best \emph{predictor}, but the \textbf{Regularized Logistic
+  Regression} offers the most \emph{interpretable} and \emph{actionable}
+  insights for the bank.
+\item
+  \textbf{Discuss:} Final actionable advice for the bank: use
+  \textbf{LogReg insights} to design campaigns and \textbf{Boosting} for
+  a final scoring model if maximum predictive accuracy is needed.
+\item
+  \textbf{Image/Plot Idea:} A visual metaphor (e.g., a scale or seesaw)
+  contrasting ``High Performance'' vs.~``High Interpretability.''
 \item
@@
 
-\begin{frame}{Discussion: Limitations \& Future Work (Slide 14)}
-\protect\phantomsection\label{discussion-limitations-future-work-slide-14}
+\begin{frame}{Discussion: Limitations \& Future Work (Slide 14) \#MUSSA}
+\protect\phantomsection\label{discussion-limitations-future-work-slide-14-mussa}
 \begin{itemize}
@@
   imbalance better or trying \textbf{Deep Learning} models.
-\item
-  \textbf{Image/Plot Idea:} A bulleted list of 2-3 \textbf{future
-  research questions} (e.g., Time-Series, SMOTE, Deep Learning).
 \end{itemize}
@@
 \item
+  Thank you
+\item
   \textbf{Image/Plot Idea:} The official UCI ML Repository logo and your

### 2025-12-07 model_fit.Rmd
slides update3

--- 2025-12-03 model_fit.Rmd
@@
 
-load clean data
+load libraries and clean data
+
+```{r, results='hide'}
+library(readr)
+library(tidyverse)
+library(boot)
+library(glmnet)
+library(caret)
+library(pROC)
+library(MASS)
+library(ggplot2)
+
+set.seed(445)
+
+bank_full <- read_delim("data/bank-full_1.csv", 
+    delim = ";", escape_double = FALSE, trim_ws = TRUE)
+```
 
 ```{r}
+bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))
+
+bank_full <- bank_full %>%
+  mutate(
+    job = as.factor(job),
+    marital = as.factor(marital),
+    education = as.factor(education),
+    default = as.factor(default),
+    housing = as.factor(housing),
+    loan = as.factor(loan),
+    contact = as.factor(contact),
+    month = as.factor(month),
+    poutcome = as.factor(poutcome)) %>% 
+  mutate(previously_contacted = as.factor(ifelse(pdays == -1, "no", "yes")))
 
@@
 
-## 2.3 Train/Test split
-- Explain purpose of an honest out-of-sample evaluation.
-- k-fold CV, k=5
+This initial phase is a crucial part of the modeling pipeline.
+This includes data cleaning, and the training/test split.
+The primary purpose of the training/test split is to perform an honest out-of-sample evaluation, a process where a model is fit exclusively on the training data and its performance is later assessed on the unseen test data to provide an estimate of the test error.
+Converting categorical variables to factors is necessary for R's statistical modeling functions, such as glm, to correctly interpret them as nominal predictors and create the appropriate dummy variables for inclusion in the model.
+The new previously_contacted variable also handles a common issue in this data set, where a value of -1 for pdays indicates the client was not previously contacted, effectively creating a binary feature.
+
+## Train/Test split
+To estimate the test error rate, the data is partitioned into a training set (80%) and a test set (20%).
+The model is fit only on the training data, and its performance is later evaluated on the unseen test data.
+The createDataPartition function is used to ensure a balanced split (stratified sampling) with respect to the response variable $Y$.
+
+```{r}
+train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
+train_data <- bank_full[train_index, ]
+test_data  <- bank_full[-train_index, ]
+
+cat(paste("Training set size:", nrow(train_data), "\n"))
+cat(paste("Test set size:", nrow(test_data), "\n"))
+```
+
+# 2. Model Selection: Backward Step-wise Selection
+Model selection is employed to identify a parsimonious subset of predictors that minimizes the estimated test error.
+We use backward step-wise selection (Chapter 6), which starts with a full model containing all predictors and iteratively removes the variable that leads to the largest decrease (or smallest increase) in AIC (Akaike Information Criterion).
+AIC is an estimate of the test error, penalizing models with more predictors to favor simplicity.
+
+### Initial Feature Selection
+We first fit a full model, and then use the step() function to perform the selection based on AIC.
+
+```{r}
+full_formula <- y ~ age + job + marital + education + default + balance + 
+  housing + loan + contact + day + month + duration + campaign + 
+  pdays + previous + poutcome + previously_contacted
+
+glm.full <- glm(full_formula, data = train_data, family = "binomial")
+
+# Perform backward selection
+glm.step <- step(glm.full, direction = "backward", trace = 0)
+
+selected_formula <- formula(glm.step)
+selected_formula
+```
+
+The aim of this process is to reduce the variance of the coefficient estimates by eliminating predictors that contribute little to the model's fit.
+The resulting selected_formula retains the most important predictors (job, housing, poutcome, duration, campaign) while eliminating others (age, default, pdays).
+Variables like duration (the time of the last contact) are known to be strong predictors, while pdays was made redundant by the engineered previously_contacted feature.
+The selected model aims for a better trade-off between bias (which may increase slightly by removing variables) and variance (which decreases, leading to a potentially lower overall test error).
+
+# 3. Baseline Logistic Regression Model
+We now fit the baseline model using the features identified through backward selection.
+Logistic regression (Chapter 4) models the probability of the response $Y=1$ (a 'yes' subscription) as:$$P(Y=1 | X) = \frac{e^{\beta_0 + \beta_1 X_1 + \dots + \beta_p X_p}}{1 + e^{\beta_0 + \beta_1 X_1 + \dots + \beta_p X_p}}$$The coefficients $\beta_j$ are interpreted as the change in the log-odds of a positive response associated with a one-unit increase in the predictor $X_j$.
+
+```{r}
+glm.baseline <- glm(selected_formula, data = train_data, family = "binomial")
+summary(glm.baseline)
+```
+
+Assessing Non-Linearity with Residuals
+A fundamental assumption in logistic regression is the linearity in the log-odds.
+We check this assumption for the continuous predictors by plotting the deviance residuals against the predictor values.
+Systematic deviation of the LOESS (locally estimated scatterplot smoothing) curve from zero suggests a need for a non-linear transformation or a more flexible model.
+
+```{r}
+# Extract deviance residuals
+residuals <- residuals(glm.baseline, type = "deviance")
+train_data$residuals <- residuals
+
+p1 <- ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+p1
+```
+
+The plot of the deviance residuals against balance shows that the LOESS curve deviates non-trivially from the zero line.
+This non-flat pattern suggests that the relationship between balance and the log-odds is non-linear, violating a core assumption of the simple linear logistic regression.
+This systematic error (or high bias) indicates that the model's fit could be substantially improved by incorporating a non-linear transformation of balance, which would better capture the true underlying relationship.
+
+```{r}
+glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
+    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
+summary(glm.tran)
+
+residuals <- residuals(glm.tran, type = "deviance")
+train_data$residuals <- residuals
+
+ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+```
+
+### Cross-Validation 
+for Test Error EstimationTo obtain a reliable, low-variance estimate of the test error, we employ $K=5$ cross-validation (Chapter 5, Section 5.1.3) using the cv.glm() function.
+The cross-validated classification error rate is often used as a direct estimator of the true test error.
+
+```{r}
+cost <- function(r, pi) {
+  mean(r != (pi > 0.5))
+}
+
+cv.error.baseline <- cv.glm(
+  data = train_data, 
+  glmfit = glm.tran, 
+  cost = cost, 
+  K = 5
+)
+
+cv_error_rate <- cv.error.baseline$delta[2]
+round(cv_error_rate, 5)
+```
+
+The cross-validation process provides an estimate of the test error without requiring us to use the final, pristine test set.
+The resulting $\mathbf{K}$-fold CV error rate of round(cv_error_rate, 5) serves as a key metric for model selection.
+This value represents the expected proportion of observations that would be misclassified by the baseline model if deployed on new data.
+
+# 4. Regularized Logistic Regression
+To further reduce the model's variance and potentially find a model with a lower test error—especially in the presence of many correlated predictors—we apply regularization (Chapter 6).
+We use the glmnet package, which requires a design matrix for predictors ($X$) and a vector for the response ($Y$).
+
+```{r}
+X_full <- model.matrix(y ~ . - 1, data = bank_full)
+Y_full <- as.numeric(bank_full$y) - 1
+
+X_train <- X_full[train_index, ]
+Y_train <- Y_full[train_index]
+
+X_test <- X_full[-train_index, ]
+Y_test <- Y_full[-train_index] 
+```
+
+### Ridge Regression ($\ell_2$ Penalty)
+Ridge regression uses an $\ell_2$ penalty ($\alpha=0$) to shrink the coefficient estimates towards zero.
+This is a form of shrinkage that helps with multicollinearity and reduces variance, but it does not perform feature selection (no coefficients are set exactly to zero).
+The optimal tuning parameter $\lambda$ is selected via cross-validation.
+
+```{r}
+cv.ridge <- cv.glmnet(X_train, Y_train, 
+                      alpha = 0, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.ridge <- cv.ridge$lambda.min
+
+plot(cv.ridge)
+```
+
+The plot shows the cross-validated misclassification error (the measure of performance) as a function of $\log(\lambda)$.
+The $\lambda_{\min}$ chosen, which minimizes this cross-validated error, is r round(lambda.min.ridge, 5).
+This value represents the optimal level of shrinkage required to minimize the estimated test error.
+Note that Ridge regression does not set coefficients exactly to zero, meaning it retains all predictors but reduces the magnitude of their estimates.
+
+### LASSO Regression ($\ell_1$ Penalty)
+LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$).
+A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6).
+
+```{r}
+cv.lasso <- cv.glmnet(X_train, Y_train, 
+                      alpha = 1, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.lasso <- cv.lasso$lambda.min
+
+plot(cv.lasso)
+```
+
+The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, r round(lambda.min.lasso, 5).
+Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.
+
+# 5. Model Assessment and Comparison
+The ultimate goal is to assess the performance of the three models (Baseline, Ridge, and LASSO) on the unseen test set.
+We focus on the classification error rate (or accuracy) and, crucially for this imbalanced dataset, the Sensitivity and Specificity (Chapter 4).
+
+### Test Set Predictions
+
+```{r}
+baseline_prob <- predict(glm.tran, newdata = test_data, type = "response")
+baseline_pred <- factor(ifelse(baseline_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+ridge_prob <- predict(cv.ridge, s = lambda.min.ridge, newx = X_test, type = "response")
+ridge_pred <- factor(ifelse(ridge_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+lasso_prob <- predict(cv.lasso, s = lambda.min.lasso, newx = X_test, type = "response")
+lasso_pred <- factor(ifelse(lasso_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+```
+
+### Confusion Matrices and Metrics
+The confusion matrix (Chapter 4) provides a detailed breakdown of the model's performance by counting True Positives (TP), True Negatives (TN), False Positives (FP), and False Negatives (FN).
+
+```{r}
+cm_baseline <- confusionMatrix(baseline_pred, test_data$y, positive = "yes")
+cm_ridge <- confusionMatrix(ridge_pred, test_data$y, positive = "yes")
+cm_lasso <- confusionMatrix(lasso_pred, test_data$y, positive = "yes")
+
+metrics <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  Accuracy = c(cm_baseline$overall['Accuracy'], cm_ridge$overall['Accuracy'], cm_lasso$overall['Accuracy']),
+  Sensitivity = c(cm_baseline$byClass['Sensitivity'], cm_ridge$byClass['Sensitivity'], cm_lasso$byClass['Sensitivity']),
+  Specificity = c(cm_baseline$byClass['Specificity'], cm_ridge$byClass['Specificity'], cm_lasso$byClass['Specificity'])
+)
+metrics
+```
+
+All three models exhibit a high overall Accuracy (around 0.89), which is expected given the significant class imbalance (Prevalence $\approx 0.11$).
+However, the critical metric, Sensitivity (the True Positive Rate for the minority class, 'yes'), is very low (around 0.17).
+This result, consistent across all three models, indicates that they struggle to correctly identify subscribers, leading to many False Negatives (FN).
+Conversely, the high Specificity (True Negative Rate $\approx 0.98$) means the models are excellent at identifying non-subscribers.
+The regularization methods (Ridge and LASSO) failed to provide any significant improvement in these metrics, suggesting that the primary limitation is one of high bias (due to insufficient flexibility) rather than high variance.
+
+### ROC Curve and AUC Comparison
+The Receiver Operating Characteristic (ROC) curve and the Area Under the Curve (AUC) (Chapter 4, Section 4.4) are threshold-independent measures of model discrimination.
+The AUC represents the probability that the model will rank a randomly chosen positive observation higher than a randomly chosen negative observation.
+
+```{r}
+roc_baseline <- roc(test_data$y, baseline_prob)
+roc_ridge <- roc(test_data$y, as.numeric(ridge_prob))
+roc_lasso <- roc(test_data$y, as.numeric(lasso_prob))
+
+roc_data <- data.frame(
+  fpr = c(1 - roc_baseline$specificities, 1 - roc_ridge$specificities, 1 - roc_lasso$specificities),
+  tpr = c(roc_baseline$sensitivities, roc_ridge$sensitivities, roc_lasso$sensitivities),
+  Model = factor(c(rep("Baseline Logit", length(roc_baseline$specificities)),
+                   rep("Ridge Logit", length(roc_ridge$specificities)),
+                   rep("LASSO Logit", length(roc_lasso$specificities))),
+                 levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit"))
+)
+
+ggplot(roc_data, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(linewidth = 1) +
+  geom_abline(linetype = "dashed", color = "grey") +
+  labs(
+    title = "ROC Curve Comparison for Logistic Models",
+    x = "False Positive Rate (1 - Specificity)",
+    y = "True Positive Rate (Sensitivity)"
+  ) +
+  theme_minimal()
+
+auc_values <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  AUC = c(auc(roc_baseline), auc(roc_ridge), auc(roc_lasso))
+)
+auc_values
+```
+
+All three models yield a very similar AUC of approximately $0.90$.
+The ROC curves are virtually identical, demonstrating that both the Ridge and LASSO regularization penalties provided negligible improvement in the overall power of the model to distinguish between the two classes when compared to the simpler Baseline logistic regression.
+This consistent performance confirms the observation from the confusion matrices: while the logistic models are good at discrimination (AUC $> 0.5$), they lack the flexibility needed to substantially reduce the bias and improve the classification of the highly valuable minority class.
 
 ---
-
-# 3. Modeling
-
-## 3.1 Logistic Regression (Baseline Model)
-- Fit logistic regression.
-- k-fold CV for estimated test error.
-- Assess linearity and consider interactions if appropriate.
-- Optional: **regularized logistic regression** (Ridge/Lasso) using CV-selected $\lambda$.
-
-## 3.2 Linear Discriminant Analysis (LDA)
-- Fit LDA model.
-- k-fold CV for estimated test error.
-- Check assumptions (normality, equal covariance).
-- Discuss bias-variance
-
-## 3.3 Quadratic Discriminant Analysis (QDA)
-- Fit QDA model.
-- k-fold CV for estimated test error.
-- Discuss overfitting risk with higher model flexibility.
-- Compare performance with LDA.
 

### 2025-12-07 model_fit.Rmd
added models that got deleted and make sure code worked, and graphics show

@@
 library(ggplot2)
+library(dplyr)
 
@@
 
-## 3.4 K-Nearest Neighbors (KNN)
 - Standardize numeric predictors.
@@
 
+## 3.4 K-Nearest Neighbors (KNN)
+KNN is a nonparametric method that relies on feature scaling and distance-based classification.
+
+Prepare scaled numeric predictors
+```{r}
+library(class)
+
+numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]
+numeric_vars <- setdiff(numeric_vars, "residuals")
+
+train_x <- train_data[, numeric_vars]
+test_x  <- test_data[, numeric_vars]
+
+train_x_scaled <- scale(train_x)
+test_x_scaled <- scale(
+  test_x,
+  center = attr(train_x_scaled, "scaled:center"),
+  scale  = attr(train_x_scaled, "scaled:scale")
+)
+
+train_y <- train_data$y
+test_y  <- test_data$y
+
+```
+
+Cross-validation to choose the best k
+
+```{r}
+k_values <- seq(1, 25, 2)
+cv_accuracy <- numeric(length(k_values))
+
+set.seed(445)
+folds <- createFolds(train_y, k = 5)
+
+for (i in seq_along(k_values)) {
+  k <- k_values[i]
+  fold_acc <- c()
+
+  for (f in folds) {
+    pred <- knn(
+      train = train_x_scaled[-f, ],
+      test  = train_x_scaled[f, ],
+      cl    = train_y[-f],
+      k = k
+    )
+    fold_acc <- c(fold_acc, mean(pred == train_y[f]))
+  }
+
+  cv_accuracy[i] <- mean(fold_acc)
+}
+
+best_k <- k_values[which.max(cv_accuracy)]
+best_k
+
+```
+
+Final KNN prediction and metrics
+
+```{r}
+knn_pred <- knn(
+  train = train_x_scaled,
+  test  = test_x_scaled,
+  cl    = train_y,
+  k = best_k
+)
+
+cm_knn <- confusionMatrix(knn_pred, test_y, positive = "yes")
+cm_knn
+
+knn_prob <- as.numeric(knn_pred == "yes")
+roc_knn <- roc(test_y, knn_prob)
+auc_knn <- auc(roc_knn)
+auc_knn
+
+```
+
 ## 3.5 Generalized Additive Models (GAM)
-- Fit GAM with smoothing splines for continuous variables.
-- Use CV to choose smoothing parameters.
-- Examine partial effect plots for interpretability.
+
+GAM allows flexible nonlinear fits through smoothing splines.
+
+```{r}
+library(gam)
+
+gam_model <- gam(
+  y ~ s(age) + s(balance) + s(duration) + s(campaign) + s(previous),
+  data = train_data,
+  family = binomial
+)
+
+summary(gam_model)
+
+gam_prob <- predict(gam_model, test_data, type = "response")
+gam_pred <- ifelse(gam_prob > 0.5, "yes", "no") |> factor(levels = c("no","yes"))
+
+cm_gam <- confusionMatrix(gam_pred, test_y, positive = "yes")
+cm_gam
+
+roc_gam <- roc(test_y, gam_prob)
+auc_gam <- auc(roc_gam)
+auc_gam
+
+```
 
 ## 3.6 Boosting (Most Flexible Model)
-- Fit gradient boosting classifier (e.g., `gbm` or `xgboost`).
-- Use cross-validation for:
-  - number of trees,
-  - interaction depth,
-  - learning rate.
-- Discuss complexity vs interpretability tradeoff. This model will be our benchmark non-interpretable model to compare interpretable models fitted above.
+
+Boosting builds many small trees sequentially, improving mistakes made by prior trees.
+
+```{r}
+
+if ("residuals" %in% colnames(train_data)) {
+  train_data$residuals <- NULL
+}
+
+train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
+test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)
+
+library(gbm)
+set.seed(445)
+
+boost_model <- gbm(
+  y_num ~ . - y,         
+  data = train_data,
+  distribution = "bernoulli",
+  n.trees = 400,       
+  shrinkage = 0.01,
+  interaction.depth = 3,
+  bag.fraction = 0.5,
+  verbose = FALSE
+)
+
+best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = FALSE)
+best_trees
+
+```
+
+Prediction + Metrics
+```{r}
+boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
+boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no","yes"))
+
+cm_boost <- confusionMatrix(boost_pred, test_y, positive = "yes")
+cm_boost
+
+roc_boost <- roc(test_y, boost_prob)
+auc_boost <- auc(roc_boost)
+auc_boost
+
+```
 
@@
 
+```{r}
+final_results <- data.frame(
+  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
+  Accuracy = c(
+    cm_baseline$overall["Accuracy"],
+    cm_ridge$overall["Accuracy"],
+    cm_lasso$overall["Accuracy"],
+    cm_knn$overall["Accuracy"],
+    cm_gam$overall["Accuracy"],
+    cm_boost$overall["Accuracy"]
+  ),
+  Sensitivity = c(
+    cm_baseline$byClass["Sensitivity"],
+    cm_ridge$byClass["Sensitivity"],
+    cm_lasso$byClass["Sensitivity"],
+    cm_knn$byClass["Sensitivity"],
+    cm_gam$byClass["Sensitivity"],
+    cm_boost$byClass["Sensitivity"]
+  ),
+  AUC = c(
+    auc(roc_baseline),
+    auc(roc_ridge),
+    auc(roc_lasso),
+    auc_knn,
+    auc_gam,
+    auc_boost
+  )
+)
+
+final_results
+
+```
+
 ## 4.2 Comparative Evaluation
@@
 - Discuss interpretability vs predictive quality.
+
+Combined ROC Curve
+```{r}
+roc_all <- rbind(
+  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
+  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
+  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
+  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
+  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
+  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
+)
+
+ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(size = 1) +
+  geom_abline(linetype = "dashed") +
+  labs(
+    title = "ROC Curve Comparison for All Models",
+    x = "False Positive Rate",
+    y = "True Positive Rate"
+  ) +
+  theme_minimal()
+
+```

### 2025-12-07 Slides.Rmd
Updated Slides KNN and boosted trees

@@
 * **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
+```{r}
+library(dplyr)
+library(class)
+library(caret)
+
+set.seed(123)
+train_index <- sample(nrow(bank_full), 0.8 * nrow(bank_full))
+bank_train <- bank_full[train_index, ]
+bank_test  <- bank_full[-train_index, ]
+
+numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
+numeric_vars <- setdiff(numeric_vars, "y")
+
+train_x <- bank_train[, numeric_vars, drop = FALSE]
+test_x  <- bank_test[, numeric_vars, drop = FALSE]
+
+train_x <- bank_train[, numeric_vars]
+test_x  <- bank_test[, numeric_vars]
+
+train_x_scaled <- scale(train_x)
+test_x_scaled <- scale(
+  test_x,
+  center = attr(train_x_scaled, "scaled:center"),
+  scale  = attr(train_x_scaled, "scaled:scale")
+)
+
+train_y <- bank_train$y
+test_y  <- bank_test$y
+
+best_k <- 5
+
+par(mfrow = c(1, 2))
+
+plot(
+  bank_full$age, bank_full$balance,
+  xlab = "Age", ylab = "Balance",
+  main = "Unscaled Features"
+)
+
+scaled_two <- scale(bank_full[, c("age", "balance")])
+
+plot(
+  scaled_two[, "age"], scaled_two[, "balance"],
+  xlab = "Scaled Age", ylab = "Scaled Balance",
+  main = "Scaled Features (mean = 0, sd = 1)"
+)
+
+par(mfrow = c(1, 1))
+
+i <- 1  
+test_point <- test_x_scaled[i, , drop = FALSE]
+
+distances <- apply(
+  train_x_scaled,
+  1,
+  function(row) sqrt(sum((row - test_point)^2))
+)
+
+nn_idx <- order(distances)[1:best_k]
+
+knn_demo_pred <- knn(
+  train = train_x_scaled,
+  test  = test_point,
+  cl    = train_y,
+  k     = best_k
+)
+
+cat("Test customer index:", i, "\n")
+cat("KNN predicted class:", knn_demo_pred, "\n")
+cat("True class:", test_y[i], "\n\n")
+
+cat("Neighbor outcome breakdown:\n")
+print(table(train_y[nn_idx]))
+
+cat("\nNearest neighbor details:\n")
+print(head(bank_train[nn_idx, c("age", "balance", "duration", "campaign", "y")]))
+
+```
+```{r}
+library(class)
+
+numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
+numeric_vars <- setdiff(numeric_vars, "y")
+
+train_x <- bank_train[, numeric_vars, drop = FALSE]
+test_x  <- bank_test[, numeric_vars, drop = FALSE]
+
+train_x_scaled <- scale(train_x)
+test_x_scaled <- scale(
+  test_x,
+  center = attr(train_x_scaled, "scaled:center"),
+  scale  = attr(train_x_scaled, "scaled:scale")
+)
+
+train_y <- bank_train$y
+test_y  <- bank_test$y
+
+knn_pred <- knn(
+  train = train_x_scaled,
+  test  = test_x_scaled,
+  cl    = train_y,
+  k     = 5
+)
+
+knn_accuracy <- mean(knn_pred == test_y)
+knn_error <- 1 - knn_accuracy
+cat("KNN Test Accuracy:", round(knn_accuracy * 100, 2), "%\n")
+cat("KNN Test Error:", round(knn_error * 100, 2), "%\n")
+
+```
+```
 
@@
 * **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
+```{r}
+library(class)
+library(caret)
+
+set.seed(123)
+
+k_values <- 1:25                  
+cv_error <- numeric(length(k_values))
+
+folds <- createFolds(train_y, k = 5, returnTrain = FALSE)
+
+for (i in seq_along(k_values)) {
+  k <- k_values[i]
+  fold_err <- numeric(length(folds))
+
+  for (j in seq_along(folds)) {
+    test_idx  <- folds[[j]]
+    train_idx <- setdiff(seq_along(train_y), test_idx)
+
+    fold_train_x <- train_x_scaled[train_idx, , drop = FALSE]
+    fold_test_x  <- train_x_scaled[test_idx,  , drop = FALSE]
+    fold_train_y <- train_y[train_idx]
+    fold_test_y  <- train_y[test_idx]
+
+    pred <- knn(
+      train = fold_train_x,
+      test  = fold_test_x,
+      cl    = fold_train_y,
+      k     = k
+    )
+
+    fold_err[j] <- mean(pred != fold_test_y)
+  }
+
+  cv_error[i] <- mean(fold_err)
+}
+
+best_k <- k_values[which.min(cv_error)]
+best_k
+plot(
+  k_values, cv_error,
+  type = "b",
+  xlab = "Number of neighbors (k)",
+  ylab = "5-fold CV Test Error",
+  main = "KNN: 5-fold CV Test Error vs k"
+)
+
+abline(v = best_k, lty = 2)
+
+```
 
@@
 * **Image/Plot Idea:** A brief table comparing the **AUC** and **Recall** of the best LogReg model and the final KNN model.
+
+```{r}
+library(pROC)
+library(caret)
+
+set.seed(123)
+
+knn_final <- knn(
+  train = train_x_scaled,
+  test  = test_x_scaled,
+  cl    = train_y,
+  k     = best_k,
+  prob  = TRUE
+)
+
+knn_prob_attr <- attr(knn_final, "prob")
+knn_prob_yes <- ifelse(knn_final == "yes", knn_prob_attr, 1 - knn_prob_attr)
+
+test_y_fac <- factor(test_y, levels = c("no", "yes"))
+
+knn_roc  <- roc(response = test_y_fac, predictor = knn_prob_yes, levels = c("no", "yes"))
+knn_auc  <- as.numeric(auc(knn_roc))
+
+knn_cm      <- confusionMatrix(knn_final, test_y_fac, positive = "yes")
+knn_recall  <- knn_cm$byClass["Recall"]
+
+logit_model <- glm(y ~ ., data = bank_train, family = binomial)
+
+logit_prob <- predict(logit_model, newdata = bank_test, type = "response")
+logit_pred <- ifelse(logit_prob >= 0.5, "yes", "no")
+logit_pred <- factor(logit_pred, levels = c("no", "yes"))
+
+logit_roc <- roc(response = test_y_fac, predictor = logit_prob, levels = c("no", "yes"))
+logit_auc <- as.numeric(auc(logit_roc))
+
+logit_cm     <- confusionMatrix(logit_pred, test_y_fac, positive = "yes")
+logit_recall <- logit_cm$byClass["Recall"]
+
+results <- data.frame(
+  Model  = c("Logistic Regression", paste0("KNN (k = ", best_k, ")")),
+  AUC    = c(logit_auc, knn_auc),
+  Recall = c(as.numeric(logit_recall), as.numeric(knn_recall))
+)
+
+results$AUC    <- round(results$AUC, 3)
+results$Recall <- round(results$Recall, 3)
+
+results
+
+```
 
@@
 * **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
+```{r}
+library(gbm)
+
+bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
+bank_test$y_num  <- ifelse(bank_test$y  == "yes", 1, 0)
+
+set.seed(123)
+
+boost_model <- gbm(
+  formula = y_num ~ . - y,       
+  data = bank_train,
+  distribution = "bernoulli",   
+  n.trees = 1000,                
+  interaction.depth = 3,         
+  shrinkage = 0.01,             
+  n.minobsinnode = 10,
+  verbose = FALSE
+)
+
+boost_prob <- predict(
+  boost_model,
+  newdata = bank_test,
+  n.trees = 1000,
+  type = "response"
+)
+
+boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")
+
+boost_accuracy <- mean(boost_pred == bank_test$y)
+boost_error    <- 1 - boost_accuracy
+
+cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
+cat("Boosting Test Error:",   round(boost_error * 100, 2), "%\n")
+
+plot(
+  boost_model$train.error,
+  type = "l",
+  xlab = "Number of Trees (Iterations)",
+  ylab = "Training Loss",
+  main = "Boosting: Sequential Error Reduction"
+)
+
+```
 
@@
 * **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
+```{r}
+boost_model <- gbm(
+  formula = y_num ~ . - y,
+  data = bank_train,
+  distribution = "bernoulli",
+  n.trees = 1000,
+  interaction.depth = 3,
+  shrinkage = 0.01,
+  n.minobsinnode = 10,
+  verbose = FALSE
+)
+
+M_values <- seq(50, 1000, by = 50)
+test_error <- numeric(length(M_values))
+
+for (i in seq_along(M_values)) {
+  M <- M_values[i]
+
+  prob_M <- predict(
+    boost_model,
+    newdata = bank_test,
+    n.trees = M,
+    type = "response"
+  )
+
+  pred_M <- ifelse(prob_M > 0.5, "yes", "no")
+  test_error[i] <- mean(pred_M != bank_test$y)
+}
+
+best_idx <- which.min(test_error)
+best_M   <- M_values[best_idx]
+
+cat("Best number of trees based on Test Error:", best_M, "\n")
+
+plot(
+  M_values, test_error,
+  type = "b",
+  xlab = "Number of Trees (M)",
+  ylab = "Test Error",
+  main = "Boosting: Test Error vs Number of Trees"
+)
+
+abline(v = best_M, lty = 2)
+
+```
 
@@
 * **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
+```{r}
+library(pROC)
+
+boost_prob <- predict(
+  boost_model,
+  newdata = bank_test,
+  n.trees = 1000,       
+  type = "response"
+)
+
+test_y_fac <- factor(bank_test$y, levels = c("no", "yes"))
+
+boost_roc <- roc(
+  response  = test_y_fac,
+  predictor = boost_prob,
+  levels    = c("no", "yes")
+)
+
+boost_auc <- as.numeric(auc(boost_roc))
+cat("Boosting Test AUC:", round(boost_auc, 3), "\n")
+
+vi <- summary(boost_model, plotit = FALSE)
+
+top_n <- min(10, nrow(vi))
+vi_top <- vi[1:top_n, ]
+
+barplot(
+  height = vi_top$rel.inf,
+  names.arg = vi_top$var,
+  las = 2,
+  ylab = "Relative Influence",
+  main = "Boosting: Variable Importance (Top Features)"
+)
+
+```
 

### 2025-12-08 model_fit.Rmd
added descripton and added the final visuals and summary of models

@@
 
-- Standardize numeric predictors.
-- Perform CV to tune **k** .
-- Report best k and corresponding CV error.
-- Discuss bias-variance behavior.
-
 ## 3.4 K-Nearest Neighbors (KNN)
@@
 
-Prepare scaled numeric predictors
+## Prepare scaled numeric predictors
 ```{r}
@@
 
-Cross-validation to choose the best k
+## Cross-validation to choose the best k
 
@@
 
-Final KNN prediction and metrics
+Using 5-fold cross-validation across odd values of kk from 1 to 25, the model achieved its highest average classification accuracy at k=23.
+This indicates that a relatively large number of neighbors provides the best bias–variance trade-off for this dataset.
+Because the Bank Marketing data is high-dimensional and noisy, smaller values of kk likely overfit the training data, while larger k values smooth out noise and generalize better.
+The fact that k=23 performs best suggests the underlying decision boundary is fairly smooth, and KNN benefits from stronger averaging across many neighbors.
+
+## Final KNN prediction and metrics
 
@@
 ```
+
+The KNN model with k=23 achieved an accuracy of 0.8911, which is comparable to the logistic regression models.
+However, due to the class imbalance, accuracy alone is misleading.
+The model’s sensitivity is very low (0.26), meaning it correctly identifies only about 26% of clients who actually subscribed.
+In contrast, the specificity is very high (0.97), showing the model is much better at identifying non-subscribers.
+The AUC of 0.617 confirms that KNN provides only limited discriminatory power for this dataset, performing slightly above random chance but worse than logistic regression and boosting models.
 
@@
 
+The GAM model allows nonlinear relationships through smoothing splines and the ANOVA results show that all five smoothed predictors—age, balance, duration, campaign, and previous—have statistically significant nonlinear effects on subscription probability (all p < 0.001).
+The model maintains an accuracy of 0.8888, similar to previous models, but like the other approaches, it struggles with the minority “yes” class.
+The sensitivity is 0.216, meaning the GAM correctly identifies only about 21% of subscribers, while specificity is very high at 0.978, reflecting accurate detection of non-subscribers.
+Despite the low sensitivity, the model achieves an AUC of 0.854, indicating strong overall discriminatory power.
+This suggests the GAM captures meaningful nonlinear patterns, but still struggles to translate these patterns into correct classifications for the rare positive cases.
+
 ## 3.6 Boosting (Most Flexible Model)
@@
   distribution = "bernoulli",
-  n.trees = 400,       
+  n.trees = 1500,       
   shrinkage = 0.01,
@@
 ```
+
+The boosting model was trained using 1500 trees, and the out-of-bag (OOB) performance evaluation selected all 1500 trees as optimal.
+This indicates that the model continued to improve throughout the entire boosting sequence, with no evidence of overfitting detected by the OOB error.
+The small residual standard error from the LOESS smoother (2.7 × 10⁻⁵) shows that the OOB improvement curve is smooth, stable, and consistently positive, suggesting reliable incremental improvement.
+The model's behavior reflects the strengths of boosting: its ability to gradually reduce bias and capture complex nonlinear and interaction effects in the data.
+The fact that performance continues to improve up to 1500 trees suggests that boosting may be the most flexible and predictive model among those considered.
 
@@
 ```
+The boosting model achieved the strongest performance among all models evaluated.
+Its overall accuracy on the test set was 0.9047, exceeding both the baseline logistic regression and other flexible models such as KNN and GAM.
+Most importantly, boosting substantially improved sensitivity, correctly identifying 41.15% of actual subscribers (“yes”), compared to roughly 21–26% for the other models.
+This improvement indicates that boosting is better able to capture the complex, nonlinear structure of the minority class, which is critical in this imbalanced classification problem.
+Specificity remained high (96.99%), showing that the model continued to classify non-subscribers accurately.
+Boosting also demonstrated stronger balance across classes, with a Balanced Accuracy of 0.6907, the highest among all models tested.
+Additionally, the model’s AUC value (insert your computed AUC here once available) reflects its enhanced ability to rank subscribers above non-subscribers.
+Together, these results reinforce that boosting provides the best predictive performance in terms of both overall classification and minority-class detection, aligning with expectations for a highly flexible, low-bias model.
 
@@
 ## 4.1 Test Set Performance
-For each model:
-- Confusion matrix  
-  - Accuracy and misclassification rate  
-  - Sensitivity / specificity (important for minority class)  
-- ROC curves and AUC
 
@@
 ```
+The overall model comparison confirms that Boosting is the strongest predictive model in this project.
+It achieved the highest accuracy (0.9047), the highest sensitivity (0.4115), and the highest AUC (0.9266).
+These metrics show that boosting not only performs well overall but is also the most effective at identifying the minority “yes” class—an essential goal in the bank marketing problem.
+
+The family of logistic models (Logistic, Ridge, and LASSO) all attained similar accuracy levels (≈0.898–0.900) and high AUC values (≈0.907), indicating solid discrimination ability.
+However, their sensitivity values (0.28–0.33) remain relatively low, meaning they miss many potential subscribers compared with boosting.
+
+KNN performed worse overall, with lower accuracy (0.8911) and the lowest AUC (0.617), showing that it struggled with this high-dimensional, imbalanced dataset.
+GAM demonstrated moderate AUC (0.854), but its low sensitivity (0.2157) suggests it was unable to flexibly model the decision boundary needed to capture the minority class.
+
+Overall, boosting delivered the most balanced and effective performance across all key metrics, supporting the conclusion that flexible ensemble methods are better suited for the nonlinear patterns present in the bank marketing data.
 

### 2025-12-08 Slides.Rmd
Fixed KNN and Boosting Trees

@@
 * **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
-```{r}
+```{r, echo=FALSE, results='asis'}
+
 library(dplyr)
@@
 ```
-```{r}
+```{r, echo=FALSE, results='asis'}
 library(class)
@@
 * **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
-```{r}
+```{r, echo=FALSE, results='asis'}
 library(class)
@@
 
-```{r}
+```{r, echo=FALSE, results='asis'}
 library(pROC)
@@
 * **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
-```{r}
+```{r, echo=FALSE, results='asis'}
 library(gbm)
@@
 * **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
-```{r}
+```{r, echo=FALSE, results='asis'}
 boost_model <- gbm(
@@
 * **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
-```{r}
+```{r, echo=FALSE, results='asis'}
 library(pROC)

### 2025-12-08 Slides.Rmd
eda slides

@@
 
-## The Bank Marketing Dataset (Slide 2)
-* **Discuss:** The dataset structure: $\approx 45,000$ client observations from a Portuguese banking institution's telemarketing campaigns.
-* **Discuss:** Brief overview of the four main predictor categories: Client Demographics, Financial Indicators, Campaign Features, and Past Outcomes.
-Mention the paper had more features, fro privacy reasons the public data set has less
-* **Discuss:** The Target Variable ($y$) is binary: Did the client subscribe to a term deposit (Yes/No)?
-* **Discuss:** Emphasize the severity of the class imbalance (e.g., typically $\approx 11\%$ 'Yes' subscriptions). This may yield higher false negatives.
-* **Image/Plot Idea:** A small table summarizing the list of the 4 key variable groups (e.g., Demographics, Financial, Campaign, Past). 
-* **Image/Plot Idea:** A **Bar Plot** showing the distribution of the target variable ($y$), visually highlighting the imbalance.
+## The Bank Marketing Dataset
 
-## Problem Statement & Methodology Overview (Slide 3)
-* **Discuss:** The project aims for a **Comparative Study**—comparing models based on two criteria: **Predictive Power** and **Interpretability**.
-* **Discuss:** The project goal is **Binary Classification**: Building statistical models that are both **predictive** and **interpretable** for management.
-* **Discuss:** Established an **80/20 Train/Test Split** and used **5-Fold Cross-Validation (CV)** on the training set for tuning and estimating out-of-sample error.
-* **Image/Plot Idea:** An icon representing **Predictive Power** next to an icon representing **Interpretability** (The two core goals).
+* **45,000 client observations** from Portuguese bank telemarketing  
+* **Target variable ($y$):** Did client subscribe to term deposit? (Yes/No)  
+* **Class imbalance:** Only ≈11% "Yes" subscriptions  
+* **Four main predictor groups:** demographics, financial indicators, campaign features, past outcomes
+
+## Term Deposit Class Imbalance
+
+* **Most clients say NO; only ≈11% say YES**
+
+```{r slide2-plot, echo=FALSE, fig.align="center", out.width="90%"}
+knitr::include_graphics("imbalance_plot.png")
+```
+
+## EDA Finding 1: Contact Method Matters
+
+```{r eda1-plot, echo=FALSE, fig.align="center", out.width="90%"}
+knitr::include_graphics("contact_plot.png")
+
+```
+
+## EDA Finding 2: Previous Success Predicts Future Success
+
+```{r eda2-plot, echo=FALSE, fig.align="center", out.width="90%"}
+knitr::include_graphics("poutcome_plot.png")
+```
+
+## EDA Finding 3: Too Many Calls Hurts Results
+
+```{r eda3-plot, echo=FALSE, fig.align="center", out.width="90%"}
+knitr::include_graphics("campaign_plot.png")
+```
+
+## Problem Statement & Methodology Overview
+
+* **Goal:** Compare models on two criteria:  
+  * **Predictive Power** – ability to correctly find YES cases  
+  * **Interpretability** – model is easy to understand
+
+* **Model evaluation:**  
+  * **80/20 Train/Test Split**  
+  * **5-Fold Cross-Validation**
 

### 2025-12-08 plots.Rmd
eda slides

---
title: "plots"
author: "Group 6"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# EDA plots as images

# 1. Class imbalance plot (from page 3 of your PDF)
png("imbalance_plot.png", width = 800, height = 500)
ggplot(bank_full, aes(x = y, fill = y)) +
  geom_bar() +
  labs(title = "Distribution of Term Deposit Subscription",
       x = "Subscribed to Term Deposit?",
       y = "Count") +
  theme_minimal(base_size = 14)
dev.off()

# 2. Contact method plot (from your contact type analysis)
png("contact_plot.png", width = 800, height = 500)
ggplot(bank_full, aes(x = contact, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Subscription Rate by Contact Type",
       x = "Contact Method",
       y = "Proportion") +
  theme_minimal(base_size = 14)
dev.off()

# 3. Job type plot
png("job_plot.png", width = 800, height = 500)
ggplot(bank_full, aes(x = job, fill = y)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Subscription Rate by Job",
       y = "Proportion") +
  theme_minimal(base_size = 12)
dev.off()

# 4. Education plot
png("education_plot.png", width = 800, height = 500)
ggplot(bank_full, aes(x = education, fill = y)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Subscription Rate by Education Level",
       y = "Proportion") +
  theme_minimal(base_size = 14)
dev.off()

# 5. Previous outcome plot
png("poutcome_plot.png", width = 800, height = 500)
ggplot(bank_full, aes(x = poutcome, fill = y)) +
  geom_bar(position = "fill") +
  labs(title = "Subscription Rate by Previous Campaign Outcome",
       y = "Proportion") +
  theme_minimal(base_size = 14)
dev.off()

# 6. Campaign contacts plot
png("campaign_plot.png", width = 800, height = 500)
bank_full %>%
  group_by(campaign) %>%
  summarise(success_rate = mean(y == "yes")) %>%
  ggplot(aes(x = campaign, y = success_rate)) +
  geom_line() +
  geom_point() +
  labs(title = "Subscription Success Rate vs. Number of Campaign Contacts",
       x = "Number of Contacts During This Campaign",
       y = "Success Rate") +
  theme_minimal(base_size = 14)
dev.off()

cat("EDA plots saved as PNG files in your current directory!\n")
cat("Files created:\n")
cat("- imbalance_plot.png\n")
cat("- contact_plot.png\n")
cat("- job_plot.png\n")
cat("- education_plot.png\n")
cat("- poutcome_plot.png\n")
cat("- campaign_plot.png\n")

```

### 2025-12-08 Slides.Rmd
eda slides
Update slides

--- 2025-12-07 Slides.Rmd
@@
     css: styles.css
-
 ---
@@
 
+## boosting
+
+if ("residuals" %in% colnames(train_data)) {
+  train_data$residuals <- NULL
+}
+
+train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
+test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)
+
+library(gbm)
+set.seed(445)
+
+boost_model <- gbm(
+  y_num ~ . - y,         
+  data = train_data,
+  distribution = "bernoulli",
+  n.trees = 1500,       
+  shrinkage = 0.01,
+  interaction.depth = 3,
+  bag.fraction = 0.5,
+  verbose = FALSE
+)
+best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = FALSE)
+
+boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
+boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no","yes"))
+
+cm_boost <- confusionMatrix(boost_pred, test_y, positive = "yes")
+cm_boost
+
+roc_boost <- roc(test_y, boost_prob)
+auc_boost <- auc(roc_boost)
+
+## test 
+final_results <- data.frame(
+  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
+  Accuracy = c(
+    cm_baseline$overall["Accuracy"],
+    cm_ridge$overall["Accuracy"],
+    cm_lasso$overall["Accuracy"],
+    cm_knn$overall["Accuracy"],
+    cm_gam$overall["Accuracy"],
+    cm_boost$overall["Accuracy"]
+  ),
+  Sensitivity = c(
+    cm_baseline$byClass["Sensitivity"],
+    cm_ridge$byClass["Sensitivity"],
+    cm_lasso$byClass["Sensitivity"],
+    cm_knn$byClass["Sensitivity"],
+    cm_gam$byClass["Sensitivity"],
+    cm_boost$byClass["Sensitivity"]
+  ),
+  AUC = c(
+    auc(roc_baseline),
+    auc(roc_ridge),
+    auc(roc_lasso),
+    auc_knn,
+    auc_gam,
+    auc_boost
+  )
+)
+# combined ROC 
+roc_all <- rbind(
+  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
+  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
+  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
+  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
+  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
+  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
+)
+
 ```
 
-## Project Motivation & Business Problem
-* **Discuss:** The bank's need for efficient marketing and the high cost of telemarketing campaigns. The goal is to maximize the success rate by predicting customer subscription.
-* **Discuss:** The challenge of less positive class (imbalanced data) and why prediction accuracy is crucial.
-* **Image/Plot Idea:** A simple, high-level info graphic of a marketing campaign funnel, highlighting the bottleneck/drop-off point.
+## Project Motivation & Business Problem (S)
+- Telemarketing campaigns are expensive, time-consuming, and have very low success rates. The bank wants to improve efficiency by identifying which clients are most likely to subscribe.
+- The dataset has a severe class imbalance: only a small fraction of customers subscribe to the term deposit. This makes prediction difficult and increases the risk of false negatives.
+- Improving predictive accuracy would allow the bank to target fewer customers, reduce costs, and increase return on investment (ROI) for future campaigns.
 
-[Image of a marketing funnel illustrating customer conversion]
+---
+
+```{r, echo=FALSE}
+library(ggplot2)
+
+ggplot(bank_full, aes(x = y, fill = y)) +
+geom_bar() +
+scale_fill_manual(values = c("#2E86C1", "#F39C12")) +
+labs(
+title = "Term Deposit Subscription (Severe Class Imbalance)",
+x = "Subscribed?",
+y = "Count"
+) +
+theme_minimal(base_size = 14)
+
+```
 
@@
 
-## Methodology: Data Preparation (Slide 5)
-* **Discuss:** **Data Wrangling** involved converting categorical variables to factors and strategies for handling 'unknown' levels (treating them as a separate category vs. missing).
-* **Image/Plot Idea:** A simple flow chart showing: Raw Data $\rightarrow$ Wrangling/Cleaning $\rightarrow$ Train/Test Split. 
-* **Image/Plot Idea:** Summary stats or informational plots
+## Methodology: Data Preparation (Slide 5) (s)
 
-[Image of a data science workflow diagram]
+- Converted all categorical variables into factors so models interpret them correctly.
+- Addressed "unknown" values by treating them as their own category to avoid losing information.
+- Created a new feature previously_contacted to properly encode pdays.
+- Split the dataset into 80% training and 20% test using stratified sampling to preserve the class imbalance.
+- Ensured numeric predictors were available for later scaling, tuning, and model evaluation.
+
+---
+
+## Breakdown of Categorical Variable
+```{r, echo=FALSE}
+library(ggplot2)
+
+ggplot(bank_full, aes(x = job)) +
+geom_bar(fill = "steelblue") +
+coord_flip() +
+labs(
+title = "Distribution of Job Categories",
+x = "Job",
+y = "Count"
+) +
+theme_minimal()
+
+```
 
@@
 ## Boosting 1/3: Model Formulation
-* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
-* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
-* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
-```{r}
-library(gbm)
+- Gradient Boosting Machines (GBM) build an ensemble of weak learners (typically shallow decision trees).
+- Each tree is fit to the errors (residuals) of the previous tree — allowing the model to improve sequentially.
+- This iterative correction makes boosting highly flexible and able to capture complex, nonlinear patterns in the data.
+- Because the model is effectively a sum of many small trees, it becomes a strong predictor but loses interpretability relative to logistic regression.
+- Boosting is expected to be our highest-performing model, especially on imbalanced datasets like this one.
 
-bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
-bank_test$y_num  <- ifelse(bank_test$y  == "yes", 1, 0)
+## Boosting 2/3: Hyperparameter Tuning 
+- We tuned the number of trees using test-error curves to prevent overfitting.
+- Test error decreased steadily and leveled off around M = 1500 trees, which we selected as the optimal value.
+- This choice balances model complexity and generalization, giving strong predictive performance without excessive tree growth.
 
-set.seed(123)
+## Boosting 3/3: Evaluation & Importance
 
-boost_model <- gbm(
-  formula = y_num ~ . - y,       
-  data = bank_train,
-  distribution = "bernoulli",   
-  n.trees = 1000,                
-  interaction.depth = 3,         
-  shrinkage = 0.01,             
-  n.minobsinnode = 10,
-  verbose = FALSE
+---
+
+## Results: Comparative Evaluation (Slide 13). (S)
+- We evaluated six models using Accuracy, Sensitivity (Recall), and AUC on the Test Set.
+- Boosting achieved the highest overall accuracy (0.905) and the highest sensitivity (0.412), meaning it identified the most true positives.
+- Logistic, Ridge, and LASSO models performed very similarly in accuracy (~0.898–0.900) and AUC (~0.907–0.908), but with lower sensitivity.
+GAM achieved moderate accuracy and AUC but very low sensitivity.
+KNN performed the worst, with the lowest AUC (0.617), indicating weak separation between classes.
+
+---
+
+```{r, echo=FALSE}
+final_results <- data.frame(
+  Model = c("Logistic", "Ridge", "LASSO", "KNN", "GAM", "Boosting"),
+  Accuracy = c(
+    cm_baseline$overall["Accuracy"],
+    cm_ridge$overall["Accuracy"],
+    cm_lasso$overall["Accuracy"],
+    cm_knn$overall["Accuracy"],
+    cm_gam$overall["Accuracy"],
+    cm_boost$overall["Accuracy"]
+  ),
+  Sensitivity = c(
+    cm_baseline$byClass["Sensitivity"],
+    cm_ridge$byClass["Sensitivity"],
+    cm_lasso$byClass["Sensitivity"],
+    cm_knn$byClass["Sensitivity"],
+    cm_gam$byClass["Sensitivity"],
+    cm_boost$byClass["Sensitivity"]
+  ),
+  AUC = c(
+    auc(roc_baseline),
+    auc(roc_ridge),
+    auc(roc_lasso),
+    auc_knn,
+    auc_gam,
+    auc_boost
+  )
 )
 
-boost_prob <- predict(
-  boost_model,
-  newdata = bank_test,
-  n.trees = 1000,
-  type = "response"
+knitr::kable(final_results, digits = 3)
+
+```
+---
+
+```{r, echo=FALSE}
+roc_all <- rbind(
+  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
+  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
+  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
+  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
+  data.frame(fpr = 1 - roc_gam$specificities,      tpr = roc_gam$sensitivities,      Model = "GAM"),
+  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
 )
 
-boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")
-
-boost_accuracy <- mean(boost_pred == bank_test$y)
-boost_error    <- 1 - boost_accuracy
-
-cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
-cat("Boosting Test Error:",   round(boost_error * 100, 2), "%\n")
-
-plot(
-  boost_model$train.error,
-  type = "l",
-  xlab = "Number of Trees (Iterations)",
-  ylab = "Training Loss",
-  main = "Boosting: Sequential Error Reduction"
-)
-
+ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(size = 1) +
+  geom_abline(linetype = "dashed") +
+  labs(
+    title = "ROC Curve Comparison for All Models",
+    x = "False Positive Rate",
+    y = "True Positive Rate"
+  ) +
+  theme_minimal()
 ```
-
-## Boosting 2/3: Hyperparameter Tuning 
-* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
-* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
-* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
-```{r}
-boost_model <- gbm(
-  formula = y_num ~ . - y,
-  data = bank_train,
-  distribution = "bernoulli",
-  n.trees = 1000,
-  interaction.depth = 3,
-  shrinkage = 0.01,
-  n.minobsinnode = 10,
-  verbose = FALSE
-)
-
-M_values <- seq(50, 1000, by = 50)
-test_error <- numeric(length(M_values))
-
-for (i in seq_along(M_values)) {
-  M <- M_values[i]
-
-  prob_M <- predict(
-    boost_model,
-    newdata = bank_test,
-    n.trees = M,
-    type = "response"
-  )
-
-  pred_M <- ifelse(prob_M > 0.5, "yes", "no")
-  test_error[i] <- mean(pred_M != bank_test$y)
-}
-
-best_idx <- which.min(test_error)
-best_M   <- M_values[best_idx]
-
-cat("Best number of trees based on Test Error:", best_M, "\n")
-
-plot(
-  M_values, test_error,
-  type = "b",
-  xlab = "Number of Trees (M)",
-  ylab = "Test Error",
-  main = "Boosting: Test Error vs Number of Trees"
-)
-
-abline(v = best_M, lty = 2)
-
-```
-
-## Boosting 3/3: Evaluation & Importance 
-* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
-* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
-* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
-```{r}
-library(pROC)
-
-boost_prob <- predict(
-  boost_model,
-  newdata = bank_test,
-  n.trees = 1000,       
-  type = "response"
-)
-
-test_y_fac <- factor(bank_test$y, levels = c("no", "yes"))
-
-boost_roc <- roc(
-  response  = test_y_fac,
-  predictor = boost_prob,
-  levels    = c("no", "yes")
-)
-
-boost_auc <- as.numeric(auc(boost_roc))
-cat("Boosting Test AUC:", round(boost_auc, 3), "\n")
-
-vi <- summary(boost_model, plotit = FALSE)
-
-top_n <- min(10, nrow(vi))
-vi_top <- vi[1:top_n, ]
-
-barplot(
-  height = vi_top$rel.inf,
-  names.arg = vi_top$var,
-  las = 2,
-  ylab = "Relative Influence",
-  main = "Boosting: Variable Importance (Top Features)"
-)
-
-```
-
-## Results: Comparative Evaluation (Slide 13). #MUSSA
-* **Discuss:** Present the **Final Summary Table** comparing the Test Set performance (AUC, Recall, Accuracy) of all models (**Best LogReg**, **KNN**, **Boosting**).
-* **Discuss:** Visually compare the models using the **Combined ROC Curve Plot** and discuss which model dominates the space.
-* **Image/Plot Idea:** The **Combined ROC Curve Plot** comparing all three final models (LogReg, KNN, and Boosting).
 

### 2025-12-08 Slides.Rmd
KNN and Boosted

--- 2025-12-08 Slides.Rmd
@@
 ---
-# 2. K-Nearest Neighbors (KNN) Model
+# 2. K-Nearest Neighbors (KNN) Model (t)
 ---
@@
 ## KNN 1/3: Feature Preparation 
-* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
-* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
-* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
+- KNN relies on distance calculations, so all numeric features must be scaled to contribute equally.
+- Predictors were standardized (mean = 0, SD = 1) so large-range variables don't dominate the distance metric.
+- Because KNN is non-parametric and has no learned coefficients, scaling is essential to ensure fair weighting across features.
+- A before-and-after scatter plot highlights this: unscaled features have mismatched ranges, while standardized features align on comparable scales.
+
+---
+
 ```{r, echo=FALSE, results='asis'}
@@
 
-```
-```{r, echo=FALSE, results='asis'}
 library(class)
@@
 ```
-```
 
 ## KNN 2/3: Hyperparameter Tuning 
-* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
-* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
-* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
-```{r, echo=FALSE, results='asis'}
+- KNN’s only hyperparameter is k, which controls model complexity
+--small k = flexible but high variance; large k = smoother but higher bias
+- Selecting the right k is essential, so we used 5-fold Cross-Validation across multiple k values.
+- The optimal k minimizes the mean CV error, typically seen as the bottom of a U-shaped error curve.
+
+---
+
+```{r, echo=FALSE}
 library(class)
@@
 ## KNN 3/3: Evaluation & Test Performance
-* **Discuss:** The final KNN model (with optimal $k$) was applied to the **Test Set**. Evaluate performance using **AUC** and compare it to the LogReg models.
-* **Discuss:** KNN generally provides a good intermediate balance—often capturing non-linearity better than LogReg but being less computationally expensive than boosting.
-* **Image/Plot Idea:** A brief table comparing the **AUC** and **Recall** of the best LogReg model and the final KNN model.
+- The final KNN model (using the optimized k) was evaluated on the Test Set using AUC and Recall.
+- KNN offers mid-level performance—capturing some non-linear patterns better than Logistic Regression while being simpler than Boosting.
+- Its AUC and Recall typically fall between LogReg and Boosting, making it a balanced, moderately interpretable model.
+- A small comparison table helps show where KNN stands relative to the other approaches.
 
-```{r, echo=FALSE, results='asis'}
+---
+
+```{r, echo=FALSE}
 library(pROC)
@@
 ---
-# 3. Boosting Model (High Predictive Benchmark)
+# 3. Boosting Model (High Predictive Benchmark) (t)
 ---
@@
 ## Boosting 1/3: Model Formulation
-* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
-* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
-* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
-```{r, echo=FALSE, results='asis'}
+- Boosting builds many small trees sequentially, each fixing errors from the previous one.
+- These weak learners combine into a strong model that captures complex, non-linear patterns.
+- It usually achieves the highest accuracy/AUC, but is less interpretable than LogReg or KNN.
+- A diagram of small trees merging into one strong model illustrates this process.
+
+---
+
+```{r, echo=FALSE}
 library(gbm)
@@
 ## Boosting 2/3: Hyperparameter Tuning 
-* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
-* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
-* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
-```{r, echo=FALSE, results='asis'}
+- Boosting requires tuning several hyperparameters, including the number of trees (M), interaction depth, and learning rate (η).
+- Cross-Validation helps find the right balance so the model is flexible enough to learn patterns but not overfit.
+- A plot of Test Error vs. Number of Trees shows where performance stops improving and begins to overfit.
+
+---
+
+```{r, echo=FALSE}
 boost_model <- gbm(
@@
 ## Boosting 3/3: Evaluation & Importance 
-* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
-* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
-* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
-```{r, echo=FALSE, results='asis'}
+- The final Boosting model achieved the highest AUC, making it the strongest predictive model among those tested.
+- Because Boosting lacks interpretable coefficients, we rely on a Variable Importance Plot to understand which features influence predictions most.
+- The importance rankings generally align with the key predictors identified by Logistic Regression, offering a post-hoc view into the model’s behavior.
+
+---
+
+```{r, echo=FALSE}
 library(pROC)

### 2025-12-08 Slides.Rmd
Update slides with cleaned code and final results

--- 2025-12-08 Slides.Rmd
@@
   ioslides_presentation:
-    css: styles.css
-
+    widescreen: true
 ---
@@
 
-## Project Motivation & Business Problem
-* **Discuss:** The bank's need for efficient marketing and the high cost of telemarketing campaigns. The goal is to maximize the success rate by predicting customer subscription.
-* **Discuss:** The challenge of less positive class (imbalanced data) and why prediction accuracy is crucial.
-* **Image/Plot Idea:** A simple, high-level info graphic of a marketing campaign funnel, highlighting the bottleneck/drop-off point.
+## Project Motivation & Business Problem (S)
+- Telemarketing campaigns are expensive, time-consuming, and have very low success rates. The bank wants to improve efficiency by identifying which clients are most likely to subscribe.
+- The dataset has a severe class imbalance: only a small fraction of customers subscribe to the term deposit. This makes prediction difficult and increases the risk of false negatives.
+- Improving predictive accuracy would allow the bank to target fewer customers, reduce costs, and increase return on investment (ROI) for future campaigns.
 
-[Image of a marketing funnel illustrating customer conversion]
+--- 
+```{r, echo=FALSE}
+library(ggplot2)
+
+ggplot(bank_full, aes(x = y, fill = y)) +
+geom_bar() +
+scale_fill_manual(values = c("#2E86C1", "#F39C12")) +
+labs(
+title = "Term Deposit Subscription (Severe Class Imbalance)",
+x = "Subscribed?",
+y = "Count"
+) +
+theme_minimal(base_size = 14)
+
+```
 
@@
 
+* Cell phone calls perform best
+
 ```{r eda1-plot, echo=FALSE, fig.align="center", out.width="90%"}
@@
 
+* Past success predicts new success
+
 ```{r eda2-plot, echo=FALSE, fig.align="center", out.width="90%"}
@@
 ## EDA Finding 3: Too Many Calls Hurts Results
+
+* More calls = worse results
 
@@
 ## KNN 1/3: Feature Preparation 
-* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
-* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
-* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
-```{r, echo=FALSE, results='asis'}
+* **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
+* The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
 
+---
+
+```{r knn-prep, echo=FALSE, message=FALSE, warning=FALSE}
 library(dplyr)
@@
 
-set.seed(123)
-train_index <- sample(nrow(bank_full), 0.8 * nrow(bank_full))
-bank_train <- bank_full[train_index, ]
-bank_test  <- bank_full[-train_index, ]
+# Use the same train/test split from appendix
+numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]
+numeric_vars <- setdiff(numeric_vars, c("y_num", "residuals"))
 
-numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
-numeric_vars <- setdiff(numeric_vars, "y")
-
-train_x <- bank_train[, numeric_vars, drop = FALSE]
-test_x  <- bank_test[, numeric_vars, drop = FALSE]
-
-train_x <- bank_train[, numeric_vars]
-test_x  <- bank_test[, numeric_vars]
+train_x <- train_data[, numeric_vars]
+test_x  <- test_data[, numeric_vars]
 
@@
 
-train_y <- bank_train$y
-test_y  <- bank_test$y
+train_y <- train_data$y
+test_y  <- test_data$y
 
-best_k <- 5
-
+# Visualization
 par(mfrow = c(1, 2))
-
 plot(
-  bank_full$age, bank_full$balance,
+  train_data$age, train_data$balance,
   xlab = "Age", ylab = "Balance",
@@
 
-scaled_two <- scale(bank_full[, c("age", "balance")])
-
+scaled_two <- scale(train_data[, c("age", "balance")])
 plot(
@@
   xlab = "Scaled Age", ylab = "Scaled Balance",
-  main = "Scaled Features (mean = 0, sd = 1)"
+  main = "Scaled Features"
 )
-
 par(mfrow = c(1, 1))
 
-i <- 1  
-test_point <- test_x_scaled[i, , drop = FALSE]
-
-distances <- apply(
-  train_x_scaled,
-  1,
-  function(row) sqrt(sum((row - test_point)^2))
-)
-
-nn_idx <- order(distances)[1:best_k]
-
-knn_demo_pred <- knn(
-  train = train_x_scaled,
-  test  = test_point,
-  cl    = train_y,
-  k     = best_k
-)
-
-cat("Test customer index:", i, "\n")
-cat("KNN predicted class:", knn_demo_pred, "\n")
-cat("True class:", test_y[i], "\n\n")
-
-cat("Neighbor outcome breakdown:\n")
-print(table(train_y[nn_idx]))
-
-cat("\nNearest neighbor details:\n")
-print(head(bank_train[nn_idx, c("age", "balance", "duration", "campaign", "y")]))
-
-```
-```{r, echo=FALSE, results='asis'}
-library(class)
-
-numeric_vars <- names(bank_full)[sapply(bank_full, is.numeric)]
-numeric_vars <- setdiff(numeric_vars, "y")
-
-train_x <- bank_train[, numeric_vars, drop = FALSE]
-test_x  <- bank_test[, numeric_vars, drop = FALSE]
-
-train_x_scaled <- scale(train_x)
-test_x_scaled <- scale(
-  test_x,
-  center = attr(train_x_scaled, "scaled:center"),
-  scale  = attr(train_x_scaled, "scaled:scale")
-)
-
-train_y <- bank_train$y
-test_y  <- bank_test$y
-
-knn_pred <- knn(
-  train = train_x_scaled,
-  test  = test_x_scaled,
-  cl    = train_y,
-  k     = 5
-)
-
-knn_accuracy <- mean(knn_pred == test_y)
-knn_error <- 1 - knn_accuracy
-cat("KNN Test Accuracy:", round(knn_accuracy * 100, 2), "%\n")
-cat("KNN Test Error:", round(knn_error * 100, 2), "%\n")
-
-```
+# Calculate accuracy with k=5 for demonstration
+knn_pred_demo <- knn(train = train_x_scaled, test = test_x_scaled, cl = train_y, k = 5)
+knn_accuracy_demo <- mean(knn_pred_demo == test_y)
 ```
 
+**Test Accuracy (k=5):** `r round(knn_accuracy_demo * 100, 2)`%
+---
+
 ## KNN 2/3: Hyperparameter Tuning 
-* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
-* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
-* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
-```{r, echo=FALSE, results='asis'}
-library(class)
-library(caret)
+* The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
+* **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
 
-set.seed(123)
+---
 
-k_values <- 1:25                  
+```{r knn-tuning, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
+set.seed(445)
+
+k_values <- seq(1, 25, by = 2)
 cv_error <- numeric(length(k_values))
@@
 best_k <- k_values[which.min(cv_error)]
-best_k
+
 plot(
@@
   ylab = "5-fold CV Test Error",
-  main = "KNN: 5-fold CV Test Error vs k"
+  main = "KNN: CV Test Error vs k"
 )
-
-abline(v = best_k, lty = 2)
-
+abline(v = best_k, lty = 2, col = "red")
+text(best_k, max(cv_error), paste("Best k =", best_k), pos = 4, col = "red")
 ```
 
+**Optimal k:** `r best_k`
+--- 
 ## KNN 3/3: Evaluation & Test Performance
-* **Discuss:** The final KNN model (with optimal $k$) was applied to the **Test Set**. Evaluate performance using **AUC** and compare it to the LogReg models.
-* **Discuss:** KNN generally provides a good intermediate balance—often capturing non-linearity better than LogReg but being less computationally expensive than boosting.
-* **Image/Plot Idea:** A brief table comparing the **AUC** and **Recall** of the best LogReg model and the final KNN model.
+* The final KNN model (with optimal $k$) was applied to the **Test Set**. We evaluate performance using **AUC** and compare it to the LogReg models.
+* KNN generally provides a good intermediate balance—often capturing non-linearity better than LogReg but being less computationally expensive than boosting.
 
-```{r, echo=FALSE, results='asis'}
+---
+
+```{r knn-results, echo=FALSE, message=FALSE, warning=FALSE}
 library(pROC)
-library(caret)
-
-set.seed(123)
 
@@
 
-logit_model <- glm(y ~ ., data = bank_train, family = binomial)
+cm_knn <- knn_cm
+roc_knn <- knn_roc
+auc_knn <- knn_auc
 
-logit_prob <- predict(logit_model, newdata = bank_test, type = "response")
-logit_pred <- ifelse(logit_prob >= 0.5, "yes", "no")
-logit_pred <- factor(logit_pred, levels = c("no", "yes"))
-
-logit_roc <- roc(response = test_y_fac, predictor = logit_prob, levels = c("no", "yes"))
-logit_auc <- as.numeric(auc(logit_roc))
-
-logit_cm     <- confusionMatrix(logit_pred, test_y_fac, positive = "yes")
-logit_recall <- logit_cm$byClass["Recall"]
-
+# Results table
 results <- data.frame(
-  Model  = c("Logistic Regression", paste0("KNN (k = ", best_k, ")")),
-  AUC    = c(logit_auc, knn_auc),
-  Recall = c(as.numeric(logit_recall), as.numeric(knn_recall))
+  Model  = c("KNN", "Baseline Logit", "Ridge", "LASSO"),
+  AUC    = c(knn_auc, auc(roc_baseline), auc(roc_ridge), auc(roc_lasso)),
+  Sensitivity = c(as.numeric(knn_recall), 
+                  cm_baseline$byClass["Sensitivity"],
+                  cm_ridge$byClass["Sensitivity"],
+                  cm_lasso$byClass["Sensitivity"])
 )
@@
 results$AUC <- round(results$AUC, 3)
-results$Recall <- round(results$Recall, 3)
+results$Sensitivity <- round(results$Sensitivity, 3)
 
-results
-
+knitr::kable(results)
 ```
@@
 ---
+```{r boosting-setup, echo=FALSE, message=FALSE, warning=FALSE}
+library(gbm)
+library(pROC)
+library(caret)
 
-## Boosting 1/3: Model Formulation
-* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
-* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
-* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
-```{r, echo=FALSE, results='asis'}
-library(gbm)
+# Clean up train_data if needed
+if ("residuals" %in% colnames(train_data)) {
+  train_data$residuals <- NULL
+}
 
-bank_train$y_num <- ifelse(bank_train$y == "yes", 1, 0)
-bank_test$y_num  <- ifelse(bank_test$y  == "yes", 1, 0)
+# Create numeric response for boosting
+train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
+test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)
 
-set.seed(123)
+set.seed(445)
 
+# Fit boosting model
 boost_model <- gbm(
-  formula = y_num ~ . - y,       
-  data = bank_train,
+  y_num ~ . - y,         
+  data = train_data,
   distribution = "bernoulli",
-  n.trees = 1000,                
+  n.trees = 1500,       
+  shrinkage = 0.01,
   interaction.depth = 3,
-  shrinkage = 0.01,             
-  n.minobsinnode = 10,
+  bag.fraction = 0.5,
   verbose = FALSE
@@
 
-boost_prob <- predict(
-  boost_model,
-  newdata = bank_test,
-  n.trees = 1000,
-  type = "response"
-)
+best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = FALSE)
 
-boost_pred <- ifelse(boost_prob > 0.5, "yes", "no")
+# Predictions
+boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
+boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
 
-boost_accuracy <- mean(boost_pred == bank_test$y)
-boost_error    <- 1 - boost_accuracy
+# Confusion matrix
+cm_boost <- confusionMatrix(boost_pred, test_data$y, positive = "yes")
 
-cat("Boosting Test Accuracy:", round(boost_accuracy * 100, 2), "%\n")
-cat("Boosting Test Error:",   round(boost_error * 100, 2), "%\n")
+# ROC and AUC
+roc_boost <- roc(test_data$y, boost_prob)
+auc_boost <- auc(roc_boost)
+```
 
+## Boosting 1/3: Model Formulation
+
+- Boosting builds many small trees sequentially, each fixing errors from the previous one.
+- These weak learners combine into a strong model that captures complex, non-linear patterns.
+- It usually achieves the highest accuracy/AUC, but is less interpretable than LogReg or KNN.
+```{r boosting-viz, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
 plot(
@@
   ylab = "Training Loss",
-  main = "Boosting: Sequential Error Reduction"
+  main = "Boosting: Sequential Error Reduction",
+  col = "steelblue",
+  lwd = 2
 )
-
 ```
@@
 ## Boosting 2/3: Hyperparameter Tuning 
-* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
-* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
-* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
-```{r, echo=FALSE, results='asis'}
-boost_model <- gbm(
-  formula = y_num ~ . - y,
-  data = bank_train,
-  distribution = "bernoulli",
-  n.trees = 1000,
-  interaction.depth = 3,
-  shrinkage = 0.01,
-  n.minobsinnode = 10,
-  verbose = FALSE
-)
-
-M_values <- seq(50, 1000, by = 50)
+- We tuned the number of trees using OOB error estimation to prevent overfitting.
+- The OOB method selected **`r best_trees`** trees as optimal.
+- This choice balances model complexity and generalization.
+```{r boosting-tuning, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
+# Test different numbers of trees
+M_values <- seq(50, 1500, by = 50)
 test_error <- numeric(length(M_values))
@@
     boost_model,
-    newdata = bank_test,
+    newdata = test_data,
     n.trees = M,
@@
   pred_M <- ifelse(prob_M > 0.5, "yes", "no")
-  test_error[i] <- mean(pred_M != bank_test$y)
+  test_error[i] <- mean(pred_M != test_data$y)
 }
@@
 best_M   <- M_values[best_idx]
-
-cat("Best number of trees based on Test Error:", best_M, "\n")
 
@@
   ylab = "Test Error",
-  main = "Boosting: Test Error vs Number of Trees"
+  main = "Boosting: Test Error vs Number of Trees",
+  col = "steelblue",
+  lwd = 2
 )
-
-abline(v = best_M, lty = 2)
-
+abline(v = best_M, lty = 2, col = "red", lwd = 2)
+text(best_M, max(test_error), paste("Best M =", best_M), pos = 4, col = "red")
 ```
 
+**Optimal number of trees (via test error):** `r best_M`
+
 ## Boosting 3/3: Evaluation & Importance 
-* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
-* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
-* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
-```{r, echo=FALSE, results='asis'}
-library(pROC)
-
-boost_prob <- predict(
-  boost_model,
-  newdata = bank_test,
-  n.trees = 1000,       
-  type = "response"
-)
-
-test_y_fac <- factor(bank_test$y, levels = c("no", "yes"))
-
-boost_roc <- roc(
-  response  = test_y_fac,
-  predictor = boost_prob,
-  levels    = c("no", "yes")
-)
-
-boost_auc <- as.numeric(auc(boost_roc))
-cat("Boosting Test AUC:", round(boost_auc, 3), "\n")
-
+- The final Boosting model achieved the highest AUC and sensitivity among all models tested.
+- Variable Importance Plot shows which features influence predictions most.
+- The importance rankings align with key predictors identified by Logistic Regression.
+```{r boosting-results, echo=FALSE, message=FALSE, warning=FALSE}
+# Variable importance
 vi <- summary(boost_model, plotit = FALSE)
-
 top_n <- min(10, nrow(vi))
@@
   ylab = "Relative Influence",
-  main = "Boosting: Variable Importance (Top Features)"
+  main = "Boosting: Variable Importance (Top 10 Features)",
+  col = "steelblue"
+)
+```
+
+**Boosting Test AUC:** `r round(auc_boost, 3)`  
+**Boosting Sensitivity:** `r round(cm_boost$byClass["Sensitivity"], 3)`  
+**Boosting Accuracy:** `r round(cm_boost$overall["Accuracy"], 3)`
+
+## Results: Comparative Evaluation
+- We evaluated five models using Accuracy, Sensitivity (Recall), and AUC on the Test Set.
+- **Boosting** achieved the highest overall accuracy and the highest sensitivity, meaning it identified the most true positives.
+- Logistic, Ridge, and LASSO models performed very similarly in accuracy and AUC, but with lower sensitivity.
+- KNN performed the worst, with the lowest AUC, indicating weak separation between classes.
+
+---
+
+## Model Performance Comparison
+```{r final-table, echo=FALSE, message=FALSE, warning=FALSE}
+# Create aliases for consistency
+cm_knn <- knn_cm
+roc_knn <- knn_roc
+
+final_results <- data.frame(
+  Model = c("Logistic", "Ridge", "LASSO", "KNN", "Boosting"),
+  Accuracy = c(
+    cm_baseline$overall["Accuracy"],
+    cm_ridge$overall["Accuracy"],
+    cm_lasso$overall["Accuracy"],
+    cm_knn$overall["Accuracy"],
+    cm_boost$overall["Accuracy"]
+  ),
+  Sensitivity = c(
+    cm_baseline$byClass["Sensitivity"],
+    cm_ridge$byClass["Sensitivity"],
+    cm_lasso$byClass["Sensitivity"],
+    cm_knn$byClass["Sensitivity"],
+    cm_boost$byClass["Sensitivity"]
+  ),
+  AUC = c(
+    auc(roc_baseline),
+    auc(roc_ridge),
+    auc(roc_lasso),
+    auc_knn,
+    auc_boost
+  )
 )
 
+knitr::kable(final_results, digits = 3)
 ```
 
-## Results: Comparative Evaluation (Slide 13). #MUSSA
-* **Discuss:** Present the **Final Summary Table** comparing the Test Set performance (AUC, Recall, Accuracy) of all models (**Best LogReg**, **KNN**, **Boosting**).
-* **Discuss:** Visually compare the models using the **Combined ROC Curve Plot** and discuss which model dominates the space.
-* **Image/Plot Idea:** The **Combined ROC Curve Plot** comparing all three final models (LogReg, KNN, and Boosting).
+---
 
-## The Trade-Off / Business Impact  #MUSSA
-* **Discuss:** Reiterate the central trade-off: **Boosting** is the best *predictor*, but the **Regularized Logistic Regression** offers the most *interpretable* and *actionable* insights for the bank.
-* **Discuss:** Final actionable advice for the bank: use **LogReg insights** to design campaigns and **Boosting** for a final scoring model if maximum predictive accuracy is needed.
+## ROC Curve Comparison
+```{r final-roc, echo=FALSE, message=FALSE, warning=FALSE}
+roc_all <- rbind(
+  data.frame(fpr = 1 - roc_baseline$specificities, tpr = roc_baseline$sensitivities, Model = "Logistic"),
+  data.frame(fpr = 1 - roc_ridge$specificities,    tpr = roc_ridge$sensitivities,    Model = "Ridge"),
+  data.frame(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    Model = "LASSO"),
+  data.frame(fpr = 1 - roc_knn$specificities,      tpr = roc_knn$sensitivities,      Model = "KNN"),
+  data.frame(fpr = 1 - roc_boost$specificities,    tpr = roc_boost$sensitivities,    Model = "Boosting")
+)
+
+ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(linewidth = 1) +
+  geom_abline(linetype = "dashed", color = "grey") +
+  labs(
+    title = "ROC Curve Comparison for All Models",
+    x = "False Positive Rate",
+    y = "True Positive Rate"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "right")
+```
+
+## Business Impact
+* The central trade-off: Boosting is the best predictor, but the **Regularized Logistic Regression** offers the most *interpretable* and *actionable* insights for the bank.
+* Sample coefficients: $\beta_{Balance^2}$ = -23.02, $\beta_{Housing}$ = -0.65, $\beta_{JobStudent}$ = 4.064
+* The chosen model can lead to cost savings and higher ROI for future campaigns.
 * **Image/Plot Idea:** A visual metaphor (e.g., a scale or seesaw) contrasting "High Performance" vs. "High Interpretability."
-* **Discuss:** How the **chosen model** (Balancing performance and interpretability) can lead to **cost savings** and **higher ROI** for future campaigns.
-* **Discuss:** Briefly touch upon **regulatory/ethical implications** in banking and the need to justify predictions (further supporting the value of interpretability).
-* **Image/Plot Idea:** A currency symbol or a bar chart showing the conceptual increase in ROI due to better targeting.
 
-## Discussion: Limitations & Future Work (Slide 14)  #MUSSA
-* **Discuss:** Acknowledge limitations, such as not using **time-series analysis** or fully accounting for the dependency of the 'duration' feature.
-* **Discuss:** Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying **Deep Learning** models.
+## Discussion
+* Consideration: regulatory/ethical implications in banking and the need to justify predictions (further supporting the value of interpretability).
+* Advanced train/test splitting methods like **time-series analysis** were not used.
+* Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying.
 
-## References (Slide 15)
-* **Discuss:** Simply list the required references for the data source and any major methodology papers.
-* **Discuss:** Provide contact information for questions.
-* Thank you
-* **Image/Plot Idea:** The official UCI ML Repository logo and your contact email.
+## References
+* UC Irvine Machine Learning Repository. Bank Marketing. https://archive.ics.uci.edu/dataset/222/bank+marketing 
+* Github link: https://github.com/dsci445-csu/project-6
+
+```{r image, echo=FALSE, fig.align='center'}
+knitr::include_graphics("UCI.png")
+```

### 2025-12-08 Slides.Rmd
correct path

@@
 ```{r slide2-plot, echo=FALSE, fig.align="center", out.width="90%"}
-knitr::include_graphics("imbalance_plot.png")
+knitr::include_graphics("EDA/imbalance_plot.png")
 ```
@@
 ```{r eda1-plot, echo=FALSE, fig.align="center", out.width="90%"}
-knitr::include_graphics("contact_plot.png")
+knitr::include_graphics("EDA/contact_plot.png")
 
@@
 ```{r eda2-plot, echo=FALSE, fig.align="center", out.width="90%"}
-knitr::include_graphics("poutcome_plot.png")
+knitr::include_graphics("EDA/poutcome_plot.png")
 ```
@@
 ```{r eda3-plot, echo=FALSE, fig.align="center", out.width="90%"}
-knitr::include_graphics("campaign_plot.png")
+knitr::include_graphics("EDA/campaign_plot.png")
 ```

### 2025-12-08 Slides.Rmd
slides update4

@@
 author: "Sandra, Tigist, Mussa Hassen, Waddah"
-subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
-  Kaplan'
+subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr. Andee Kaplan'
 output:
@@
 ---
+
 ```{r, echo=FALSE}
@@
 
-* **Most clients say NO; only ≈11% say YES**
+* **Most clients say NO; only about 11% say YES **
 
@@
 
-## Methodology: Data Preparation (Slide 5) (s)
+## Methodology: Data Preparation (s)
 
@@
 
-## LogReg: Feature Selection & Linearity  #MUSSA
+## LogReg: Feature Selection & Linearity  
 * Initial feature selection was performed using **Backward Stepwise Elimination** on the full model to find the most useful set of main effects.
@@
 
-$y \sim job + marital + education + balance + housing + $
+$y \sim job + marital + education + balance + housing + loan +$
 
-$loan + contact + day + month + duration + $
+$contact + day + month + duration + campaign + previous+ poutcome$
 
-$campaign + previous+ poutcome$
 * **Residual Plots** were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.
@@
 
-## LogReg: Regularization & Tuning   #MUSSA
+## LogReg: Regularization & Tuning  
 * We explored Ridge and LASSO regularization as an alternative approach for feature selection.
@@
 
-```{r, echo=FALSE, results='asis', fig.width=4, fig.height=3}
+```{r, echo=FALSE, results='asis', fig.width=5, fig.height=3.5}
 
@@
 
-## LogReg: Evaluation & Coefficients  #MUSSA
+## LogReg: Evaluation & Coefficients
 * The final selected LogReg model was evaluated on the 5-fold CV error rate, along with metrics focused on the confusion matrix.
@@
 ---
+
 # 2. K-Nearest Neighbors (KNN) Model
+
 ---
@@
 **Test Accuracy (k=5):** `r round(knn_accuracy_demo * 100, 2)`%
+
 ---
@@
 **Optimal k:** `r best_k`
+
 ---
+
 ## KNN 3/3: Evaluation & Test Performance
@@
 ---
+
 # 3. Boosting Model (High Predictive Benchmark)
+
 ---
+
 ```{r boosting-setup, echo=FALSE, message=FALSE, warning=FALSE}

### 2025-12-09 Slides.Rmd
Update slides

--- 2025-12-08 Slides.Rmd
@@
 
----
-
 ## Breakdown of Categorical Variable
@@
 
----
 # 2. K-Nearest Neighbors (KNN) Model
----
 
@@
 **Test Accuracy (k=5):** `r round(knn_accuracy_demo * 100, 2)`%
----
 
@@
 **Optimal k:** `r best_k`
---- 
+
 ## KNN 3/3: Evaluation & Test Performance
@@
 
----
 # 3. Boosting Model (High Predictive Benchmark)
----
+
 ```{r boosting-setup, echo=FALSE, message=FALSE, warning=FALSE}
@@
 - It usually achieves the highest accuracy/AUC, but is less interpretable than LogReg or KNN.
+
+---
+
 ```{r boosting-viz, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
@@
 - This choice balances model complexity and generalization.
+---
+
 ```{r boosting-tuning, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
@@
 - The importance rankings align with key predictors identified by Logistic Regression.
+
+---
+
 ```{r boosting-results, echo=FALSE, message=FALSE, warning=FALSE}
@@
 - KNN performed the worst, with the lowest AUC, indicating weak separation between classes.
-
----
 
@@
 ```
-
----
 

### 2025-12-09 Slides.Rmd
Spaces
summary slides

@@
 
-## LogReg: Feature Selection & Linearity  #MUSSA
+# 1. Logistic Regression Model
+
+## Feature Selection & Linearity  
 * Initial feature selection was performed using **Backward Stepwise Elimination** on the full model to find the most useful set of main effects.
-* Our selected formula resutled:
+* Our selected formula resulted:
 
-$y \sim job + marital + education + balance + housing + $
+$y \sim job + marital + education + balance + housing + loan + contact +$
 
-$loan + contact + day + month + duration + $
+$day + month + duration + campaign + previous+ poutcome$
 
-$campaign + previous+ poutcome$
-* **Residual Plots** were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.
+* Residual Plots were used to assess the linearity assumption for numeric predictors (`age`, `balance`). This led to testing a model with non-linear terms.
 
@@
 
-## LogReg: Models with non-linear components and interactions
+## Models with non-linear components and interactions
 * A **quadratic term** $I(\text{balance}^4)$—was added to the formula to reduce the model's bias.
 * We then tested for crucial **interaction terms** (between campaign features and demographics). 
-* Our model had a cross-validated test error rate 0.09846
+* Our model had a cross-validated test error rate `r round(round(cv_error_rate, 5), 4)``
 
-## LogReg: Regularization & Tuning   #MUSSA
+## Regularization & Tuning  
 * We explored Ridge and LASSO regularization as an alternative approach for feature selection.
@@
 
-```{r, echo=FALSE, results='asis', fig.width=4, fig.height=3}
-
+```{r, echo=FALSE, results='asis', fig.width=5, fig.height=4}
 plot(cv.ridge) 
@@
 
-## LogReg: Evaluation & Coefficients  #MUSSA
+## Evaluation 
 * The final selected LogReg model was evaluated on the 5-fold CV error rate, along with metrics focused on the confusion matrix.
-* Models performed relatively similar. We would prefer a model with higher **sensitivity** (few false negatives)
+* Models performed relatively similar. We would prefer a model with higher *sensitivity* (few false negatives)
 
@@
 - This choice balances model complexity and generalization.
+
 ---
@@
 * Consideration: regulatory/ethical implications in banking and the need to justify predictions (further supporting the value of interpretability).
-* Advanced train/test splitting methods like **time-series analysis** were not used.
-* Suggest future improvements: Exploring advanced techniques like **resampling methods** (SMOTE) to handle imbalance better or trying.
+* Advanced train/test splitting methods like *time-series analysis* were not used.
+* Suggest future improvements: Exploring advanced techniques like *resampling methods* (SMOTE) to handle imbalance better or trying.
 

### 2025-12-09 Slides.Rmd
slides update6

@@
 ---
-title: "Statistical Machine Learning Models on A Bank Marketing dataset"
+title: "Statistical Machine Learning Models on A Bank Marketing Dataset"
 author: "Sandra, Tigist, Mussa Hassen, Waddah"
@@
 
-```{r slide2-plot, echo=FALSE, fig.align="center", out.width="90%"}
+```{r slide2-plot, echo=FALSE, fig.align="center", out.width="75%"}
 knitr::include_graphics("EDA/imbalance_plot.png")
@@
 
-```{r eda1-plot, echo=FALSE, fig.align="center", out.width="90%"}
+```{r eda1-plot, echo=FALSE, fig.align="center", out.width="75%"}
 knitr::include_graphics("EDA/contact_plot.png")
-
 ```
@@
 
-```{r eda2-plot, echo=FALSE, fig.align="center", out.width="90%"}
+```{r eda2-plot, echo=FALSE, fig.align="center", out.width="68%"}
 knitr::include_graphics("EDA/poutcome_plot.png")
@@
 
-```{r eda3-plot, echo=FALSE, fig.align="center", out.width="90%"}
+```{r eda3-plot, echo=FALSE, fig.align="center", out.width="75%"}
 knitr::include_graphics("EDA/campaign_plot.png")
 ```
-
-## Problem Statement & Methodology Overview
-
-* **Goal:** Compare models on two criteria:  
-  * **Predictive Power** – ability to correctly find YES cases  
-  * **Interpretability** – model is easy to understand
-
-* **Model evaluation:**  
-  * **80/20 Train/Test Split**  
-  * **5-Fold Cross-Validation**
 
@@
 * The chosen model can lead to cost savings and higher ROI for future campaigns.
-* **Image/Plot Idea:** A visual metaphor (e.g., a scale or seesaw) contrasting "High Performance" vs. "High Interpretability."
 

### 2025-12-09 Slides.Rmd
My changes to slides

--- 2025-12-09 Slides.Rmd
@@
 title: "Statistical Machine Learning Models on A Bank Marketing dataset"
-author: "Sandra, Tigist, Mussa Hassen, Waddah"
+author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah"
 subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee

### 2025-12-09 Slides.Rmd
final slides update

--- 2025-12-09 Slides.Rmd
@@
 title: "Statistical Machine Learning Models on A Bank Marketing Dataset"
-author: "Sandra, Tigist, Mussa Hassen, Waddah"
+author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah"
 subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
@@
 
-## Project Motivation & Business Problem (S)
+## Project Motivation & Business Problem
 - Telemarketing campaigns are expensive, time-consuming, and have very low success rates. The bank wants to improve efficiency by identifying which clients are most likely to subscribe.
@@
 - Improving predictive accuracy would allow the bank to target fewer customers, reduce costs, and increase return on investment (ROI) for future campaigns.
-
----
-
-```{r, echo=FALSE}
-library(ggplot2)
-
-ggplot(bank_full, aes(x = y, fill = y)) +
-geom_bar() +
-scale_fill_manual(values = c("#2E86C1", "#F39C12")) +
-labs(
-title = "Term Deposit Subscription (Severe Class Imbalance)",
-x = "Subscribed?",
-y = "Count"
-) +
-theme_minimal(base_size = 14)
-
-```
 
@@
 
-## Methodology: Data Preparation (Slide 5) (s)
+## Methodology: Data Preparation
 
@@
 - We tuned the number of trees using OOB error estimation to prevent overfitting.
-- The OOB method selected **`r best_trees`** trees as optimal.
+- The OOB method selected 1100 trees as optimal.
 - This choice balances model complexity and generalization.

### 2025-12-09 Slides.Rmd
final slides title without numbers

@@
 
-## KNN 1/3: Feature Preparation 
+## KNN: Feature Preparation 
 * **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
@@
 
-## KNN 2/3: Hyperparameter Tuning 
+## KNN: Hyperparameter Tuning 
 * The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
@@
 
-## KNN 3/3: Evaluation & Test Performance
+## KNN: Evaluation & Test Performance
 * The final KNN model (with optimal $k$) was applied to the **Test Set**. We evaluate performance using **AUC** and compare it to the LogReg models.
@@
 
-## Boosting 1/3: Model Formulation
+## Boosting: Model Formulation
 
@@
 
-## Boosting 2/3: Hyperparameter Tuning 
+## Boosting: Hyperparameter Tuning 
 - We tuned the number of trees using OOB error estimation to prevent overfitting.
@@
 
-## Boosting 3/3: Evaluation & Importance 
+## Boosting: Evaluation & Importance 
 - The final Boosting model achieved the highest AUC and sensitivity among all models tested.

### 2025-12-09 Slides.Rmd
slides update7

@@
 title: "Statistical Machine Learning Models on A Bank Marketing Dataset"
-author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah"
+author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah Alaahtani"
 subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
@@
 
-```{r residuals_slide, echo=FALSE, results='asis'}
+```{r residuals_slide, echo=FALSE, fig.align="center", results='asis'}
 residuals <- residuals(glm.baseline, type = "deviance")
@@
 
-```{r knn-prep, echo=FALSE, message=FALSE, warning=FALSE}
+```{r knn-prep, echo=FALSE, message=FALSE, fig.align="center", warning=FALSE}
 library(dplyr)
@@
 
-```{r knn-tuning, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
+```{r knn-tuning, echo=FALSE, message=FALSE, fig.align="center", warning=FALSE, cache=TRUE}
 set.seed(445)
@@
 
----
-
-```{r knn-results, echo=FALSE, message=FALSE, warning=FALSE}
+```{r knn-results, echo=FALSE, message=FALSE, fig.align="center",  out.width="75%", warning=FALSE}
 library(pROC)
@@
 
-# 3. Boosting Model (High Predictive Benchmark)
+# 3. Boosting Tress Model
 
-```{r boosting-setup, echo=FALSE, message=FALSE, warning=FALSE}
+```{r boosting-setup, echo=FALSE, fig.align="center", message=FALSE, warning=FALSE}
 library(gbm)
@@
 
-```{r boosting-viz, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
+```{r boosting-viz, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", out.width="80%"}
 plot(
@@
 
-```{r boosting-tuning, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
+```{r boosting-tuning, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", out.width="75%", cache=TRUE}
 # Test different numbers of trees
@@
 ## ROC Curve Comparison
-```{r final-roc, echo=FALSE, message=FALSE, warning=FALSE}
+```{r final-roc, echo=FALSE, message=FALSE, fig.align="center", warning=FALSE}
 roc_all <- rbind(

### 2025-12-10 Paper.Rmd
Paper.Rmd

--- 2025-12-06 draft_group-6.Rmd
@@
 output:
-  pdf_document: default
   html_document:
     df_print: paged
+  pdf_document: default
 ---
-
-*Paper (40%)* A paper detailing the motivation, methodology, and results of your project, complete with references. (https://dsci445-csu.github.io/project/)
-
-# Outline
 

### 2025-12-10 Paper.Rmd
Added motivation, dataset description, problem statement, and KNN section

@@
 title: "draft_group-6"
-author: "Sandra, Tigist, Mussa, Waddah"
+author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alaahtani"
 output:
@@
 
-## Motivation
-(13% of the content)\
-The Bank Marketing dataset contains information collected from a Portuguese banking institution during multiple direct telemarketing campaigns (source1).
-*Why??* (We want to justify the efforts of modeling this data, then whwy it needs to be accurate)
+## Motivation # TT
+Bank Marketing dataset The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns.
+These campaigns were done to encourage clients to subscribe to a long-term deposit product.
+Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
+Predictive modeling bears great value in this scenario, as it helps the bank focus resources on only those customers who depict a higher possibility of subscribing.
+In the subscription rate of this dataset, the problem is even more difficult as it is very low.
+Most individuals say "no," and only a small percentage decides to open a term deposit.
+If one class is rare, human judgment or simple rules fail to identify those clients who could actually be interested.
+A well-constructed model may present potential improvements in targeting accuracy, reduce the number of unsuccessful calls, and raise the return on investment.
+That is why it is necessary to research this dataset and look at which modeling approaches work best.
 
-**Description of dataset** \
-Each observation in the dataset corresponds to a single client contacted during the bank’s telemarketing campaigns. The dataset includes 16 predictor variables, which cover:
+**Description of dataset**  #TT 
+Each entry of the dataset represents one client contacted in one of the bank's telemarketing campaigns.
+This dataset contains 16 predictors carrying significant information about the client's background, financial situation, and interactions with previous campaigns.
+These variables fall into several key categories:
+Demographics include such factors as age, occupation, marital status, and education. These help describe who the clients are and often relate to financial behavior or economic stability.
+Financial indicators: Loan status, credit history, and average account balance.
+These variables describe the client's existing financial commitments and capacity, which may influence their willingness to invest in a long-term product like a term deposit.
+Marketing campaign features include the type of contact: cellular or telephone, month of contact, and day of contact, and duration of the call.
+These are operational decisions made by the marketing team and may influence a client's likelihood of subscribing.
+For example, longer call duration sometimes is indicative of higher client engagement.
+Campaign outcomes: Number of previous contacts and if the past interactions have resulted in a successful outcome.
+These variables highlight whether the client has been interested in the past or has been contacted many times without success.
+The data includes both numerical and categorical variables, and this can serve as a good example to compare the many different classification methods with respect to how each treats different data types: for instance, linear models require the conversion of categorical variables, whereas tree-based or distance-based methods would naturally handle those in different ways.
+This allows us to explore how each approach might respond to the same set of features.
+The response variable, y, tells whether the client has subscribed to the term deposit or not.
+Only a small fraction of the contacted clients actually subscribe, therefore this is a strongly imbalanced dataset.
+The class imbalance is because, in general, marketing datasets are imbalanced since the majority of customers do not respond positively when outreach is performed.
+This skewness biases models toward the "no" prediction since this prediction would naturally be correct for most of the cases.
+Therefore, careful handling of this bias will be critical to generate models capable of identifying the small minority of clients who are likely to subscribe-the group that the bank cares most about.
 
-- Demographics: age, occupation, marital status, education
-- Financial indicators: loan status, credit history, average account balance
-- Marketing campaign features: type of contact, month/day of contact, call duration
-- Past campaign outcomes: number of previous contacts and whether they were successful
-
-These variables include both categorical and quantitative features, making the dataset suitable for comparing a variety of classification methods.
-
-The target variable, y, is a binary indicator specifying whether the client subscribed to a term deposit (yes/no).
-As is typical in marketing data, the positive class (“yes”) is rare, meaning most clients do not subscribe (source 2).
-This creates a class imbalance, which can bias models toward predicting the majority class unless addressed appropriately.
-
-**Problem Statement/Assumptions** \
-The goal of this project is to build statistical models that can predict whether a client will subscribe to a term deposit based on their demographic, financial, and campaign-related characteristics.
-This is a binary classification problem.
-
-Our focus is on models that provide both good predictive performance and interpretability, since managers need results they can explain and translate into marketing strategies.
-Interpretability is especially valuable in domains such as banking, where decisions must be justified for ethical, customer-centric, and regulatory reasons.
-
-To keep the scope focused, we make several simplifying assumptions:
-
-- We treat all observations as independent, even though clients may appear in multiple campaigns.
-- We do not use time-series methods, even though campaign month/day could be part of a temporal structure.
-- We prioritize models covered in class, especially those that illustrate differences in linearity, flexibility, and underlying assumptions (e.g., covariance structure for LDA and QDA).
-- We assume the data used for modeling reflects the population targeted by the campaign, meaning the model can be applied to future clients with similar characteristics.
-
-These choices allow us to compare several classification approaches under a consistent framework.
+**Problem Statement/Assumptions**  #TT
+It is now required to develop statistical models based on the available information in the dataset that will predict with accuracy whether a client will subscribe to a term deposit.
+This problem is of a binary classification nature, and our work focuses on comparing several models that differ in assumptions, flexibility, and interpretability.
+Understanding how such models perform is useful to a marketing team for making informed decisions about which clients to contact in future campaigns.
+We focused our attention on well-performing models that give insightful knowledge of which factors drive the choice of a client.
+In the banking industry, understanding why a model has yielded some prediction is crucial.
+This supports ethical marketing practices, helps managers explain decisions, and can be important for regulatory compliance.
+For instance, highly flexible models may perform very well but may not indicate which client characteristics are actually relevant.
+More interpretable models do not always have the strongest accuracy but can guide marketing strategies more directly.
+We make the following assumptions to keep the analysis focused and manageable:
+Independence of observations: We treat the observations as independent, even though multiple contacts with the same client may appear across different campaigns.
+This assumption makes the training and evaluation of the models easier, even if some repeated-measure structure in the data is ignored.
+No time-series modeling: variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns.
+However, we do not model this data as a time series.
+This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.
+Use of class-relevant models: We restrict our approach to models presented in class because this enables us to contrast multiple concepts, including linear decision boundaries, distance-based approaches, and flexible tree-based structures.
+This renders the analysis suitable for a course project while still answering substantial questions regarding classification performance.
+Generalization from sample to population: We assume that this dataset is a good representation of the client population to which the bank will offer its services in forthcoming campaigns.
+If we couldn't assume this, our predictions wouldn't generalize properly to real-world decision-making.
+By selecting these assumptions, we are assured of a standard framework that will allow the comparison of different classification techniques on equal footing.
+The aim is not only to identify which model performs the best but also to understand how different modeling strategies behave when applied to imbalanced marketing data; this may provide practical insights into how a bank would structure its future campaigns and what features of clients are most relevant for predicting subscription behaviors.
 
@@
 It captures nonlinear relationships but can be sensitive to scaling and high-dimensional data.
+#TT
+
+K-Nearest Neighbors is a nonparametric classification method that predicts an outcome for a new observation by looking at the k most similar clients in the training dataset.
+Rather than estimating a set of coefficients or assuming any functional form between predictors and the response, KNN makes predictions based purely on proximity in predictor space.
+This makes the method particularly well-suited when the underlying relationship between predictors and subscription behavior is nonlinear or too complicated for a parametric model to capture.
+Because KNN relies on distance calculations, feature scaling is an important preprocessing step.
+Otherwise, predictors that are measured on a larger scale-e.g., account balance or call duration-would dominate the distance metric and distort which observations are considered "nearest." In our implementation, we standardized all numeric variables so that each feature was given equal weight in determining the similarity between any two customers.
+This preprocessing step is especially important for marketing datasets which often involve heterogeneous variables.
+A major modeling decision in KNN, which involves a choice, is the value of k.
+For small values of k, the classifier is highly flexible and closely follows the training data with a potential risk of overfitting noise.
+Larger values produce smoother, more stable decision boundaries; however, this has a potential risk of missing the important patterns.
+We used 5-fold cross-validation to identify the value of k that minimized the estimated test error, balancing the bias-variance trade-off.
+While KNN presents an intuitive method of finding clients with similar profiles, it has several limitations in this context.
+The performance may suffer for high-dimensional data, with points becoming further apart and distances becoming less informative-the "curse of dimensionality".
+Second, KNN is less interpretable than models like logistic regression, where the effect of every predictor can be expressed quantitatively.
+It does, however, provide a useful baseline when assessing the performance of alternative nonlinear classification methods for predicting term deposit subscriptions.
 

### 2025-12-11 Slides.Rmd
EDA slides

@@
 title: "Statistical Machine Learning Models on A Bank Marketing Dataset"
-author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah Alaahtani"
+author: "Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah Alqahtani"
 subtitle: 'Colorado State University, DSCI 445: Statistical Machine Learning, Dr.Andee
@@
 
-```{r slide2-plot, echo=FALSE, fig.align="center", out.width="75%"}
-knitr::include_graphics("EDA/imbalance_plot.png")
+```{r imbalance-plot, echo=FALSE, fig.align="center", out.width="75%"}
+ggplot(bank_full, aes(x = y, fill = y)) +
+  geom_bar(color = "white") +
+  scale_fill_manual(values = c("no" = "grey60", "yes" = "forestgreen")) +
+  labs(
+    title = "Distribution of Term Deposit Subscription",
+    x = "Subscribed?",
+    y = "Count"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(legend.position = "none")
 ```
@@
 
-```{r eda1-plot, echo=FALSE, fig.align="center", out.width="75%"}
-knitr::include_graphics("EDA/contact_plot.png")
+```{r contact-plot, echo=FALSE, fig.align="center", out.width="75%"}
+ggplot(bank_full, aes(x = contact, fill = y)) +
+  geom_bar(position = "fill", color = "white") +
+  scale_fill_manual(values = c("no" = "grey60", "yes" = "forestgreen")) +
+  scale_y_continuous(labels = scales::percent) +
+  labs(
+    title = "Subscription Rate by Contact Method",
+    x = "Contact Method",
+    y = "Proportion (Yes)"
+  ) +
+  theme_minimal(base_size = 14)
 ```
@@
 
-```{r eda2-plot, echo=FALSE, fig.align="center", out.width="68%"}
-knitr::include_graphics("EDA/poutcome_plot.png")
+```{r poutcome-plot, echo=FALSE, fig.align="center", out.width="70%"}
+ggplot(bank_full, aes(x = poutcome, fill = y)) +
+  geom_bar(position = "fill", color = "white") +
+  scale_fill_manual(values = c("no" = "grey60", "yes" = "forestgreen")) +
+  scale_y_continuous(labels = scales::percent) +
+  labs(
+    title = "Subscription Rate by Previous Outcome",
+    x = "Previous Outcome",
+    y = "Proportion (Yes)"
+  ) +
+  theme_minimal(base_size = 14)
 ```
@@
 
-```{r eda3-plot, echo=FALSE, fig.align="center", out.width="75%"}
-knitr::include_graphics("EDA/campaign_plot.png")
+```{r campaign-plot, echo=FALSE, fig.align="center", out.width="75%"}
+bank_full %>%
+  group_by(campaign) %>%
+  summarise(success_rate = mean(y == "yes")) %>%
+  ggplot(aes(x = campaign, y = success_rate)) +
+  geom_line(color = "steelblue", linewidth = 1.2) +
+  geom_point(color = "darkorange", size = 3) +
+  scale_y_continuous(labels = scales::percent) +
+  labs(
+    title = "Success Rate vs. Number of Campaign Contacts",
+    x = "Number of Contacts",
+    y = "Success Rate"
+  ) +
+  theme_minimal(base_size = 14)
 ```
@@
 
----
-
 ```{r knn-prep, echo=FALSE, message=FALSE, fig.align="center", warning=FALSE}
@@
 * **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
-
----
 

### 2025-12-13 Draft Folder/Intro.Rmd
Draft finalization

--- 2025-12-03 Intro.Rmd
@@
 
-Plots for categorical variables
+Plots for categorical variables (S.A.)
 
@@
 ```
+**Description:** These plots show that subscription rates vary meaningfully across categorical variables, with certain jobs, higher education levels, specific contact methods (especially cellular), particular months, and prior campaign successes all associated with higher proportions of clients subscribing to a term deposit.

### 2025-12-13 Draft2.Rmd
Draft finalization

---
title: "Draft2"
output: pdf_document
---

## Motivation
- Bank marketing dynamics (cite the paper)

Bank telemarketing campaigns are a major channel through which financial institutions promote term deposit products. Thsee marketing campaigns come with a cost and are often inefficient due to low client response rates. 
Prior work by Moro et al. (2014) discover that client characteristics and campaign related factors strongly influence subscription outcomes. Motivating a data-driven approach to improve targeting and decision-making in bank marketing. 
    In their approach a Portuguese retail bank was studied using campaign data collected between 2008 and 2013. \

- Introduce dataset
    - Description of variables (under factors(
    - Description of target variable (EDA of several factors with target; plots) 
    - note imbalance 
    - Include wrangling of dataset (as.factor() and addition of variable for pdays)

Our dataset was a publicily available daaset, thus had less variables for privacy concerns of the bank customers...
The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts. A list of the 16 variable found below...
The response variable indicates whether a client subscribed to a term deposit which defines a binary classification task.

Exploratory analysis shows that the numeric predictors (balance, duration, pdays, previously contacted) are heavily skewed.
Also, analysis reveals substantial class imbalance, in the response variable, which motivates the use of appropriate modeling and evaluation for our classification problem.
Visualizations of subscription rates by job education marital status contact type and previous outcome reveal variations in client responsiveness across different factors.

Data preprocessing involved converting categorical variables to factors and adding derived variables such as pdays to account for clients not previously contacted. Converting the categorical variables (initilay character type) to factors creates an indicator for each class in the variable (i.e. if job has 20 factors, 20 indicator variables will be considered in the model fit). As for the previously contaced variable (days since previously contacted), customers who haven't been prreviously contacted will recieve a value of -1. This can cause improper modeling of customers, since the variable now takes on the value of negative numbers. Instead we set -1 to zeros and created a new variable to indicate customers that have not been previously coontacted before. These steps ensure that the dataset is suitable for statistical learning models and enable accurate assessment of predictive performance. \

- Problem Statement
    - introduction of ML (as in ISLR)
        - Chapter 2: Estimating f(x)
            - How some models are more flexible, or more accurate. How some are less interpretable 
            - Assessing model accuracy
        - Chapter 4: We are dealing with classification problem

This study is motivated by the availability of bank marketing datasets that enable comparison of machine learning methods for predicting term deposit subscription behavior.

The specification of the problem begins with the recognition that we are addressing a binary classification task where the goal is to predict whether a client will subscribe to a term deposit based on demographic financial and campaign features. According to the framework presented in Introduction to Statistical Learning (ISLR), the fundamental goal is to estimate the unknown function, f(x) that relates the predictors, X to the response, Y. Some models are highly flexible and can capture complex non-linear relationships between predictors and the response. In most Cases these models achieve higher predictive accuracy. However, these flexible models often suffer from lower interpretability, making it difficult to understand the influence of individual features. Simpler models such as logistic regression are less flexible but provide coefficients that can be directly interpreted in terms of odds ratios, offering insight into feature importance.

Chapter 5 of Introduction to Statistical Learning emphasizes the importance of obtaining an unbiased estimate of a model’s test error to assess its predictive performance on unseen data. Test error quantifies how well a model generalizes beyond the data used for training, which is critical for selecting among competing models. The chapter presents several methods for estimating test error, including the simple training/test split, K-fold cross validation, and the leave-one-out cross validation (LOOCV) approach. Each method balances bias and variance differently, with cross validation generally providing a lower-variance estimate compared to a single split. These techniques are essential for model selection, tuning hyperparameters, and comparing the expected performance of alternative models. For the Bank Marketing dataset, estimating test error accurately is particularly important due to the class imbalance and the presence of both categorical and numeric predictors, which can influence model generalization.

In the following chpater of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting. It introduces methods such as subset selection, shrinkage (Ridge and LASSO regression), and dimension reduction to assess and enhance model accuracy while controlling variance. These approaches allow practitioners to identify parsimonious models that generalize well to new data, providing complementary strategies to traditional test error estimation. We apply these concepts to the models we fit on the bank data, to reduce overfitting and ultimately improve the generalizability of our models to new clients. \

    - Models we will implement (without diving into details)
We start with a logistic regression (described in methodology section), then build up adjustments and regularizations to the logestic regression to make our baseline. 
We then move to another interpretable model: KNN. This model is considered transperent in a unique, due to it's neighboring factor.
Finally we fit a Boosting model. This state-of-the-art model is not considered interpretable. Although, there are methods used to extract the contributing variables in such models. We included this model is used to gage how much accucracy we are sacrificing for interpretability.

    - Bank marketing dynamics (cite the paper) differentiate our methodology
The authors employed time ordered data splitting rolling window evaluation. 
For feature enrichment, they implimented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features. 
They compared Logistic regression with Decission trees, SVMs, and Neural networks; all complex 'black-box' models.
While Moro et al. prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).

## Methodology
For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets.

1. Logistic regression 
    * Describe fitting algorithm (from ISLR)
    * code form model_fir
    * short paragraph after each action
        * “we did this because…” or “we didn’t do this because…”
Logistic regression is used to model the probability of a binary outcome as a function of predictor variables. The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$

Formula: f(x) =

Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data.

Feature selection is performed using backward stepwise selection to identify a parsimonious subset of predictors that minimizes estimated test error as measured by AIC. Variables such as duration and previous marketing outcomes remain in the model due to their strong predictive power, whereas less informative variables are excluded to reduce variance. Categorical predictors are converted to factors, and a new binary variable previously_contacted is created to handle clients with no prior campaign contact. This preprocessing ensures that the data is suitable for statistical modeling and facilitates accurate interpretation of model coefficients.

Next, we check for potential non-linearity in continuous predictors from our predictors remaining after back-ward stepwise selection. We plotted deviance residuals against numeric features.
Based on residual diagnostics, we incorporated a polynomial term for balance to account for non-linear effects.

Regularization techniques such as Ridge (l2 penalty) and LASSO (l1 penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors. Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error. Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority.

Ridge and LASSO models yield similar results, indicating that additional shrinkage does not meaningfully improve classification of subscribers. We assume this is due to our modeling being majority indicator variables. This binary aspect of the indicator variable causes it's coefficient to be less affected by regularization techniques.
The ROC curves and AUC values of approximately 0.90 confirm that all models perform similarly in terms of discrimination between the two classes. These findings suggest that the primary limitation is bias from model flexibility rather than variance, and highlight the need for potentially more flexible classifiers to improve minority class prediction.

  2. KNN
  3. Boasted trees

## Results and Discussion
- Declare we are looking for specificity and interpretable (and cite back to specification of problem from introduction)
- Stats
    - Table of metrics for all models 
    - plot AUC curve 
- Business impact
    - Demonstrate interpretable models benefits
- Future work

Our methodology differentiates from prior work by Moro et al. (2014) in focusing on a comparative evaluation of interpretable models with minimal feature engineering, emphasizing a balance between predictive power and model transparency.

Model performance was primarily assessed using the area under the ROC curve which provides a threshold independent measure of classification quality. In contrast our work focuses on the original bank marketing dataset and emphasizes comparative evaluation of standard statistical learning models with reduced feature complexity and greater interpretability.

## References
1. dataset
2. Paper on website
3. ISLR v2

### 2025-12-13 Paper.Rmd
Draft finalization

@@
 ---
-title: "draft_group-6"
+title: "DSCI 445 Final Report"
+subtitle: "Machine Learning Models on A Bank Marketing Dataset"
 author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alaahtani"
-output:
-  html_document:
-    df_print: paged
-  pdf_document: default
+output: pdf_document
 ---

### 2025-12-14 Draft2.Rmd
Paper update1

@@
 
+```{r, include=FALSE}
+library(readr)
+library(tidyverse)
+library(boot)
+library(glmnet)
+library(caret)
+library(pROC)
+library(MASS)
+library(ggplot2)
+library(dplyr)
+
+set.seed(445)
+
+bank_full <- read_delim("data/bank-full_1.csv", 
+    delim = ";", escape_double = FALSE, trim_ws = TRUE)
+
+bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))
+
+bank_full <- bank_full %>%
+  mutate(
+    job = as.factor(job),
+    marital = as.factor(marital),
+    education = as.factor(education),
+    default = as.factor(default),
+    housing = as.factor(housing),
+    loan = as.factor(loan),
+    contact = as.factor(contact),
+    month = as.factor(month),
+    poutcome = as.factor(poutcome)) %>% 
+  mutate(previously_contacted = as.factor(ifelse(pdays == -1, "no", "yes")))
+
+## Train/Test split
+
+train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
+train_data <- bank_full[train_index, ]
+test_data  <- bank_full[-train_index, ]
+```
+
 ## Motivation
-- Bank marketing dynamics (cite the paper)
+- **Bank marketing dynamics (cite the paper)**
 
@@
 
-- Introduce dataset
+- **Introduce dataset**
     - Description of variables (under factors(
@@
 
-Our dataset was a publicily available daaset, thus had less variables for privacy concerns of the bank customers...
+Our dataset was a publicily available daaset, thus had less variables for privacy concerns of the bank customers.
+The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns.
+These campaigns were done to encourage clients to subscribe to a long-term deposit product.
+Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
 The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts.
 A list of the 16 variable found below...
+
+* age: Age in years 
+* job: Occupation (Categorical)
+* marital: Marital status (Categorical)
+* education: Highest level of education (Categorical)
+* default: has credit in default? (binary)
+* balance: average yearly balance
+* housing: has housing loan? (binary)
+* loan: has personal loan? (binary)
+* Contract: contact communication type (Categorical)
+* day: last contact day (of the month)
+* month: last contact month of year (month; 1=January)
+* duration: last contact duration, in seconds
+* campaign: number of contacts performed during this campaign and for this client
+* pdays: number of days that passed by after the client was last contacted from a previous campaign (-1 means client was not previously contacted)
+* previous: number of contacts performed before this campaign and for this client
+* poutcome: outcome of the previous marketing campaign (Categorical)
+* y: Indicator of whether the client subscribed a term deposit (Binary target)
+
 The response variable indicates whether a client subscribed to a term deposit which defines a binary classification task.
@@
 Visualizations of subscription rates by job education marital status contact type and previous outcome reveal variations in client responsiveness across different factors.
+
+*[STREAMLINE Plots and EDA summary HERE]*
 
@@
 
-- Problem Statement
+- **Problem Statement**
     - introduction of ML (as in ISLR)
@@
 For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets.
+Our models do not apply time-series modeling, since our data does not include sequintial variables.
+Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series.
+This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.
 
-1. Logistic regression 
-    * Describe fitting algorithm (from ISLR)
-    * code form model_fir
-    * short paragraph after each action
-        * “we did this because…” or “we didn’t do this because…”
+1. **Logistic regression**
+
+```{r, include=FALSE}
+full_formula <- y ~ age + job + marital + education + default + balance + 
+  housing + loan + contact + day + month + duration + campaign + 
+  pdays + previous + poutcome + previously_contacted
+
+glm.full <- glm(full_formula, data = train_data, family = "binomial")
+
+# Perform backward selection
+glm.step <- step(glm.full, direction = "backward", trace = 0)
+
+selected_formula <- formula(glm.step)
+
+## 
+glm.baseline <- glm(selected_formula, data = train_data, family = "binomial")
+summary(glm.baseline)
+
+ctrl <- trainControl(
+  method = "cv",
+  number = 5,
+  classProbs = TRUE, 
+  summaryFunction = twoClassSummary 
+)
+
+set.seed(445) 
+cv_fit <- train(
+  selected_formula, 
+  data = train_data, 
+  method = "glm", 
+  family = "binomial",
+  trControl = ctrl, 
+  metric = "ROC" 
+)
+
+test_probabilities <- predict(glm.baseline, newdata = test_data, type = "response")
+test_predictions <- factor(ifelse(test_probabilities > 0.5, "yes", "no"), levels = levels(test_data$y))
+
+conf_matrix <- confusionMatrix(data = test_predictions, reference = test_data$y, positive = "yes" )
+
+# Extract deviance residuals
+residuals <- residuals(glm.baseline, type = "deviance")
+train_data$residuals <- residuals
+
+p1 <- ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+
+glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
+    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
+summary(glm.tran)
+tran.coef <- summary(glm.tran)$coef
+
+residuals <- residuals(glm.tran, type = "deviance")
+train_data$residuals <- residuals
+
+p2 <- ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+
+## Regularization
+X_full <- model.matrix(y ~ . - 1, data = bank_full)
+Y_full <- as.numeric(bank_full$y) - 1
+
+X_train <- X_full[train_index, ]
+Y_train <- Y_full[train_index]
+
+X_test <- X_full[-train_index, ]
+Y_test <- Y_full[-train_index]
+
+cv.ridge <- cv.glmnet(X_train, Y_train, 
+                      alpha = 0, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.ridge <- cv.ridge$lambda.min
+
+cv.lasso <- cv.glmnet(X_train, Y_train, 
+                      alpha = 1, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.lasso <- cv.lasso$lambda.min
+
+baseline_prob <- predict(glm.tran, newdata = test_data, type = "response")
+baseline_pred <- factor(ifelse(baseline_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+ridge_prob <- predict(cv.ridge, s = lambda.min.ridge, newx = X_test, type = "response")
+ridge_pred <- factor(ifelse(ridge_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+lasso_prob <- predict(cv.lasso, s = lambda.min.lasso, newx = X_test, type = "response")
+lasso_pred <- factor(ifelse(lasso_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+cm_baseline <- confusionMatrix(baseline_pred, test_data$y, positive = "yes")
+cm_ridge <- confusionMatrix(ridge_pred, test_data$y, positive = "yes")
+cm_lasso <- confusionMatrix(lasso_pred, test_data$y, positive = "yes")
+
+metrics <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  Accuracy = c(cm_baseline$overall['Accuracy'], cm_ridge$overall['Accuracy'], cm_lasso$overall['Accuracy']),
+  Sensitivity = c(cm_baseline$byClass['Sensitivity'], cm_ridge$byClass['Sensitivity'], cm_lasso$byClass['Sensitivity']),
+  Specificity = c(cm_baseline$byClass['Specificity'], cm_ridge$byClass['Specificity'], cm_lasso$byClass['Specificity'])
+)
+
+roc_baseline <- roc(test_data$y, baseline_prob)
+roc_ridge <- roc(test_data$y, as.numeric(ridge_prob))
+roc_lasso <- roc(test_data$y, as.numeric(lasso_prob))
+
+roc_data <- data.frame(
+  fpr = c(1 - roc_baseline$specificities, 1 - roc_ridge$specificities, 1 - roc_lasso$specificities),
+  tpr = c(roc_baseline$sensitivities, roc_ridge$sensitivities, roc_lasso$sensitivities),
+  Model = factor(c(rep("Baseline Logit", length(roc_baseline$specificities)),
+                   rep("Ridge Logit", length(roc_ridge$specificities)),
+                   rep("LASSO Logit", length(roc_lasso$specificities))),
+                 levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit"))
+)
+
+```
+
 Logistic regression is used to model the probability of a binary outcome as a function of predictor variables.
-The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$
+The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$.
+This model fits with assumptions of independent observations and normality.
 
-Formula: f(x) =
-
-Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data.
+Formula: $$f(x) = P(Y=1 \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k)}}$$  
+Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data (ISLR).
 
@@
 This preprocessing ensures that the data is suitable for statistical modeling and facilitates accurate interpretation of model coefficients.
+We resulted 12 features:
+
+```{r}
+selected_formula
+```
+
+Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error.
+Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority.
+This model performed will with mean 5-fold CV test error metrics:
+
+```{r, echo=FALSE}
+print(conf_matrix)
+```
 
 Next, we check for potential non-linearity in continuous predictors from our predictors remaining after back-ward stepwise selection. We plotted deviance residuals against numeric features.
-Based on residual diagnostics, we incorporated a polynomial term for balance to account for non-linear effects.
+Based on residual diagnostics, we incorporated a polynomial term (using the poly function: *poly(balance, 4)*) for balance to account for non-linear effects.
+The Balance squared and power 4 term produced significant coefficients (Last 4 on the list of coefficients).
+
+```{r, out.width="45%", echo=FALSE}
+p1
+p2
+```
+
+```{r}
+tran.coef
+```
+
+Although this polynomial transformation didn't provide desired metrics in expense of the nonlinear effect.
+From the deviance residual plots we can observe there weren't significant deviance from a normal residual line.
+We will not include this model in our comparison.
 
 Regularization techniques such as Ridge (l2 penalty) and LASSO (l1 penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors.
-Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error.
-Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority.
+LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$).
+A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6).
+The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`.
+Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.
+
+```{r, out.width="45%", echo=FALSE}
+plot(cv.ridge)
+plot(cv.lasso)
+```
+
+Here is a comparison of the models fitted thus far:
+
+```{r}
+metrics
+```
 
@@
 This binary aspect of the indicator variable causes it's coefficient to be less affected by regularization techniques.
-The ROC curves and AUC values of approximately 0.90 confirm that all models perform similarly in terms of discrimination between the two classes.
-These findings suggest that the primary limitation is bias from model flexibility rather than variance, and highlight the need for potentially more flexible classifiers to improve minority class prediction.
 
-  2. KNN
-  3. Boasted trees
+```{r, out.width="45%", echo=FALSE}
+ggplot(roc_data, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(linewidth = 1) +
+  geom_abline(linetype = "dashed", color = "grey") +
+  labs(
+    title = "ROC Curve Comparison for Logistic Models",
+    x = "False Positive Rate (1 - Specificity)",
+    y = "True Positive Rate (Sensitivity)"
+  ) +
+  theme_minimal()
+```
+
+All three models exhibit a high overall Accuracy (around 0.89), which is expected given the significant class imbalance (Prevalence $\approx 0.11$).
+However, the critical metric, Sensitivity (the True Positive Rate for the minority class, 'yes'), is very low (around 0.17).
+This result, consistent across all three models, indicates that they struggle to correctly identify subscribers, leading to many False Negatives (FN).
+Conversely, the high Specificity (True Negative Rate $\approx 0.98$) means the models are excellent at identifying non-subscribers.
+The regularization methods (Ridge and LASSO) failed to provide any significant improvement in these metrics, suggesting that the primary limitation is one of high bias (due to insufficient flexibility) rather than high variance.
+\
+
+  2. **KNN** 
+      - Describe fitting algorithm (from ISLR)
+      - sample code and plots form model_fit file
+      - short paragraph after each action (“we did this because…” or “we didn’t do this because…”) \
+  3. **Boasted trees** 
+      - Describe fitting algorithm (from ISLR)
+      - sample code and plots form model_fit file
+      - short paragraph after each action (“we did this because…” or “we didn’t do this because…”) \
 

### 2025-12-14 model_fit.Rmd
Paper update1

@@
 
+```{r}
+ctrl <- trainControl(
+  method = "cv",
+  number = 5,
+  classProbs = TRUE, 
+  summaryFunction = twoClassSummary 
+)
+
+set.seed(445)
+cv_fit <- train(
+  selected_formula, 
+  data = train_data, 
+  method = "glm", 
+  family = "binomial",
+  trControl = ctrl, 
+  metric = "ROC" 
+)
+
+print(cv_fit)
+
+test_probabilities <- predict(
+  glm.baseline, 
+  newdata = test_data, 
+  type = "response"
+)
+
+test_predictions <- factor(
+  ifelse(test_probabilities > 0.5, "yes", "no"),
+  levels = levels(test_data$y)
+)
+
+conf_matrix <- confusionMatrix(
+  data = test_predictions, 
+  reference = test_data$y, 
+  positive = "yes" 
+)
+
+print(conf_matrix)
+
+```
+
 Assessing Non-Linearity with Residuals
@@
 summary(glm.tran)
+tran.coef <- summary(glm.tran)$coef
 
@@
 
-ggplot(train_data, aes(x = balance, y = residuals)) +
+p2 <- ggplot(train_data, aes(x = balance, y = residuals)) +
   geom_point(alpha = 0.5) +
@@
 
-The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, r round(lambda.min.lasso, 5).
+The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`.
 Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.

### 2025-12-14 Paper.Rmd
did the discussion, the result, the model training, the boosting, and data cleaning section

@@
 Predictive modeling bears great value in this scenario, as it helps the bank focus resources on only those customers who depict a higher possibility of subscribing.
+
 In the subscription rate of this dataset, the problem is even more difficult as it is very low.
@@
 These variables describe the client's existing financial commitments and capacity, which may influence their willingness to invest in a long-term product like a term deposit.
+
 Marketing campaign features include the type of contact: cellular or telephone, month of contact, and day of contact, and duration of the call.
@@
 These variables highlight whether the client has been interested in the past or has been contacted many times without success.
+
 The data includes both numerical and categorical variables, and this can serve as a good example to compare the many different classification methods with respect to how each treats different data types: for instance, linear models require the conversion of categorical variables, whereas tree-based or distance-based methods would naturally handle those in different ways.
@@
 Understanding how such models perform is useful to a marketing team for making informed decisions about which clients to contact in future campaigns.
+
 We focused our attention on well-performing models that give insightful knowledge of which factors drive the choice of a client.
@@
 More interpretable models do not always have the strongest accuracy but can guide marketing strategies more directly.
+
 We make the following assumptions to keep the analysis focused and manageable:
@@
 This renders the analysis suitable for a course project while still answering substantial questions regarding classification performance.
+
 Generalization from sample to population: We assume that this dataset is a good representation of the client population to which the bank will offer its services in forthcoming campaigns.
@@
 
-## Methodology
+#### Data Wrangling and Cleaning:
+The analysis began by loading the Bank Marketing dataset and converting all categorical predictors into factor variables so they could be properly handled by the statistical modeling procedures in R.
+Several variables contained “unknown” values, which serve as placeholders rather than true missing observations.
+Multiple strategies for handling these values were considered, including treating them as missing and excluding the affected observations.
+Ultimately, “unknown” values were retained as their own category in order to preserve information and avoid altering the underlying class imbalance.
 
-#### Data Wrangling and Cleaning:
+An additional feature, previously_contacted, was engineered to address the special coding of the pdays variable, where a value of −1 indicates that a client was not previously contacted.
+This transformation allowed the model to distinguish between first-time and repeat contacts more clearly.
 
-We began by loading the dataset, converting categorical variables into factors, and identifying “unknown” levels used in place of missing data.
-We evaluated several approaches to handling these values—including keeping “unknown” as its own category and treating them as missing, but ultimately selected the method that preserved the most information without distorting class balance.
-
-A train/test split was created, and all models were trained using the same set of cleaned predictors for fair comparison.
+To evaluate model performance fairly, the dataset was partitioned into an 80% training set and a 20% test set using stratified sampling based on the response variable.
+This ensured that the original class imbalance was preserved in both subsets.
+All models were trained on the same cleaned training data and evaluated on the same test data, allowing for direct and unbiased comparison across modeling approaches.
 
@@
 This makes the method particularly well-suited when the underlying relationship between predictors and subscription behavior is nonlinear or too complicated for a parametric model to capture.
+
 Because KNN relies on distance calculations, feature scaling is an important preprocessing step.
@@
 We used 5-fold cross-validation to identify the value of k that minimized the estimated test error, balancing the bias-variance trade-off.
+
 While KNN presents an intuitive method of finding clients with similar profiles, it has several limitations in this context.
@@
 
-- Boosting (e.g., AdaBoost or Gradient Boosting) — An ensemble method that sequentially builds weak learners to minimize classification error.
-Boosting is often superior in predictive accuracy but is less interpretable.
+Boosting was used as the most flexible modeling approach in this study to address the strong class imbalance and complex predictor relationships present in the Bank Marketing dataset.
+Unlike logistic regression and KNN, boosting does not rely on a fixed functional form and can naturally capture nonlinear effects and interactions among demographic, financial, and campaign-related variables.
+
+The boosting model was trained on the same 80% training set as the other models using a Bernoulli loss function appropriate for binary classification.
+To allow the model sufficient capacity to learn complex structure, the model was initially trained with 1,500 trees.
+Model complexity was controlled using a small learning rate and shallow tree depth to encourage gradual improvement rather than abrupt overfitting.
+The optimal number of trees was selected using out-of-bag (OOB) error estimation.
+This procedure identified approximately 1,100 trees as the point at which predictive performance was maximized.
+Beyond this value, additional trees did not lead to meaningful improvements, indicating that the model had effectively converged.
+Final predictions and evaluation metrics were therefore computed using 1,100 trees to balance flexibility and generalization.
+
+When evaluated on the test set, boosting achieved the strongest overall performance among all models considered.
+It produced the highest accuracy, sensitivity, and AUC, substantially improving the identification of the minority “yes” class compared to logistic regression and KNN.
+In particular, boosting correctly identified over 40% of subscribing clients while maintaining high specificity, demonstrating its ability to reduce false negatives without substantially increasing false positives.
+These results indicate that boosting is well suited for imbalanced marketing data where identifying potential subscribers is a primary objective.
 
 ### Model Training, Formulation, and Validation
+All models were trained and evaluated within a consistent validation framework to ensure fair comparison and reliable estimation of out-of-sample performance.
+An 80/20 stratified train–test split was used, and all modeling decisions, including feature selection and hyperparameter tuning, were made using only the training data.
+This prevented information leakage and ensured that the test set provided an honest evaluation of generalization performance.
 
-All models were trained using k-fold cross-validation to estimate out-of-sample performance. This approach reduces variance and ensures that results do not depend on a single train/test split.
+For the logistic regression models, feature selection was performed using backward stepwise selection based on AIC to obtain a parsimonious set of predictors.
+Residual diagnostics were used to assess model assumptions, leading to the inclusion of nonlinear terms where appropriate.
+Regularization techniques, including Ridge and LASSO regression, were applied to address multicollinearity and reduce variance.
+The regularization parameter was selected using 5-fold cross-validation.
 
-We also carried out:
-- Feature selection using stepwise regression and subset selection for logistic regression
-- Regularization techniques (ridge/lasso) when appropriate
-- Dimensionality reduction (e.g., PCA) if needed to improve KNN performance
-- Hyperparameter tuning for KNN (choice of k)
+K-Nearest Neighbors required additional preprocessing due to its reliance on distance-based calculations.
+All numeric predictors were standardized to ensure equal contribution to the distance metric.
+The number of neighbors,
+k, was selected using 5-fold cross-validation by identifying the value that minimized classification error on the training folds.
 
-This section of the paper will include the formal model equations and the details of the validation procedure.
+Boosting was trained using a large number of trees to allow sufficient model flexibility, with the optimal number of trees selected using out-of-bag error estimation.
+This tuning procedure served a similar role to cross-validation by identifying the point at which additional complexity no longer improved performance.
+Across all models, final performance was assessed on the held-out test set using accuracy, sensitivity, and AUC.
+Because the dataset is highly imbalanced, particular emphasis was placed on sensitivity and AUC as measures of a model’s ability to identify the minority “yes” class.
 
@@
 
-Once our models are trained, we will compare their performance using test error rate, confusion matrices, and metrics such as accuracy, recall, and AUC.
-Because the dataset is imbalanced, we expect recall and AUC to be especially important for evaluating how well each model identifies clients who subscribe.
+This study evaluated multiple statistical and machine learning models for predicting term deposit subscriptions using the Bank Marketing dataset.
+Due to the strong class imbalance, model performance was assessed using accuracy, sensitivity, and AUC rather than accuracy alone.
+Logistic regression and its regularized variants demonstrated strong overall discrimination and high AUC values, but they struggled to identify the minority class of subscribing clients.
+K-Nearest Neighbors performed less effectively, showing limited discriminatory power in this high-dimensional, imbalanced setting.
 
-We will also create ROC curves to compare models visually and use information criteria (AIC/BIC) for parametric models where appropriate.
-After summarizing these results, we will determine which model performs best overall and which provides the most useful insights for guiding future marketing decisions
+Boosting outperformed all other models, achieving the highest accuracy, sensitivity, and AUC on the test set.
+Its superior performance indicates that flexible ensemble methods are better suited for capturing complex, nonlinear relationships in marketing data.
+However, logistic regression remains valuable due to its interpretability and ability to provide actionable insights.
+Overall, the results highlight an important trade-off between predictive performance and interpretability when modeling imbalanced marketing data.
 
@@
 
-Since we are still exploring the dataset and fitting our models, we have not yet determined which classification method will perform best.
-However, based on the structure of the Bank Marketing data and findings from similar studies, we expect that boosting is likely to give the best predictive accuracy, while logistic regression will likely be the most interpretable and actionable for managers.
+The results of this study highlight an important trade-off between predictive performance and interpretability when modeling imbalanced bank marketing data.
+Among all models evaluated, boosting achieved the strongest overall predictive performance, particularly in terms of sensitivity and AUC.
+This indicates that boosting is the most effective method for identifying clients who are likely to subscribe to a term deposit, which is the primary objective of the bank.
+By capturing complex nonlinear relationships and interactions among predictors, boosting substantially reduced the number of missed potential subscribers compared to simpler models.
 
-In the final version of this section, we will compare the models based on their test performance, interpretability, and robustness to class imbalance.
-We will also discuss any limitations of our approach and suggest future improvements, such as trying resampling methods or evaluating additional machine learning models.
+However, while boosting offers superior predictive accuracy, it lacks transparency.
+In contrast, regularized logistic regression provides more interpretable results that can be directly translated into business insights.
+Coefficient estimates from the logistic models reveal how specific client characteristics influence subscription probability.
+For example, higher account balances were associated with lower marginal increases in subscription probability at extreme values, housing loan status was negatively associated with subscription, and student job status showed a strong positive association.
+These interpretable effects allow marketing teams to better understand customer behavior and design targeted strategies, even if the model itself is not the most accurate.
+
+Interpretability is particularly important in the banking sector, where predictive decisions may be subject to regulatory oversight and ethical scrutiny.
+Models that can clearly justify why a client was targeted or excluded are easier to audit and defend.
+From this perspective, logistic regression remains a valuable tool despite its lower sensitivity.
+A practical strategy for the bank may involve using boosting as a primary screening model to identify high-potential clients, followed by logistic regression to provide explanations and support decision-making.
+Several limitations should be noted.
+The analysis did not incorporate time-aware modeling, even though campaign timing may influence client responses.
+Additionally, no resampling techniques were used to directly address class imbalance.
+Future work could explore methods such as SMOTE or cost-sensitive learning to further improve minority-class detection.
+Other extensions may include experimenting with alternative ensemble methods or adjusting decision thresholds to better align predictions with business costs.
+
+Overall, this study demonstrates that flexible ensemble models offer substantial gains in predictive performance for imbalanced marketing problems, while interpretable models remain essential for transparency and actionable insight.
+Balancing these two objectives is critical for effective and responsible deployment of predictive models in real-world banking applications.
 

### 2025-12-14 Slides.tex
did the discussion, the result, the model training, the boosting, and data cleaning section

--- 2025-12-07 Slides.tex
@@
 \PassOptionsToPackage{hyphens}{url}
+%
 \documentclass[
@@
 ]{beamer}
-\newif\ifbibliography
 \usepackage{pgfpages}
@@
 \beamertemplatenavigationsymbolsempty
-% remove section numbering
+% Prevent slide breaks in the middle of a paragraph
+\widowpenalties 1 10000
+\raggedbottom
 \setbeamertemplate{part page}{
@@
 }
-% Prevent slide breaks in the middle of a paragraph
-\widowpenalties 1 10000
-\raggedbottom
 \AtBeginPart{
@@
 }
+\usepackage{amsmath,amssymb}
 \usepackage{iftex}
@@
 \makeatother
+\usepackage{xcolor}
+\newif\ifbibliography
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+\usepackage{caption}
+% Make caption package work with longtable
+\makeatletter
+\def\fnum@table{\tablename~\thetable}
+\makeatother
+\usepackage{graphicx}
+\makeatletter
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
+% Set default figure placement to htbp
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Statistical Machine Learning Models on A Bank Marketing dataset},
-  pdfauthor={Sandra, Tigist, Mussa, Waddah},
+  pdftitle={Statistical Machine Learning Models on A Bank Marketing Dataset},
+  pdfauthor={Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah Alqahtani},
   hidelinks,
@@
 
-\title{Statistical Machine Learning Models on A Bank Marketing dataset}
+\title{Statistical Machine Learning Models on A Bank Marketing Dataset}
 \subtitle{Colorado State University, DSCI 445: Statistical Machine
 Learning, Dr.Andee Kaplan}
-\author{Sandra, Tigist, Mussa, Waddah}
+\author{Sandra Afrifa, Tigist Kefelew, Mussa Hassen, Waddah Alqahtani}
+\date{}
 
@@
 \begin{frame}{Project Motivation \& Business Problem}
-\protect\phantomsection\label{project-motivation-business-problem}
+\phantomsection\label{project-motivation-business-problem}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} The bank's need for \textbf{efficient marketing} and
-  the high cost of telemarketing campaigns. The goal is to maximize the
-  success rate by predicting customer subscription.
+  Telemarketing campaigns are expensive, time-consuming, and have very
+  low success rates. The bank wants to improve efficiency by identifying
+  which clients are most likely to subscribe.
 \item
-  \textbf{Discuss:} The challenge of the \textbf{rare positive class}
-  (imbalanced data) and why prediction accuracy is crucial for
-  \textbf{resource allocation}.
+  The dataset has a severe class imbalance: only a small fraction of
+  customers subscribe to the term deposit. This makes prediction
+  difficult and increases the risk of false negatives.
 \item
-  \textbf{Image/Plot Idea:} A simple, high-level infographic of a
-  marketing campaign funnel, highlighting the bottleneck/drop-off point.
+  Improving predictive accuracy would allow the bank to target fewer
+  customers, reduce costs, and increase return on investment (ROI) for
+  future campaigns.
 \end{itemize}
-
-{[}Image of a marketing funnel illustrating customer conversion{]}
 \end{frame}
 
-\begin{frame}{The Bank Marketing Dataset (Slide 2)}
-\protect\phantomsection\label{the-bank-marketing-dataset-slide-2}
+\begin{frame}{The Bank Marketing Dataset}
+\phantomsection\label{the-bank-marketing-dataset}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} The dataset structure: \(\approx 45,000\) client
-  observations from a Portuguese banking institution's telemarketing
-  campaigns.
+  \textbf{45,000 client observations} from Portuguese bank
+  telemarketing\\
 \item
-  \textbf{Discuss:} Brief overview of the \textbf{four main predictor
-  categories}: Demographics, Financial Indicators, Campaign Features,
-  and Past Outcomes.
+  \textbf{Target variable (\(y\)):} Did client subscribe to term
+  deposit? (Yes/No)\\
 \item
-  \textbf{Image/Plot Idea:} A small table summarizing the number of
-  observations and the list of the 4 key variable groups (e.g.,
-  Demographics, Financial, Campaign, Past).
+  \textbf{Class imbalance:} Only ≈11\% ``Yes'' subscriptions\\
+\item
+  \textbf{Four main predictor groups:} demographics, financial
+  indicators, campaign features, past outcomes
 \end{itemize}
@@
 
-\begin{frame}{Target Variable \& Class Imbalance (Slide 3)}
-\protect\phantomsection\label{target-variable-class-imbalance-slide-3}
+\begin{frame}{Term Deposit Class Imbalance}
+\phantomsection\label{term-deposit-class-imbalance}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} The \textbf{Target Variable (\(y\))} is binary: Did
-  the client subscribe to a term deposit (Yes/No)?
+  \textbf{Most clients say NO; only ≈11\% say YES}
+\end{itemize}
+
+\begin{center}\includegraphics[width=0.75\linewidth]{Slides_files/figure-beamer/imbalance-plot-1} \end{center}
+\end{frame}
+
+\begin{frame}{EDA Finding 1: Contact Method Matters}
+\phantomsection\label{eda-finding-1-contact-method-matters}
+\begin{itemize}
+\tightlist
 \item
-  \textbf{Discuss:} Emphasize the severity of the \textbf{class
-  imbalance} (e.g., typically \(\approx 11\%\) `Yes' subscriptions).
-  This means models must be evaluated beyond simple accuracy.
+  Cell phone calls perform best
+\end{itemize}
+
+\begin{center}\includegraphics[width=0.75\linewidth]{Slides_files/figure-beamer/contact-plot-1} \end{center}
+\end{frame}
+
+\begin{frame}{EDA Finding 2: Previous Success Predicts Future Success}
+\phantomsection\label{eda-finding-2-previous-success-predicts-future-success}
+\begin{itemize}
+\tightlist
 \item
-  \textbf{Image/Plot Idea:} A \textbf{Bar Plot} showing the distribution
-  of the target variable (\(y\)), visually highlighting the imbalance.
+  Past success predicts new success
+\end{itemize}
+
+\begin{center}\includegraphics[width=0.7\linewidth]{Slides_files/figure-beamer/poutcome-plot-1} \end{center}
+\end{frame}
+
+\begin{frame}{EDA Finding 3: Too Many Calls Hurts Results}
+\phantomsection\label{eda-finding-3-too-many-calls-hurts-results}
+\begin{itemize}
+\tightlist
+\item
+  More calls = worse results
+\end{itemize}
+
+\begin{center}\includegraphics[width=0.75\linewidth]{Slides_files/figure-beamer/campaign-plot-1} \end{center}
+\end{frame}
+
+\begin{frame}{Methodology: Data Preparation}
+\phantomsection\label{methodology-data-preparation}
+\begin{itemize}
+\tightlist
+\item
+  Converted all categorical variables into factors so models interpret
+  them correctly.
+\item
+  Addressed ``unknown'' values by treating them as their own category to
+  avoid losing information.
+\item
+  Created a new feature previously\_contacted to properly encode pdays.
+\item
+  Split the dataset into 80\% training and 20\% test using stratified
+  sampling to preserve the class imbalance.
+\item
+  Ensured numeric predictors were available for later scaling, tuning,
+  and model evaluation.
 \end{itemize}
@@
 
-\begin{frame}{Problem Statement \& Key Assumptions (Slide 4)}
-\protect\phantomsection\label{problem-statement-key-assumptions-slide-4}
+\begin{frame}{Breakdown of Categorical Variable}
+\phantomsection\label{breakdown-of-categorical-variable}
+\pandocbounded{\includegraphics[keepaspectratio]{Slides_files/figure-beamer/unnamed-chunk-1-1.pdf}}
+\end{frame}
+
+\section{1. Logistic Regression Model}\label{logistic-regression-model}
+
+\begin{frame}[fragile]{Feature Selection \& Linearity}
+\phantomsection\label{feature-selection-linearity}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} The project goal is \textbf{Binary Classification}:
-  Building statistical models that are both \textbf{predictive} and
-  \textbf{interpretable} for management.
+  Initial feature selection was performed using \textbf{Backward
+  Stepwise Elimination} on the full model to find the most useful set of
+  main effects.
 \item
-  \textbf{Discuss:} Two key assumptions: Data observations are
-  \textbf{independent} (no time-series methods), and we focus on
-  \textbf{course-covered models} for direct comparison.
+  Our selected formula resulted:
+\end{itemize}
+
+\(y \sim job + marital + education + balance + housing + loan + contact +\)
+
+\(day + month + duration + campaign + previous+ poutcome\)
+
+\begin{itemize}
+\tightlist
 \item
-  \textbf{Image/Plot Idea:} An icon representing \textbf{Predictive
-  Power} next to an icon representing \textbf{Interpretability} (The two
-  core goals).
+  Residual Plots were used to assess the linearity assumption for
+  numeric predictors (\texttt{age}, \texttt{balance}). This led to
+  testing a model with non-linear terms.
 \end{itemize}
@@
 
-\begin{frame}{Methodology: Data Preparation (Slide 5)}
-\protect\phantomsection\label{methodology-data-preparation-slide-5}
+\begin{frame}[fragile]
+\begin{verbatim}
+## `geom_smooth()` using formula = 'y ~ x'
+\end{verbatim}
+
+\begin{center}\includegraphics{Slides_files/figure-beamer/residuals_slide-1} \end{center}
+\end{frame}
+
+\begin{frame}{Models with non-linear components and interactions}
+\phantomsection\label{models-with-non-linear-components-and-interactions}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} \textbf{Data Wrangling} involved converting
-  categorical variables to factors and strategies for handling `unknown'
-  levels (treating them as a separate category vs.~missing).
+  A \textbf{quadratic term} \(I(\text{balance}^4)\)---was added to the
+  formula to reduce the model's bias.
 \item
-  \textbf{Discuss:} Established a \textbf{train/test split} (e.g.,
-  70/30) to ensure models are evaluated on unseen data for a fair
-  comparison of out-of-sample error.
+  We then tested for crucial \textbf{interaction terms} (between
+  campaign features and demographics).
 \item
-  \textbf{Image/Plot Idea:} A simple flow chart showing: Raw Data
-  \(\rightarrow\) Wrangling/Cleaning \(\rightarrow\) Train/Test Split.
+  Our model had a cross-validated test error rate 0.098`
 \end{itemize}
-
-{[}Image of a data science workflow diagram{]}
 \end{frame}
 
-\begin{frame}{Methodology: Models Selected (Slide 6)}
-\protect\phantomsection\label{methodology-models-selected-slide-6}
+\begin{frame}{Regularization \& Tuning}
+\phantomsection\label{regularization-tuning}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} \textbf{Logistic Regression:} Chosen for its
-  \textbf{linearity} and \textbf{high interpretability} (provides clear
-  marketing insights via coefficients).
+  We explored Ridge and LASSO regularization as an alternative approach
+  for feature selection.
 \item
-  \textbf{Discuss:} \textbf{K-Nearest Neighbors (KNN):} A
-  \textbf{non-parametric} and \textbf{flexible} model to capture
-  non-linearities, though sensitive to scale.
+  \textbf{5-Fold Cross-Validation} was used to select the optimal
+  penalty parameter, \(\lambda\), which minimized the mean CV error for
+  each regularized model.
+\end{itemize}
+
+\pandocbounded{\includegraphics[keepaspectratio]{Slides_files/figure-beamer/unnamed-chunk-2-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{Slides_files/figure-beamer/unnamed-chunk-2-2.pdf}}
+\end{frame}
+
+\begin{frame}{Evaluation}
+\phantomsection\label{evaluation}
+\begin{itemize}
+\tightlist
 \item
-  \textbf{Image/Plot Idea:} A side-by-side visual comparing the
-  conceptual difference between a linear boundary (LogReg) and a local,
-  non-linear one (KNN).
+  The final selected LogReg model was evaluated on the 5-fold CV error
+  rate, along with metrics focused on the confusion matrix.
+\item
+  Models performed relatively similar. We would prefer a model with
+  higher \emph{sensitivity} (few false negatives)
+\end{itemize}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\toprule\noalign{}
+Model & Accuracy & Sensitivity & Specificity & AUC \\
+\midrule\noalign{}
+\endhead
+Baseline Logit & 0.8996792 & 0.3301798 & 0.9750752 & 0.9077627 \\
+Ridge Logit & 0.8982413 & 0.2800378 & 0.9800852 & 0.9080043 \\
+LASSO Logit & 0.8981307 & 0.3008515 & 0.9772044 & 0.9072429 \\
+\bottomrule\noalign{}
+\end{longtable}
+\end{frame}
+
+\section{2. K-Nearest Neighbors (KNN)
+Model}\label{k-nearest-neighbors-knn-model}
+
+\begin{frame}{KNN: Feature Preparation}
+\phantomsection\label{knn-feature-preparation}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{KNN} requires scaling: all numeric predictors were
+  \textbf{standardized} (mean=0, SD=1) to prevent features with larger
+  scales from dominating the distance calculation.
+\item
+  The non-parametric nature means \textbf{no explicit feature selection}
+  via coefficients, but the scaling step is vital for ensuring equal
+  contribution from all dimensions.
+\end{itemize}
+
+\begin{center}\includegraphics{Slides_files/figure-beamer/knn-prep-1} \end{center}
+
+\textbf{Test Accuracy (k=5):} 88.55\%
+\end{frame}
+
+\begin{frame}{KNN: Hyperparameter Tuning}
+\phantomsection\label{knn-hyperparameter-tuning}
+\begin{itemize}
+\tightlist
+\item
+  The sole hyperparameter, \textbf{\(k\)} (the number of neighbors),
+  determines the model's complexity (i.e., the bias-variance trade-off).
+\item
+  \textbf{5-Fold Cross-Validation} was performed across a range of \(k\)
+  values to find the optimal \(k\) that minimized the mean CV error.
+\end{itemize}
+
+\begin{center}\includegraphics{Slides_files/figure-beamer/knn-tuning-1} \end{center}
+
+\textbf{Optimal k:} 23
+\end{frame}
+
+\begin{frame}{KNN: Evaluation \& Test Performance}
+\phantomsection\label{knn-evaluation-test-performance}
+\begin{itemize}
+\tightlist
+\item
+  The final KNN model (with optimal \(k\)) was applied to the
+  \textbf{Test Set}. We evaluate performance using \textbf{AUC} and
+  compare it to the LogReg models.
+\item
+  KNN generally provides a good intermediate balance---often capturing
+  non-linearity better than LogReg but being less computationally
+  expensive than boosting.
+\end{itemize}
+
+\begin{longtable}[]{@{}lrr@{}}
+\toprule\noalign{}
+Model & AUC & Sensitivity \\
+\midrule\noalign{}
+\endhead
+KNN & 0.852 & 0.260 \\
+Baseline Logit & 0.908 & 0.330 \\
+Ridge & 0.908 & 0.280 \\
+LASSO & 0.907 & 0.301 \\
+\bottomrule\noalign{}
+\end{longtable}
+\end{frame}
+
+\section{3. Boosting Tress Model}\label{boosting-tress-model}
+
+\begin{frame}{Boosting: Model Formulation}
+\phantomsection\label{boosting-model-formulation}
+\begin{itemize}
+\tightlist
+\item
+  Boosting builds many small trees sequentially, each fixing errors from
+  the previous one.
+\item
+  These weak learners combine into a strong model that captures complex,
+  non-linear patterns.
+\item
+  It usually achieves the highest accuracy/AUC, but is less
+  interpretable than LogReg or KNN.
 \end{itemize}
@@
 
-\begin{frame}{Methodology: Ensemble/Advanced Models (Slide 7)}
-\protect\phantomsection\label{methodology-ensembleadvanced-models-slide-7}
+\begin{frame}
+\begin{center}\includegraphics[width=0.8\linewidth]{Slides_files/figure-beamer/boosting-viz-1} \end{center}
+\end{frame}
+
+\begin{frame}{Boosting: Hyperparameter Tuning}
+\phantomsection\label{boosting-hyperparameter-tuning}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} \textbf{Boosting (e.g., AdaBoost/Gradient
-  Boosting):} An ensemble method expected to provide the \textbf{highest
-  predictive accuracy} by sequentially correcting errors of weak
-  learners.
+  We tuned the number of trees using OOB error estimation to prevent
+  overfitting.
 \item
-  \textbf{Discuss:} \textbf{Model Training Strategy:} All models used
-  \textbf{k-fold Cross-Validation} (CV) on the training set to prevent
-  overfitting and select optimal hyperparameters.
+  The OOB method selected 1100 trees as optimal.
 \item
-  \textbf{Image/Plot Idea:} A conceptual diagram illustrating the
-  boosting process (e.g., weak learners combining to form a strong
-  learner).
+  This choice balances model complexity and generalization.
 \end{itemize}
@@
 
-\begin{frame}{Model Tuning \& Validation Details (Slide 8)}
-\protect\phantomsection\label{model-tuning-validation-details-slide-8}
+\begin{frame}
+\begin{center}\includegraphics[width=0.75\linewidth]{Slides_files/figure-beamer/boosting-tuning-1} \end{center}
+
+\textbf{Optimal number of trees (via test error):} 1100
+\end{frame}
+
+\begin{frame}{Boosting: Evaluation \& Importance}
+\phantomsection\label{boosting-evaluation-importance}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Mentioned specific tuning steps:
-  \textbf{Hyperparameter tuning} (e.g., finding optimal \(k\) for KNN)
-  and \textbf{Feature Selection} (e.g., stepwise for LogReg) to simplify
-  models.
+  The final Boosting model achieved the highest AUC and sensitivity
+  among all models tested.
 \item
-  \textbf{Discuss:} For models like LogReg, we also explored
-  \textbf{Regularization} (Ridge/Lasso) to prevent overfitting and
-  improve generalization.
+  Variable Importance Plot shows which features influence predictions
+  most.
 \item
-  \textbf{Image/Plot Idea:} A plot showing CV error vs.~\(k\) (for KNN)
-  or \(\lambda\) (for Lasso/Ridge) to demonstrate hyperparameter
-  selection.
+  The importance rankings align with key predictors identified by
+  Logistic Regression.
 \end{itemize}
@@
 
-\begin{frame}{Results: Model Evaluation Metrics (Slide 9)}
-\protect\phantomsection\label{results-model-evaluation-metrics-slide-9}
+\begin{frame}
+\pandocbounded{\includegraphics[keepaspectratio]{Slides_files/figure-beamer/boosting-results-1.pdf}}
+
+\textbf{Boosting Test AUC:} 0.927\\
+\textbf{Boosting Sensitivity:} 0.412\\
+\textbf{Boosting Accuracy:} 0.905
+\end{frame}
+
+\begin{frame}{Results: Comparative Evaluation}
+\phantomsection\label{results-comparative-evaluation}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Primary metrics are \textbf{Test Error Rate} and
-  \textbf{AUC (Area Under the ROC Curve)}, which are robust to class
-  imbalance.
+  We evaluated five models using Accuracy, Sensitivity (Recall), and AUC
+  on the Test Set.
 \item
-  \textbf{Discuss:} Also emphasize \textbf{Recall} (sensitivity) because
-  identifying the few `Yes' subscribers is more critical than high
-  accuracy on the majority `No' class.
+  \textbf{Boosting} achieved the highest overall accuracy and the
+  highest sensitivity, meaning it identified the most true positives.
 \item
-  \textbf{Image/Plot Idea:} A visual diagram of a \textbf{Confusion
-  Matrix}, highlighting the meaning of True Positives (TP) and False
-  Negatives (FN).
+  Logistic, Ridge, and LASSO models performed very similarly in accuracy
+  and AUC, but with lower sensitivity.
+\item
+  KNN performed the worst, with the lowest AUC, indicating weak
+  separation between classes.
 \end{itemize}
@@
 
-\begin{frame}{Results: Performance Comparison (Slide 10)}
-\protect\phantomsection\label{results-performance-comparison-slide-10}
+\begin{frame}{Model Performance Comparison}
+\phantomsection\label{model-performance-comparison}
+\begin{longtable}[]{@{}lrrr@{}}
+\toprule\noalign{}
+Model & Accuracy & Sensitivity & AUC \\
+\midrule\noalign{}
+\endhead
+Logistic & 0.900 & 0.330 & 0.908 \\
+Ridge & 0.898 & 0.280 & 0.908 \\
+LASSO & 0.898 & 0.301 & 0.907 \\
+KNN & 0.891 & 0.260 & 0.852 \\
+Boosting & 0.905 & 0.412 & 0.927 \\
+\bottomrule\noalign{}
+\end{longtable}
+\end{frame}
+
+\begin{frame}{ROC Curve Comparison}
+\phantomsection\label{roc-curve-comparison}
+\begin{center}\includegraphics{Slides_files/figure-beamer/final-roc-1} \end{center}
+\end{frame}
+
+\begin{frame}{Business Impact}
+\phantomsection\label{business-impact}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Present the \textbf{final performance comparison},
-  identifying the model with the lowest Test Error/Highest AUC (e.g.,
-  \textbf{Boosting}).
+  The central trade-off: Boosting is the best predictor, but the
+  \textbf{Regularized Logistic Regression} offers the most
+  \emph{interpretable} and \emph{actionable} insights for the bank.
 \item
-  \textbf{Discuss:} Acknowledge the trade-off: The \textbf{best
-  predictive model} may not be the \textbf{most interpretable}.
+  Sample coefficients: \(\beta_{Balance^2}\) = -23.02,
+  \(\beta_{Housing}\) = -0.65, \(\beta_{JobStudent}\) = 4.064
 \item
-  \textbf{Image/Plot Idea:} A \textbf{Table} or \textbf{Bar Chart}
-  summarizing AUC, Accuracy, and Recall for \textbf{Logistic
-  Regression}, \textbf{KNN}, and \textbf{Boosting}.
+  The chosen model can lead to cost savings and higher ROI for future
+  campaigns.
 \end{itemize}
@@
 
-\begin{frame}{Results: ROC Curve Analysis (Slide 11)}
-\protect\phantomsection\label{results-roc-curve-analysis-slide-11}
+\begin{frame}{Discussion}
+\phantomsection\label{discussion}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} A visual comparison of models using the
-  \textbf{Receiver Operating Characteristic (ROC) Curve}. A curve closer
-  to the top-left is better.
+  Consideration: regulatory/ethical implications in banking and the need
+  to justify predictions (further supporting the value of
+  interpretability).
 \item
-  \textbf{Discuss:} This curve provides insight into how each model
-  balances the \textbf{True Positive Rate} (Recall) vs.~\textbf{False
-  Positive Rate}.
+  Advanced train/test splitting methods like \emph{time-series analysis}
+  were not used.
 \item
-  \textbf{Image/Plot Idea:} The final \textbf{ROC Curve plot} comparing
-  the performance of all tested models.
+  Suggest future improvements: Exploring advanced techniques like
+  \emph{resampling methods} (SMOTE) to handle imbalance better or
+  trying.
 \end{itemize}
@@
 
-\begin{frame}{Discussion: Interpreting the Best Model (Slide 12)}
-\protect\phantomsection\label{discussion-interpreting-the-best-model-slide-12}
+\begin{frame}{References}
+\phantomsection\label{references}
 \begin{itemize}
@@
 \item
-  \textbf{Discuss:} Focus on the \textbf{most interpretable model
-  (Logistic Regression)}---what were the key coefficients/insights?
-  (e.g., `Duration of last contact' is a strong predictor).
+  UC Irvine Machine Learning Repository. Bank Marketing.
+  \url{https://archive.ics.uci.edu/dataset/222/bank+marketing}
 \item
-  \textbf{Discuss:} Discuss how these feature importances can be
-  directly translated into \textbf{actionable marketing strategies} for
-  the bank.
-\item
-  \textbf{Image/Plot Idea:} A \textbf{Feature Importance Plot} (for
-  Boosting) or a \textbf{Coefficient Plot} (for LogReg), highlighting
-  the top 3-5 most influential variables.
+  Github link: \url{https://github.com/dsci445-csu/project-6}
 \end{itemize}
-\end{frame}
 
-\begin{frame}{Discussion: Business Impact (Slide 13)}
-\protect\phantomsection\label{discussion-business-impact-slide-13}
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Discuss:} How the \textbf{chosen model} (Balancing performance
-  and interpretability) can lead to \textbf{cost savings} and
-  \textbf{higher ROI} for future campaigns.
-\item
-  \textbf{Discuss:} Briefly touch upon \textbf{regulatory/ethical
-  implications} in banking and the need to justify predictions (further
-  supporting the value of interpretability).
-\item
-  \textbf{Image/Plot Idea:} A currency symbol or a bar chart showing the
-  conceptual increase in ROI due to better targeting.
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Discussion: Limitations \& Future Work (Slide 14)}
-\protect\phantomsection\label{discussion-limitations-future-work-slide-14}
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Discuss:} Acknowledge limitations, such as not using
-  \textbf{time-series analysis} or fully accounting for the dependency
-  of the `duration' feature.
-\item
-  \textbf{Discuss:} Suggest future improvements: Exploring advanced
-  techniques like \textbf{resampling methods} (SMOTE) to handle
-  imbalance better or trying \textbf{Deep Learning} models.
-\item
-  \textbf{Image/Plot Idea:} A bulleted list of 2-3 \textbf{future
-  research questions} (e.g., Time-Series, SMOTE, Deep Learning).
-\end{itemize}
-\end{frame}
-
-\begin{frame}{References (Slide 15)}
-\protect\phantomsection\label{references-slide-15}
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Discuss:} Simply list the required references for the data
-  source and any major methodology papers.
-\item
-  \textbf{Discuss:} Provide contact information for questions.
-\item
-  \textbf{Image/Plot Idea:} The official UCI ML Repository logo and your
-  contact email.
-\end{itemize}
+\begin{center}\includegraphics{UCI} \end{center}
 \end{frame}

### 2025-12-14 Paper.Rmd
Final report update

@@
 
-## Motivation # TT
-Bank Marketing dataset The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns.
+```{r, include=FALSE}
+library(readr)
+library(tidyverse)
+library(boot)
+library(glmnet)
+library(caret)
+library(pROC)
+library(MASS)
+library(ggplot2)
+library(dplyr)
+
+set.seed(445)
+
+bank_full <- read_delim("data/bank-full_1.csv", 
+    delim = ";", escape_double = FALSE, trim_ws = TRUE)
+
+bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))
+
+bank_full <- bank_full %>%
+  mutate(
+    job = as.factor(job),
+    marital = as.factor(marital),
+    education = as.factor(education),
+    default = as.factor(default),
+    housing = as.factor(housing),
+    loan = as.factor(loan),
+    contact = as.factor(contact),
+    month = as.factor(month),
+    poutcome = as.factor(poutcome)) %>% 
+  mutate(previously_contacted = as.factor(ifelse(pdays == -1, "no", "yes")))
+
+## Train/Test split
+
+train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
+train_data <- bank_full[train_index, ]
+test_data  <- bank_full[-train_index, ]
+```
+
+## Motivation
+
+Bank telemarketing campaigns are a major channel through which financial institutions promote term deposit products.
+Thsee marketing campaigns come with a cost and are often inefficient due to low client response rates.
+Prior work by Moro et al.
+(2014) discover that client characteristics and campaign related factors strongly influence subscription outcomes.
+Motivating a data-driven approach to improve targeting and decision-making in bank marketing.
+In their approach a Portuguese retail bank was studied using campaign data collected between 2008 and 2013.
+
+Our dataset was a publicily available dataset, thus had less variables for privacy concerns of the bank customers.
+The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns.
 These campaigns were done to encourage clients to subscribe to a long-term deposit product.
 Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
-Predictive modeling bears great value in this scenario, as it helps the bank focus resources on only those customers who depict a higher possibility of subscribing.
+The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts.
+A list of the 16 variable found below...
 
-In the subscription rate of this dataset, the problem is even more difficult as it is very low.
-Most individuals say "no," and only a small percentage decides to open a term deposit.
-If one class is rare, human judgment or simple rules fail to identify those clients who could actually be interested.
-A well-constructed model may present potential improvements in targeting accuracy, reduce the number of unsuccessful calls, and raise the return on investment.
-That is why it is necessary to research this dataset and look at which modeling approaches work best.
+* age: Age in years 
+* job: Occupation (Categorical)
+* marital: Marital status (Categorical)
+* education: Highest level of education (Categorical)
+* default: has credit in default? (binary)
+* balance: average yearly balance
+* housing: has housing loan? (binary)
+* loan: has personal loan? (binary)
+* Contract: contact communication type (Categorical)
+* day: last contact day (of the month)
+* month: last contact month of year (month; 1=January)
+* duration: last contact duration, in seconds
+* campaign: number of contacts performed during this campaign and for this client
+* pdays: number of days that passed by after the client was last contacted from a previous campaign (-1 means client was not previously contacted)
+* previous: number of contacts performed before this campaign and for this client
+* poutcome: outcome of the previous marketing campaign (Categorical)
+* y: Indicator of whether the client subscribed a term deposit (Binary target)
 
-**Description of dataset**  #TT 
-Each entry of the dataset represents one client contacted in one of the bank's telemarketing campaigns.
-This dataset contains 16 predictors carrying significant information about the client's background, financial situation, and interactions with previous campaigns.
-These variables fall into several key categories:
-Demographics include such factors as age, occupation, marital status, and education. These help describe who the clients are and often relate to financial behavior or economic stability.
-Financial indicators: Loan status, credit history, and average account balance.
-These variables describe the client's existing financial commitments and capacity, which may influence their willingness to invest in a long-term product like a term deposit.
+The response variable indicates whether a client subscribed to a term deposit which defines a binary classification task.
 
-Marketing campaign features include the type of contact: cellular or telephone, month of contact, and day of contact, and duration of the call.
-These are operational decisions made by the marketing team and may influence a client's likelihood of subscribing.
-For example, longer call duration sometimes is indicative of higher client engagement.
-Campaign outcomes: Number of previous contacts and if the past interactions have resulted in a successful outcome.
-These variables highlight whether the client has been interested in the past or has been contacted many times without success.
+Exploratory analysis shows that the numeric predictors (balance, duration, pdays, previously contacted) are heavily skewed.
+Also, analysis reveals substantial class imbalance, in the response variable, which motivates the use of appropriate modeling and evaluation for our classification problem.
+Visualizations of subscription rates by job education marital status contact type and previous outcome reveal variations in client responsiveness across different factors.
 
-The data includes both numerical and categorical variables, and this can serve as a good example to compare the many different classification methods with respect to how each treats different data types: for instance, linear models require the conversion of categorical variables, whereas tree-based or distance-based methods would naturally handle those in different ways.
-This allows us to explore how each approach might respond to the same set of features.
-The response variable, y, tells whether the client has subscribed to the term deposit or not.
-Only a small fraction of the contacted clients actually subscribe, therefore this is a strongly imbalanced dataset.
-The class imbalance is because, in general, marketing datasets are imbalanced since the majority of customers do not respond positively when outreach is performed.
-This skewness biases models toward the "no" prediction since this prediction would naturally be correct for most of the cases.
-Therefore, careful handling of this bias will be critical to generate models capable of identifying the small minority of clients who are likely to subscribe-the group that the bank cares most about.
+*[STREAMLINE Plots and EDA summary HERE]*
 
-**Problem Statement/Assumptions**  #TT
-It is now required to develop statistical models based on the available information in the dataset that will predict with accuracy whether a client will subscribe to a term deposit.
-This problem is of a binary classification nature, and our work focuses on comparing several models that differ in assumptions, flexibility, and interpretability.
-Understanding how such models perform is useful to a marketing team for making informed decisions about which clients to contact in future campaigns.
+Data preprocessing involved converting categorical variables to factors and adding derived variables such as pdays to account for clients not previously contacted.
+Converting the categorical variables (initilay character type) to factors creates an indicator for each class in the variable (i.e.
+if job has 20 factors, 20 indicator variables will be considered in the model fit).
+As for the previously contaced variable (days since previously contacted), customers who haven't been prreviously contacted will recieve a value of -1.
+This can cause improper modeling of customers, since the variable now takes on the value of negative numbers.
+Instead we set -1 to zeros and created a new variable to indicate customers that have not been previously coontacted before.
+These steps ensure that the dataset is suitable for statistical learning models and enable accurate assessment of predictive performance.
+\
 
-We focused our attention on well-performing models that give insightful knowledge of which factors drive the choice of a client.
-In the banking industry, understanding why a model has yielded some prediction is crucial.
-This supports ethical marketing practices, helps managers explain decisions, and can be important for regulatory compliance.
-For instance, highly flexible models may perform very well but may not indicate which client characteristics are actually relevant.
-More interpretable models do not always have the strongest accuracy but can guide marketing strategies more directly.
+### Problem Statement     
+This study is motivated by the availability of bank marketing datasets that enable comparison of machine learning methods for predicting term deposit subscription behavior.
 
-We make the following assumptions to keep the analysis focused and manageable:
-Independence of observations: We treat the observations as independent, even though multiple contacts with the same client may appear across different campaigns.
-This assumption makes the training and evaluation of the models easier, even if some repeated-measure structure in the data is ignored.
-No time-series modeling: variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns.
-However, we do not model this data as a time series.
+The specification of the problem begins with the recognition that we are addressing a binary classification task where the goal is to predict whether a client will subscribe to a term deposit based on demographic financial and campaign features.
+According to the framework presented in Introduction to Statistical Learning (ISLR), the fundamental goal is to estimate the unknown function, f(x) that relates the predictors, X to the response, Y.
+Some models are highly flexible and can capture complex non-linear relationships between predictors and the response.
+In most Cases these models achieve higher predictive accuracy.
+However, these flexible models often suffer from lower interpretability, making it difficult to understand the influence of individual features.
+Simpler models such as logistic regression are less flexible but provide coefficients that can be directly interpreted in terms of odds ratios, offering insight into feature importance.
+
+Chapter 5 of Introduction to Statistical Learning emphasizes the importance of obtaining an unbiased estimate of a model’s test error to assess its predictive performance on unseen data.
+Test error quantifies how well a model generalizes beyond the data used for training, which is critical for selecting among competing models.
+The chapter presents several methods for estimating test error, including the simple training/test split, K-fold cross validation, and the leave-one-out cross validation (LOOCV) approach.
+Each method balances bias and variance differently, with cross validation generally providing a lower-variance estimate compared to a single split.
+These techniques are essential for model selection, tuning hyperparameters, and comparing the expected performance of alternative models.
+For the Bank Marketing dataset, estimating test error accurately is particularly important due to the class imbalance and the presence of both categorical and numeric predictors, which can influence model generalization.
+
+In the following chpater of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting.
+It introduces methods such as subset selection, shrinkage (Ridge and LASSO regression), and dimension reduction to assess and enhance model accuracy while controlling variance.
+These approaches allow practitioners to identify parsimonious models that generalize well to new data, providing complementary strategies to traditional test error estimation.
+We apply these concepts to the models we fit on the bank data, to reduce overfitting and ultimately improve the generalization of our models to new clients.
+\
+
+We start with a logistic regression (described in methodology section), then build up adjustments and regularizations to the logestic regression to make our baseline.
+We then move to another interpretable model: KNN.
+This model is considered transperent in a unique, due to it's neighboring factor.
+Finally we fit a Boosting model.
+This state-of-the-art model is not considered interpretable.
+Although, there are methods used to extract the contributing variables in such models.
+We included this model is used to gage how much accucracy we are sacrificing for interpretability.
+
+The authors employed time ordered data splitting rolling window evaluation.
+For feature enrichment, they implimented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features.
+They compared Logistic regression with Decission trees, SVMs, and Neural networks; all complex 'black-box' models.
+While Moro et al.
+prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).
+
+## Methodology
+For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets.
+Our models do not apply time-series modeling, since our data does not include sequintial variables.
+Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series.
 This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.
-Use of class-relevant models: We restrict our approach to models presented in class because this enables us to contrast multiple concepts, including linear decision boundaries, distance-based approaches, and flexible tree-based structures.
-This renders the analysis suitable for a course project while still answering substantial questions regarding classification performance.
 
-Generalization from sample to population: We assume that this dataset is a good representation of the client population to which the bank will offer its services in forthcoming campaigns.
-If we couldn't assume this, our predictions wouldn't generalize properly to real-world decision-making.
-By selecting these assumptions, we are assured of a standard framework that will allow the comparison of different classification techniques on equal footing.
-The aim is not only to identify which model performs the best but also to understand how different modeling strategies behave when applied to imbalanced marketing data; this may provide practical insights into how a bank would structure its future campaigns and what features of clients are most relevant for predicting subscription behaviors.
+1. **Logistic regression**
 
-#### Data Wrangling and Cleaning:
-The analysis began by loading the Bank Marketing dataset and converting all categorical predictors into factor variables so they could be properly handled by the statistical modeling procedures in R.
-Several variables contained “unknown” values, which serve as placeholders rather than true missing observations.
-Multiple strategies for handling these values were considered, including treating them as missing and excluding the affected observations.
-Ultimately, “unknown” values were retained as their own category in order to preserve information and avoid altering the underlying class imbalance.
+```{r, include=FALSE}
+full_formula <- y ~ age + job + marital + education + default + balance + 
+  housing + loan + contact + day + month + duration + campaign + 
+  pdays + previous + poutcome + previously_contacted
 
-An additional feature, previously_contacted, was engineered to address the special coding of the pdays variable, where a value of −1 indicates that a client was not previously contacted.
-This transformation allowed the model to distinguish between first-time and repeat contacts more clearly.
+glm.full <- glm(full_formula, data = train_data, family = "binomial")
 
-To evaluate model performance fairly, the dataset was partitioned into an 80% training set and a 20% test set using stratified sampling based on the response variable.
-This ensured that the original class imbalance was preserved in both subsets.
-All models were trained on the same cleaned training data and evaluated on the same test data, allowing for direct and unbiased comparison across modeling approaches.
+# Perform backward selection
+glm.step <- step(glm.full, direction = "backward", trace = 0)
+
+selected_formula <- formula(glm.step)
+
+## 
+glm.baseline <- glm(selected_formula, data = train_data, family = "binomial")
+summary(glm.baseline)
+
+ctrl <- trainControl(
+  method = "cv",
+  number = 5,
+  classProbs = TRUE, 
+  summaryFunction = twoClassSummary 
+)
+
+set.seed(445) 
+cv_fit <- train(
+  selected_formula, 
+  data = train_data, 
+  method = "glm", 
+  family = "binomial",
+  trControl = ctrl, 
+  metric = "ROC" 
+)
+
+test_probabilities <- predict(glm.baseline, newdata = test_data, type = "response")
+test_predictions <- factor(ifelse(test_probabilities > 0.5, "yes", "no"), levels = levels(test_data$y))
+
+conf_matrix <- confusionMatrix(data = test_predictions, reference = test_data$y, positive = "yes" )
+
+# Extract deviance residuals
+residuals <- residuals(glm.baseline, type = "deviance")
+train_data$residuals <- residuals
+
+p1 <- ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+
+glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
+    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
+summary(glm.tran)
+tran.coef <- summary(glm.tran)$coef
+
+residuals <- residuals(glm.tran, type = "deviance")
+train_data$residuals <- residuals
+
+p2 <- ggplot(train_data, aes(x = balance, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "loess", color = "red", se = FALSE) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
+  labs(title = "Deviance Residuals vs. Balance",
+       y = "Deviance Residuals") +
+  theme_minimal()
+
+## Regularization
+X_full <- model.matrix(y ~ . - 1, data = bank_full)
+Y_full <- as.numeric(bank_full$y) - 1
+
+X_train <- X_full[train_index, ]
+Y_train <- Y_full[train_index]
+
+X_test <- X_full[-train_index, ]
+Y_test <- Y_full[-train_index]
+
+cv.ridge <- cv.glmnet(X_train, Y_train, 
+                      alpha = 0, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.ridge <- cv.ridge$lambda.min
+
+cv.lasso <- cv.glmnet(X_train, Y_train, 
+                      alpha = 1, 
+                      family = "binomial", 
+                      type.measure = "class")
+lambda.min.lasso <- cv.lasso$lambda.min
+
+baseline_prob <- predict(glm.tran, newdata = test_data, type = "response")
+baseline_pred <- factor(ifelse(baseline_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+ridge_prob <- predict(cv.ridge, s = lambda.min.ridge, newx = X_test, type = "response")
+ridge_pred <- factor(ifelse(ridge_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+lasso_prob <- predict(cv.lasso, s = lambda.min.lasso, newx = X_test, type = "response")
+lasso_pred <- factor(ifelse(lasso_prob > 0.5, "yes", "no"), levels = c("no", "yes"))
+
+cm_baseline <- confusionMatrix(baseline_pred, test_data$y, positive = "yes")
+cm_ridge <- confusionMatrix(ridge_pred, test_data$y, positive = "yes")
+cm_lasso <- confusionMatrix(lasso_pred, test_data$y, positive = "yes")
+
+metrics <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
+  Accuracy = c(cm_baseline$overall['Accuracy'], cm_ridge$overall['Accuracy'], cm_lasso$overall['Accuracy']),
+  Sensitivity = c(cm_baseline$byClass['Sensitivity'], cm_ridge$byClass['Sensitivity'], cm_lasso$byClass['Sensitivity']),
+  Specificity = c(cm_baseline$byClass['Specificity'], cm_ridge$byClass['Specificity'], cm_lasso$byClass['Specificity'])
+)
+
+roc_baseline <- roc(test_data$y, baseline_prob)
+roc_ridge <- roc(test_data$y, as.numeric(ridge_prob))
+roc_lasso <- roc(test_data$y, as.numeric(lasso_prob))
+
+roc_data <- data.frame(
+  fpr = c(1 - roc_baseline$specificities, 1 - roc_ridge$specificities, 1 - roc_lasso$specificities),
+  tpr = c(roc_baseline$sensitivities, roc_ridge$sensitivities, roc_lasso$sensitivities),
+  Model = factor(c(rep("Baseline Logit", length(roc_baseline$specificities)),
+                   rep("Ridge Logit", length(roc_ridge$specificities)),
+                   rep("LASSO Logit", length(roc_lasso$specificities))),
+                 levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit"))
+)
+
+```
+
+Logistic regression is used to model the probability of a binary outcome as a function of predictor variables.
+The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$.
+This model fits with assumptions of independent observations and normality.
+
+Formula: $$f(x) = P(Y=1 \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k)}}$$  
+Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data (ISLR).
+
+Feature selection is performed using backward stepwise selection to identify a parsimonious subset of predictors that minimizes estimated test error as measured by AIC.
+Variables such as duration and previous marketing outcomes remain in the model due to their strong predictive power, whereas less informative variables are excluded to reduce variance.
+Categorical predictors are converted to factors, and a new binary variable previously_contacted is created to handle clients with no prior campaign contact.
+This preprocessing ensures that the data is suitable for statistical modeling and facilitates accurate interpretation of model coefficients.
+We resulted 12 features:
+
+```{r}
+selected_formula
+```
+
+Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error.
+Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority.
+This model performed will with mean 5-fold CV test error metrics:
+
+```{r, echo=FALSE}
+print(conf_matrix)
+```
+
+Next, we check for potential non-linearity in continuous predictors from our predictors remaining after back-ward stepwise selection. We plotted deviance residuals against numeric features.
+Based on residual diagnostics, we incorporated a polynomial term (using the poly function: *poly(balance, 4)*) for balance to account for non-linear effects.
+The Balance squared and power 4 term produced significant coefficients (Last 4 on the list of coefficients).
+
+```{r, out.width="45%", echo=FALSE}
+p1
+p2
+```
+
+```{r}
+tran.coef
+```
+
+Although this polynomial transformation didn't provide desired metrics in expense of the nonlinear effect.
+From the deviance residual plots we can observe there weren't significant deviance from a normal residual line.
+We will not include this model in our comparison.
+
+Regularization techniques such as Ridge ($\ell_2$ penalty) and LASSO ($\ell_1$ penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors.
+LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$).
+A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6).
+The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`.
+Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.
+
+```{r, out.width="45%", echo=FALSE}
+plot(cv.ridge)
+plot(cv.lasso)
+```
+
+Here is a comparison of the models fitted thus far:
+
+```{r}
+metrics
+```
+
+Ridge and LASSO models yield similar results, indicating that additional shrinkage does not meaningfully improve classification of subscribers.
+We assume this is due to our modeling being majority indicator variables.
+This binary aspect of the indicator variable causes it's coefficient to be less affected by regularization techniques.
+
+```{r, out.width="45%", echo=FALSE}
+ggplot(roc_data, aes(x = fpr, y = tpr, color = Model)) +
+  geom_line(linewidth = 1) +
+  geom_abline(linetype = "dashed", color = "grey") +
+  labs(
+    title = "ROC Curve Comparison for Logistic Models",
+    x = "False Positive Rate (1 - Specificity)",
+    y = "True Positive Rate (Sensitivity)"
+  ) +
+  theme_minimal()
+```
+
+All three models exhibit a high overall Accuracy (around 0.89), which is expected given the significant class imbalance (Prevalence $\approx 0.11$).
+However, the critical metric, Sensitivity (the True Positive Rate for the minority class, 'yes'), is very low (around 0.17).
+This result, consistent across all three models, indicates that they struggle to correctly identify subscribers, leading to many False Negatives (FN).
+Conversely, the high Specificity (True Negative Rate $\approx 0.98$) means the models are excellent at identifying non-subscribers.
+The regularization methods (Ridge and LASSO) failed to provide any significant improvement in these metrics, suggesting that the primary limitation is one of high bias (due to insufficient flexibility) rather than high variance.
+\
 
@@
 
-We evaluated five main models from the course:
-
-- Logistic Regression — A linear classifier that estimates the log-odds of subscription. It is highly interpretable and provides coefficients that can be directly translated into marketing insights.
+We evaluated five main models from the course: ***#only 3 models***
 

### 2025-12-15 Paper.Rmd
added eda section

@@
 subtitle: "Machine Learning Models on A Bank Marketing Dataset"
-author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alaahtani"
+author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alqahtani"
 output: pdf_document
+header-includes:
+  - \usepackage{float}
 ---
@@
 
-*[STREAMLINE Plots and EDA summary HERE]*
+## Exploratory Data Analysis (EDA)
+
+### Data Characteristics and Class Imbalance
+
+Our exploratory analysis examined 45,211 client records from the Portuguese banking institution.
+Analysis of the target variable revealed severe class imbalance, with only 11.7% of clients subscribing to term deposits compared to 88.3% who declined (Figure 1).
+This imbalance represented a key challenge for predictive modeling.
+
+```{r fig1, echo=FALSE, fig.cap="Class imbalance in term deposit subscriptions.
+Only 11.7% of clients subscribed, presenting a significant modeling challenge.", fig.align='center', out.width='70%', fig.pos='H'}
+
+ggplot(bank_full, aes(x = y, fill = y)) +
+  geom_bar(color = "white", linewidth = 0.4) +
+  scale_fill_manual(
+    values = c("no" = "grey75", "yes" = "seagreen3")
+  ) +
+  labs(
+    x = "Subscribed?",
+    y = "Count"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(
+    legend.position = "none",
+    panel.grid.major.x = element_blank()
+  )
+
+```
+
+### Demographic Patterns
+
+Demographic patterns emerged from the data: retired clients and students were most likely to subscribe, while blue-collar workers were least likely.
+Higher education levels correlated with increased subscription rates.
+
+```{r fig2, echo=FALSE, fig.cap="Distribution of job categories in the dataset.
+Administrative, blue-collar, and technician occupations are most common.", fig.align='center', out.width='70%', fig.pos='H'}
+
+ggplot(bank_full, aes(x = job)) +
+  geom_bar(fill = "steelblue3") +
+  coord_flip() +
+  labs(
+    x = "Job",
+    y = "Count"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(
+    panel.grid.major.y = element_blank()
+  )
+
+```
+
+### Historical Patterns
+
+Analysis of previous campaign outcomes revealed powerful patterns: clients with previous successful subscriptions demonstrated approximately 65% likelihood of subscribing again, compared to only 15% among those with previous unsuccessful outcomes.
+
+```{r fig3, echo=FALSE, fig.cap="Previous campaign outcome influence.
+Clients with previous success show 65% subscription rates compared to 15% for previous failures.", fig.align='center', out.width='70%', fig.pos='H'}
+
+ggplot(bank_full, aes(x = poutcome, fill = y)) +
+  geom_bar(position = "fill", color = "white", linewidth = 0.3) +
+  scale_fill_manual(
+    values = c("no" = "grey75", "yes" = "seagreen3")
+  ) +
+  scale_y_continuous(labels = scales::percent) +
+  labs(
+    x = "Previous Outcome",
+    y = "Proportion Subscribed"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(
+    legend.position = "none"
+  )
+
+```
+
+### Campaign Dynamics and Diminishing Returns
+
+Campaign persistence analysis showed a clear downward trend in success rates as contact frequency increased, with substantial decline after five attempts and minimal success beyond thirty contacts.
+
+```{r fig4, echo=FALSE, fig.cap="Diminishing returns from repeated contacts. Success rates drop significantly after 5 contacts.", fig.align='center', out.width='70%', fig.pos='H'}
+
+bank_full %>%
+  group_by(campaign) %>%
+  summarise(success_rate = mean(y == "yes")) %>%
+  ggplot(aes(x = campaign, y = success_rate)) +
+  geom_line(color = "steelblue3", linewidth = 1.2) +
+  geom_point(color = "darkorange2", size = 2.8) +
+  scale_y_continuous(labels = scales::percent) +
+  labs(
+    x = "Number of Contacts",
+    y = "Success Rate"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(
+    panel.grid.minor = element_blank()
+  )
+
+```
 

### 2025-12-16 Paper.Rmd
paper polishing

@@
 
-## Motivation
+# Motivation
 
 Bank telemarketing campaigns are a major channel through which financial institutions promote term deposit products.
-Thsee marketing campaigns come with a cost and are often inefficient due to low client response rates.
+These marketing campaigns come with a cost and are often inefficient due to low client response rates.
 Prior work by Moro et al.
@@
 
-Our dataset was a publicily available dataset, thus had less variables for privacy concerns of the bank customers.
+Our dataset was a publicly available dataset, thus had less variables for privacy concerns of the bank customers.
 The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns.
@@
 
-### Problem Statement     
+## Problem Statement     
 This study is motivated by the availability of bank marketing datasets that enable comparison of machine learning methods for predicting term deposit subscription behavior.
@@
 
-In the following chpater of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting.
+In the following chapter of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting.
 It introduces methods such as subset selection, shrinkage (Ridge and LASSO regression), and dimension reduction to assess and enhance model accuracy while controlling variance.
@@
 
-We start with a logistic regression (described in methodology section), then build up adjustments and regularizations to the logestic regression to make our baseline.
+We start with a logistic regression (described in methodology section), then build up adjustments and regularization to the logestic regression to make our baseline.
 We then move to another interpretable model: KNN.
-This model is considered transperent in a unique, due to it's neighboring factor.
+This model is considered transparent in a unique, due to it's neighboring factor.
 Finally we fit a Boosting model.
@@
 Although, there are methods used to extract the contributing variables in such models.
-We included this model is used to gage how much accucracy we are sacrificing for interpretability.
+We included this model is used to gage how much accuracy we are sacrificing for interpretability.
 
-The authors employed time ordered data splitting rolling window evaluation.
-For feature enrichment, they implimented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features.
-They compared Logistic regression with Decission trees, SVMs, and Neural networks; all complex 'black-box' models.
-While Moro et al.
-prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).
+The authors in (Moro et al.) employed time ordered data splitting rolling window evaluation.
+For feature enrichment, they implemented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features.
+They compared Logistic regression with Decision trees, SVMs, and Neural networks; all complex 'black-box' models.
+While (Moro et al.) prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).
 
-## Methodology
+# Methodology
 For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets.
-Our models do not apply time-series modeling, since our data does not include sequintial variables.
+Our models do not apply time-series modeling, since our data does not include sequential variables.
 Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series.
 This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.
+An 80/20 random train–test split was used, and all modeling decisions, including feature selection and hyper-parameter tuning, were made using only the training data.
+This prevented information leakage and ensured that the test set provided an honest evaluation of generalization performance.
 
@@
 
+### Interpretability
+To understand the meaning of the coefficients, we can transform the formula into the log-odds (logit) form:
+$$ \log\left(\frac{P(Y=1 \mid \mathbf{x})}{1 - P(Y=1 \mid \mathbf{x})}\right) = \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k$$
+Consider a single coefficient, say $\beta_{balance}$, which relates to a client's average bank account balance.
+If the estimated coefficient for balance is $\beta_{balance} = 0.001$, this means that a one-unit increase in the client's balance is associated with a change of 0.001 in the log-odds of subscribing to a term deposit, assuming all other variables remain constant.
+This change in log-odds translates to a multiplicative change of $e^{0.001} \approx 1.001$ in the odds.
+Therefore, for every unit increase in balance, the odds of subscription increase by a factor of 1.001.
+This shows a clear relationship between the coefficient and the target variable.
+
+Without loss of generality, we can generalize this interpretation and estimation process to logistic regression models with multiple variables and their variations, such as incorporating Lasso regularization (for variable selection and complexity control) or applying polynomial transformations to the predictor variables (to capture non-linear relationships).
+
+### Model fit
+
 Feature selection is performed using backward stepwise selection to identify a parsimonious subset of predictors that minimizes estimated test error as measured by AIC.
@@
 
-```{r}
-tran.coef
+```{r, echo=FALSE}
+full_coefs <- summary(glm.tran)$coefficients
+ballance_coefs <- c("poly(balance, 4)1", "poly(balance, 4)2", "poly(balance, 4)3", "poly(balance, 4)4")
+subset_coefs <- full_coefs[ballance_coefs, c(1, 4)]
+
+print(subset_coefs)
 ```
@@
 LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$).
-A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6).
+A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (ISLR Chapter 6).
 The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`.
@@
 
-We evaluated five main models from the course: ***#only 3 models***
+We evaluated five main models from the course:
+
+2. **K-Nearest Neighbors (KNN)**
 
@@
 
+3. **Boosting**
+
 Boosting was used as the most flexible modeling approach in this study to address the strong class imbalance and complex predictor relationships present in the Bank Marketing dataset.
@@
 ### Model Training, Formulation, and Validation
-All models were trained and evaluated within a consistent validation framework to ensure fair comparison and reliable estimation of out-of-sample performance.
-An 80/20 stratified train–test split was used, and all modeling decisions, including feature selection and hyperparameter tuning, were made using only the training data.
-This prevented information leakage and ensured that the test set provided an honest evaluation of generalization performance.
-
-For the logistic regression models, feature selection was performed using backward stepwise selection based on AIC to obtain a parsimonious set of predictors.
-Residual diagnostics were used to assess model assumptions, leading to the inclusion of nonlinear terms where appropriate.
-Regularization techniques, including Ridge and LASSO regression, were applied to address multicollinearity and reduce variance.
-The regularization parameter was selected using 5-fold cross-validation.
 
@@
 
-## Results
+# Results
 
@@
 
-## Discussion
+# Discussion
 
@@
 https://archive.ics.uci.edu/dataset/222/bank+marketing
+
+James, Gareth, et al. An Introduction to Statistical Learning : With Applications in R. Springer, 2013.
+
+Moro, Sérgio, et al. A Data-Driven Approach to Predict the Success of Bank Telemarketing. Decision Support Systems, vol. 62, June 2014, pp. 22–31.

### 2025-12-17 Paper.Rmd
finished the visuals of the paper for boosting, KNN, and results

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+library(readr)
+library(tidyverse)
+```
 
@@
 It captures nonlinear relationships but can be sensitive to scaling and high-dimensional data.
-#TT
 
@@
 
+```{r}
+# KNN Model
+library(class)
+
+# Remove any extra columns that might have been added
+train_data_clean <- train_data %>% 
+  dplyr::select(-any_of(c("residuals", "y_num")))
+test_data_clean <- test_data %>% 
+  dplyr::select(-any_of(c("residuals", "y_num")))
+
+# Prepare data for KNN (standardize numeric variables)
+numeric_vars <- c("age", "balance", "day", "duration", "campaign", "pdays", "previous")
+
+# Create model matrices from combined data to ensure same dimensions
+full_data_temp <- rbind(
+  train_data_clean %>% mutate(set = "train"),
+  test_data_clean %>% mutate(set = "test")
+)
+
+# Standardize numeric variables
+full_data_temp[numeric_vars] <- scale(full_data_temp[numeric_vars])
+
+# Create model matrix for all data at once
+X_full_knn <- model.matrix(y ~ . - set - 1, data = full_data_temp)
+
+# Split back into train and test
+X_train_knn <- X_full_knn[full_data_temp$set == "train", ]
+X_test_knn <- X_full_knn[full_data_temp$set == "test", ]
+
+# Find optimal k using cross-validation
+set.seed(445)
+k_values <- c(3, 5, 7, 9, 11, 15, 20, 25, 30)
+
+# Use caret for easier CV with KNN
+ctrl_knn <- trainControl(method = "cv", number = 5)
+
+set.seed(445)
+knn_cv <- train(
+  y ~ .,
+  data = train_data_clean,
+  method = "knn",
+  trControl = ctrl_knn,
+  tuneGrid = data.frame(k = k_values)
+)
+
+best_k <- knn_cv$bestTune$k
+
+# Fit final KNN model
+knn_pred <- knn(
+  train = X_train_knn,
+  test = X_test_knn,
+  cl = train_data_clean$y,
+  k = best_k,
+  prob = TRUE
+)
+
+# Get probabilities
+knn_prob <- attr(knn_pred, "prob")
+knn_prob <- ifelse(knn_pred == "yes", knn_prob, 1 - knn_prob)
+
+# Confusion matrix
+cm_knn <- confusionMatrix(knn_pred, test_data_clean$y, positive = "yes")
+
+# ROC curve
+roc_knn <- roc(test_data_clean$y, knn_prob)
+```
+
+```{r, out.width="45%",}
+# Confusion Matrix
+cm_knn
+
+# Best k value
+cat("\nOptimal number of neighbors (k):", best_k, "\n")
+cat("Cross-validation accuracy:", round(max(knn_cv$results$Accuracy), 4), "\n")
+```
+
+```{r, out.width="45%",}
+# CV results plot
+plot(knn_cv, main = "KNN Cross-Validation Performance")
+```
+
+```{r, out.width="45%",}
+# ROC curve
+plot(roc_knn, col = "steelblue", lwd = 2)
+abline(a = 0, b = 1, lty = 2, col = "gray")
+text(0.6, 0.3, paste("AUC =", round(auc(roc_knn), 4)), cex = 1.2)
+```
+
 3. **Boosting**
@@
 
-The boosting model was trained on the same 80% training set as the other models using a Bernoulli loss function appropriate for binary classification.
-To allow the model sufficient capacity to learn complex structure, the model was initially trained with 1,500 trees.
-Model complexity was controlled using a small learning rate and shallow tree depth to encourage gradual improvement rather than abrupt overfitting.
-The optimal number of trees was selected using out-of-bag (OOB) error estimation.
-This procedure identified approximately 1,100 trees as the point at which predictive performance was maximized.
-Beyond this value, additional trees did not lead to meaningful improvements, indicating that the model had effectively converged.
-Final predictions and evaluation metrics were therefore computed using 1,100 trees to balance flexibility and generalization.
+We implemented gradient boosting using a Bernoulli loss function for binary classification.
+The model was trained on the same 80% training set as the other methods.
+To encourage gradual learning and prevent overfitting, we used a small learning rate (shrinkage = 0.01) and shallow trees (interaction depth = 3).
+The model was initially trained with 1,500 trees to allow sufficient flexibility.
+
+The optimal number of trees was selected using out-of-bag (OOB) error estimation, which serves a similar role to cross-validation in ensemble methods.
+The OOB procedure identified the point at which additional trees no longer meaningfully improved predictive performance.
+Final predictions were computed using the optimal number of trees selected by this procedure.
+
+```{r, out.width="45%", boosting-fit, echo=FALSE, fig.cap="Out-of-bag error used to select the optimal number of boosting trees.", fig.pos='H'}
+library(gbm)
+
+train_data <- train_data %>% dplyr::select(-any_of("residuals"))
+test_data  <- test_data  %>% dplyr::select(-any_of("residuals"))
+
+train_data$y_num <- ifelse(train_data$y == "yes", 1, 0)
+test_data$y_num  <- ifelse(test_data$y == "yes", 1, 0)
+
+set.seed(445)
+boost_model <- gbm(
+  y_num ~ . - y,
+  data = train_data,
+  distribution = "bernoulli",
+  n.trees = 1500,
+  shrinkage = 0.01,
+  interaction.depth = 3,
+  bag.fraction = 0.5,
+  verbose = FALSE
+)
+
+best_trees <- gbm.perf(boost_model, method = "OOB", plot.it = TRUE)
+
+```
+
+```{r, out.width="45%",}
+boost_prob <- predict(boost_model, test_data, n.trees = best_trees, type = "response")
+boost_pred <- factor(ifelse(boost_prob > 0.5, "yes", "no"), levels = c("no","yes"))
+
+cm_boost <- confusionMatrix(boost_pred, test_data$y, positive = "yes")
+cm_boost
+
+```
+
+```{r, out.width="45%",}
+roc_boost <- roc(test_data$y, boost_prob)
+plot(roc_boost, col = "darkgreen", lwd = 2)
+abline(a = 0, b = 1, lty = 2)
+
+```
 
 When evaluated on the test set, boosting achieved the strongest overall performance among all models considered.
-It produced the highest accuracy, sensitivity, and AUC, substantially improving the identification of the minority “yes” class compared to logistic regression and KNN.
-In particular, boosting correctly identified over 40% of subscribing clients while maintaining high specificity, demonstrating its ability to reduce false negatives without substantially increasing false positives.
-These results indicate that boosting is well suited for imbalanced marketing data where identifying potential subscribers is a primary objective.
+It produced the highest accuracy (0.905), sensitivity (0.412), and AUC (0.927).
+In particular, boosting correctly identified over 40% of subscribing clients while maintaining high specificity.
+This represents a substantial improvement over logistic regression, KNN, and GAM, all of which struggled to identify the minority “yes” class.
+These results highlight the advantage of flexible ensemble methods in reducing bias and capturing complex nonlinear patterns in imbalanced marketing data.
 
@@
 
-This study evaluated multiple statistical and machine learning models for predicting term deposit subscriptions using the Bank Marketing dataset.
-Due to the strong class imbalance, model performance was assessed using accuracy, sensitivity, and AUC rather than accuracy alone.
-Logistic regression and its regularized variants demonstrated strong overall discrimination and high AUC values, but they struggled to identify the minority class of subscribing clients.
-K-Nearest Neighbors performed less effectively, showing limited discriminatory power in this high-dimensional, imbalanced setting.
+Model performance was evaluated on a held-out test set using accuracy, sensitivity, and AUC.
+Because the dataset is highly imbalanced, particular emphasis was placed on sensitivity and AUC as measures of a model’s ability to identify the minority “yes” class.
 
-Boosting outperformed all other models, achieving the highest accuracy, sensitivity, and AUC on the test set.
-Its superior performance indicates that flexible ensemble methods are better suited for capturing complex, nonlinear relationships in marketing data.
-However, logistic regression remains valuable due to its interpretability and ability to provide actionable insights.
-Overall, the results highlight an important trade-off between predictive performance and interpretability when modeling imbalanced marketing data.
+```{r}
+# Create final results comparison table
+final_results <- data.frame(
+  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit", "KNN", "Boosting"),
+  Accuracy = c(
+    cm_baseline$overall['Accuracy'],
+    cm_ridge$overall['Accuracy'],
+    cm_lasso$overall['Accuracy'],
+    cm_knn$overall['Accuracy'],  # You'll need to create this
+    cm_boost$overall['Accuracy']
+  ),
+  Sensitivity = c(
+    cm_baseline$byClass['Sensitivity'],
+    cm_ridge$byClass['Sensitivity'],
+    cm_lasso$byClass['Sensitivity'],
+    cm_knn$byClass['Sensitivity'],  # You'll need to create this
+    cm_boost$byClass['Sensitivity']
+  ),
+  AUC = c(
+    auc(roc_baseline),
+    auc(roc_ridge),
+    auc(roc_lasso),
+    auc(roc_knn),  # You'll need to create this
+    auc(roc_boost)
+  )
+)
+
+# Round for better display
+final_results[, 2:4] <- round(final_results[, 2:4], 4)
+```
+
+```{r}
+final_results
+```
+
+Boosting outperformed all other models across all key performance metrics.
+It achieved the highest accuracy (0.9047), sensitivity (0.4115), and AUC (0.9266), indicating superior ability to identify clients likely to subscribe to a term deposit.
+This improvement in sensitivity is particularly important in the bank marketing context, where failing to identify potential subscribers represents a missed business opportunity.
+
+The logistic regression models (baseline, Ridge, and LASSO) achieved similar accuracy levels (approximately 0.898–0.900) and strong AUC values (around 0.907), indicating good overall discrimination.
+However, their sensitivity remained relatively low, suggesting that these models frequently misclassified subscribing clients as non-subscribers.
+Regularization did not substantially improve performance, indicating that model bias rather than variance was the primary limitation.
+
+KNN performed the weakest among all models, with the lowest AUC (0.617), reflecting poor discriminatory power in this high-dimensional setting.
+GAM demonstrated improved AUC (0.854) but still failed to substantially improve sensitivity, indicating limited gains from smooth nonlinear terms alone.
+
+```{r, out.width="45%",}
+# Combine all ROC data for plotting
+roc_all <- data.frame(
+  fpr = c(
+    1 - roc_baseline$specificities,
+    1 - roc_ridge$specificities,
+    1 - roc_lasso$specificities,
+    1 - roc_knn$specificities,
+    1 - roc_boost$specificities
+  ),
+  tpr = c(
+    roc_baseline$sensitivities,
+    roc_ridge$sensitivities,
+    roc_lasso$sensitivities,
+    roc_knn$sensitivities,
+    roc_boost$sensitivities
+  ),
+  Model = factor(
+    c(
+      rep("Baseline Logit", length(roc_baseline$specificities)),
+      rep("Ridge Logit", length(roc_ridge$specificities)),
+      rep("LASSO Logit", length(roc_lasso$specificities)),
+      rep("KNN", length(roc_knn$specificities)),
+      rep("Boosting", length(roc_boost$specificities))
+    ),
+    levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit", "KNN", "Boosting")
+  )
+)
+```
+
+```{r, out.width="45%",}
+ggplot(roc_all, aes(x = fpr, y = tpr, color = Model)) +
+geom_line(linewidth = 1) +
+geom_abline(linetype = "dashed") +
+labs(
+x = "False Positive Rate",
+y = "True Positive Rate"
+) +
+theme_minimal()
+
+```
+
+The ROC curves reinforce the numerical results.
+Boosting dominates across most thresholds, confirming its superior ranking ability.
+Logistic regression models cluster closely together, while KNN performs only slightly better than random classification.
 
